---
title: "The `afli` Oracle database — table inventory and schema evolution"
author: "Einar Hjörleifsson"
date: today
---

```{r setup}
#| include: false
library(nanoparquet)
library(dplyr)
library(lubridate)
library(ggplot2)
```

## Overview

The `afli` Oracle schema is Fiskistofa's primary operational catch-and-effort
database. Its 96 tables span five decades of Icelandic fishing records, and its
structure reflects the successive layers of technology through which logbook
data have been collected: paper forms, early electronic entry, XML-based digital
submissions, and on-board automatic loggers.

The parquet dump in `data-raw/data-dump/afli/` is a static snapshot exported
via `data-raw/data-dump/oracle_dump.R`. Metadata tables `_overview_tables.parquet`
and `_overview_fields.parquet` carry Oracle table comments and column
documentation where available.

The tables can be grouped into six functional families, described in the
sections below.

---

## Schema evolution and consolidation {#sec-eras}

The naming convention, temporal coverage, and `visir` sign distribution of the
tables reveal how the database evolved — and how it is structured as a
consolidated whole.

### The `visir` key space

`visir` (Icelandic: pointer / index) is the universal station key throughout
the schema. Its sign distinguishes provenance:

| `visir` sign | Origin |
|---|---|
| **Positive** | Paper-era record (transcribed from physical logbook) |
| **Negative** | Electronic-era record (loaded from XML digital submission) |

### Master tables vs staging tables

The central finding from examining the `visir` distributions is that `stofn`,
`afli`, `toga`, `lineha`, `hringn`, and `gildra` are **consolidated master
tables** — they contain records from both eras:

| Master table | Positive visir (paper) | Negative visir (electronic) | Electronic share |
|---|---|---|---|
| `stofn` | 5,610,300 | 1,275,274 | 18.5% |
| `afli` | 11,610,187 | 4,064,218 | 26.0% |
| `toga` | 4,447,596 | 1,029,820 | 18.8% |
| `lineha` | 1,056,760 | 236,254 | 18.3% |

Cross-matching confirms that ~96% of `rafr_stofn` rows reappear in `stofn`
(negative visir), and similarly for `rafr_toga` (~97%) and `rafr_lineha`
(~97%). The `rafr_*` tables are therefore the **import staging layer** —
records are loaded from XML into `rafr_stofn` (and its gear subtables) and
simultaneously written to the master `stofn` (and its gear subtables).

However, `stofn` has 1,275,274 negative-visir rows while `rafr_stofn`
has only 1,149,278 — the extra ~166K come from a **second electronic
pathway** (see Phase 4 below) that loaded records directly into `stofn`
without creating `rafr_*` staging rows. A further ~40K `rafr_stofn` rows
do not appear in `stofn`: 16,912 with `eytt = 1` (officially deleted) and
~23K that were likely rejected during Oracle consolidation.

The distinction matters for analysis: the master tables (`stofn` / `afli` /
`toga` …) give the most complete single view spanning all eras, while the
`rafr_*` staging tables are the only place to find the XML back-links
(`xml_sending_visir`, `xml_faerslu_visir`) needed to trace a record back to
the original submission.

### Chronology — three phases

**Phase 1: Paper logbooks (~1950s–2007).** The core tables predate any
electronic system. Data were transcribed by inspectors from physical logbooks
submitted at landing. Station positions recorded as DDMM integers — when
recorded at all; early records have a high rate of missing coordinates.
`stofn.vedags` (fishing date, paper era only) shows continuous data from the
1970s onward, with scattered records back to 1950. In `stofn`, positive-visir
records continue through 2022 (declining in volume after ~2014 as digital
submissions displaced paper).

**Phase 2: Early electronic submissions (~2003–2007).** A small number of
vessels began submitting XML logbooks from ~2003. The `rafr_stofn.hledsla_dags`
column (the Oracle import timestamp) shows the volume clearly:

| Year | Stations loaded |
|---|---|
| 2003 | 408 |
| 2004 | 2,282 |
| 2005–2007 | ~7,000–10,000 / yr |
| **2008** | **43,263** ← mandatory rollout |

**Phase 3: Mandatory electronic logbooks (~2008–2020).** The 2008 inflection
in `hledsla_dags` marks a regulatory change requiring large vessels to submit
digital logbooks. Annual volumes stabilise at ~80,000–100,000 stations/year
through 2019. From 2020 onwards the `afli`/`rafr_*` series tapers as the
newer `fs_afladagbok` and `adb` systems take over reception of fresh
submissions.

**Phase 4: Direct-load electronic records (2020–2022).** `stofn` contains
~165K negative-visir records in 2020–2022 that have no counterpart in
`rafr_stofn`. These appear to be electronic submissions that entered `stofn`
directly — bypassing the `rafr_*` staging pipeline — during the transition
period when `fs_afladagbok` was becoming the primary reception system. They
use the same old gear codes (veidarf 1–15) and the same DDMM coordinate
encoding as the Phase 3 records; they are otherwise indistinguishable in the
master tables.

The `afladagb_skil` / `afladagb_skil_linur` compliance tables were added in
**January 2014** (per Oracle comment) to formally track whether vessels met
their monthly submission obligations. Physical logbook delivery is recorded in
`afladagbok_afhending` for vessels that did not participate digitally.

---

## Family 1 — Core fishing records {#sec-core}

### `stofn` — fishing stations (consolidated master)

The central fact table. One row per fishing operation (set, tow, or haul).

```{r}
#| label: tbl-stofn
#| tbl-cap: "Key columns of `stofn`."
tribble(
  ~Column,       ~Type,    ~Description,
  "visir",       "int",    "Unique station key",
  "skipnr",      "int",    "Vessel registration number",
  "veidarf",     "int",    "Gear type code (old vocabulary)",
  "ahofn",       "int",    "Crew size (Oracle: *Fjöldi manna í áhöfn skipsins*)",
  "vedags",      "date",   "Fishing date",
  "breidd",      "int",    "Start latitude — DDMM integer (e.g. 6534 = 65°34′N)",
  "lengd",       "int",    "Start longitude — DDMM integer (stored unsigned; West assumed for Iceland)",
  "breidd_lok",  "int",    "End latitude — DDMM integer",
  "lengd_lok",   "int",    "End longitude — DDMM integer",
  "reitur",      "int",    "Statistical square (1°×1°)",
  "smareitur",   "int",    "Statistical sub-square (0.5°×0.5°)",
  "dypi",        "int",    "Depth (m)",
  "ldags",       "date",   "Landing date",
  "lhofn",       "int",    "Landing harbour"
) |> knitr::kable()
```

`stofn` has **6,885,574** rows. It is the consolidated master station table
spanning all collection eras: positive `visir` = paper-era records (1950s–2022,
bulk in 1970s–2010s); negative `visir` = electronic-era records (2003–2024,
~96% overlap with `rafr_stofn`). The `visir` primary key is shared by all
gear-specific subtables in both eras. Coordinates are DDMM integers and must be
converted with `geo::geoconvert.1()` (multiply by 100 before passing in; negate
longitude for West hemisphere). Missing coordinates are common in older records;
`pos_lon` / `pos_lat` columns provide decimal-degree fallbacks where the Oracle
import process was able to derive them from statistical squares.

### `afli` — catch records (consolidated master)

One row per species per station. Oracle comment: *"Sameiginleg fyrir öll
veiðarfæri"* (common to all gear types). Like `stofn`, this is a consolidated
master table: 11.6M positive-visir (paper-era) rows plus 4.1M negative-visir
(electronic-era) rows, the latter overlapping with `rafr_afli`. Columns:
`visir` (→ `stofn`), `tegund` (species code), `magn` (catch weight), `afli`
(catch in standard units), `astand` (condition: S = poor, L = good).

**15,674,405** catch rows total; `visir` is the join key to `stofn`.

### Gear subtables

Each links to `stofn` via `visir`. Together they cover all gear classes.

| Table | Rows | Gear | Key effort columns |
|---|---|---|---|
| `toga` | 5,477,416 | Mobile (trawl, seine) | `togtimi` (min), `grandarar` (sweep width m), `hoflina` (head-line m), `moskvi` (mesh mm) |
| `lineha` | 1,293,014 | Longline / gillnet | `onglar` (hooks/set), `bjod` (sets), `dregin` (nets hauled), `naetur` (net-nights), `faeri` (jigs), `klst` (hours) |
| `hringn` | 107,778 | Purse seine / ring-net | `l_notar` (net length m), `d_notar` (net depth m), `kast_nr` (cast number) |
| `gildra` | 5,269 | Traps | `gildrur` (trap count), `klst` (soak hours) |

### Ancillary station tables

| Table | Rows | Description |
|---|---|---|
| `beita` | 153,536 | Bait used in longline operations — species (`tegund`), kg |
| `medafli` | 2,708 | Incidental catch of marine mammals and birds |
| `soknartegund` | 2,291,420 | Gear-type breakdown per station; `visir` + `tegund` + `hlutfall` (proportion) |

---

## Family 2 — Electronic logbook staging tables (`rafr_*`) {#sec-rafr}

### `rafr_stofn` — electronic station import table

The staging table for XML-submitted fishing stations. **1,149,278** rows; valid
dates (from `hledsla_dags`, the Oracle import timestamp) 2003–2020 with the
bulk arriving after the 2008 mandatory rollout. Approximately 96.5% of these
rows are also present in `stofn` (negative visir). The principal reason to use
`rafr_stofn` rather than `stofn` directly is its XML traceability columns,
which `stofn` does not carry. Has the same core columns as `stofn` plus:

| Column | Description |
|---|---|
| `xml_faerslu_visir` | Links to the individual FAERSLA within the XML submission |
| `xml_sending_visir` | Links to `afladagb_xml_mottaka.id` |
| `xml_londun_sending_visir` | Links to the XML landing-confirmation sending |
| `hledsla_dags` | Date the record was loaded into Oracle from XML |
| `eytt` | Deletion flag (0 = live, 1 = deleted) |

Coordinates follow the same DDMM integer convention as `stofn`.

### Electronic gear staging subtables

These mirror the paper-era gear subtables and feed into the same master tables
via the negative-visir consolidation. ~97% of `rafr_toga` and `rafr_lineha`
rows reappear in `toga` and `lineha` respectively.

| Staging table | Rows | Master table | Gear |
|---|---|---|---|
| `rafr_toga` | 931,260 | `toga` | Mobile gear |
| `rafr_lineha` | 209,257 | `lineha` | Longline / gillnet |
| `rafr_hringn` | 8,673 | `hringn` | Purse seine |
| `rafr_beita` | 136,067 | `beita` | Bait |
| `rafr_medafli` | 1,807 | `medafli` | Marine mammal / bird bycatch |
| `rafr_afli` | 3,625,502 | `afli` | Catch by species |
| `rafr_soknartegund` | 1,994,085 | `soknartegund` | Gear-type breakdown |

### `rafr_stofn_*` — station detail subtables

A family of ~15 small tables provides additional electronic-era station detail,
each linked to `rafr_stofn` by `visir`. Most have fewer than 100 rows in the
dump, suggesting they record rare events or optional supplementary fields.

| Table | Rows | Description |
|---|---|---|
| `rafr_stofn_fisktegundir` | — | Species recorded for a station |
| `rafr_stofn_fisktegundir_sleppt` | — | Species released / discarded |
| `rafr_stofn_veidarf` | 13 | Gear used per station |
| `rafr_stofn_beita` / `_beita_t` | — / 67 | Bait detail |
| `rafr_stofn_botngerd` | 15 | Bottom type observations |
| `rafr_stofn_sjolag` | 10 | Sea conditions |
| `rafr_stofn_hofn` | 91 | Harbour detail |
| `rafr_stofn_fotreipi` | 7 | Foot-rope detail |
| `rafr_stofn_gerd_kroka` | 8 | Hook type |
| `rafr_stofn_gerd_poka` | 7 | Codend type |
| `rafr_stofn_skiljugerd` | 6 | Sorting grid |
| `rafr_stofn_spendyr_fuglar` | 68 | Marine mammal / bird detail |
| `rafr_stofn_aths_skipstjora` | 45 | Skipper's remarks |
| `rafr_stofn_aths_eftirlit_t` | 13 | Inspector's remarks |

---

## Family 3 — Automatic logger tracks {#sec-logger}

Vessels equipped with automatic position loggers record position, speed and
heading at regular intervals (typically every 10 minutes) throughout a trip.
These data arrive in the XML submission inside `SJALFVIRKIR_MAELAR/LOG` elements
and are stored in two related tables.

### `sjalfvirkir_maelar` vs `rafr_sjalfvirkir_maelar`

Both tables carry all-negative `visir` values and **identical coordinate
values** for overlapping records (685,656 shared `visir`s confirmed by exact
diff). They differ in temporal coverage and column set:

| | `sjalfvirkir_maelar` | `rafr_sjalfvirkir_maelar` |
|---|---|---|
| Rows | 28,234,223 | 23,610,639 |
| Valid `timi` range | 2000–2024 | 2000–2020 |
| Extra columns | `botndypi`, `hofudlinudypi`, `botnhiti`, `hofudlinuhiti`, `vindhradi`, `vindstefna` | fewer |

`sjalfvirkir_maelar` is the later, larger extract — it extends four more years
(2021–2024) and carries additional sensor columns (wind speed/direction, column
temperatures). `rafr_sjalfvirkir_maelar` appears to be an earlier Oracle
extraction that was superseded. **Use `sjalfvirkir_maelar` as the primary
source** for logger track analysis; `rafr_sjalfvirkir_maelar` adds nothing
not already present in the wider table.

Both tables link to `rafr_stofn` via `visir` (negative values throughout).
Unlike the gear subtables, there is no positive-visir master-table equivalent
for logger data — `sjalfvirkir_maelar` is itself the master.

### Wacky coordinate error {#sec-wacky}

Positions in both logger tables are stored as decimal-degree floats, but the
values are incorrect for approximately 96% of records. The error is not a
simple encoding mistake that can be exactly reversed.

**Root cause.** On-board GPS loggers encode positions as DDMMmm integers (degrees,
minutes, hundredths-of-a-minute; e.g. `633367` = 63°33.67′N). The logbook
software (Trackwell / SeaData) converted these to decimal degrees by treating
the last two digits as *seconds* (valid range 0–59) rather than hundredths of a
minute (range 0–99). When the original hundredths value was ≥ 60, the software
rolled the minute counter forward and subtracted 60 from the remainder, shifting
the reported position by a fixed `1/150° ≈ 300–740 m`.

**Evidence that the error is sender-side.** The same XML submission that
contains wacky decimal logger positions also contains DDMM integer coordinates
in the `STOFN/STADSETNING` elements (vessel position at start/end of tow,
entered manually by the skipper). These integer values were stored verbatim in
`rafr_stofn.breidd` / `rafr_stofn.lengd` without any conversion by the
recipient. The logger positions, by contrast, arrive in the XML already as
decimal-degree floats — the erroneous conversion was applied by the logbook
software before transmission. The recipient stored both verbatim; the error
originated upstream.

```{r logger-coord-illustration}
#| echo: false
#| label: tbl-wacky-example
#| tbl-cap: "A single SJALFVIRKIR_MAELAR/LOG entry from submission id 52066. The stored decimal values are the result of the DDMMmm → DMS conversion error applied by the logbook software. The STOFN/STADSETNING entry from the same FAERSLA is stored as an integer in rafr_stofn."
tribble(
  ~Source,          ~`XML node`,                     ~`Stored value`,  ~Format,
  "Logger position", "SJALFVIRKIR_MAELAR/LOG/STADSETNING", "BREIDD='63.3366667'  LENGD='19.4022222'", "Wacky decimal",
  "Station position", "STOFN/STADSETNING",           "BREIDD='6320'  LENGD='1928'",   "DDMM integer"
) |> knitr::kable()
```

**Recovery.** From any wacky stored value $w$ the original DDMMmm components
can be extracted:

$$dd = \lfloor |w| \rfloor, \quad
mm = \lfloor (|w| - dd) \times 60 \rfloor, \quad
ss = \operatorname{round}\bigl(((|w| - dd) \times 60 - mm) \times 60\bigr)$$

where $ss$ is the original hundredths-of-a-minute that was mis-handled. The
ambiguity arises because $ss \in [0, 39]$ is consistent with two interpretations:
the original hundredths could be $ss$ (non-wacky) or $ss + 60$ (wacky with
rollover). Only when $ss \in [40, 59]$ is the solution unique (~4% of records).
A greedy forward-backward smoother using recorded speed and heading resolves
most ambiguous records with a residual positional uncertainty of ~300–740 m.
Full methodology is in `wacky_recovery.qmd`.

**An alternative to algorithmic recovery** is to obtain corrected data directly
from the logbook software vendors (Trackwell / SeaData), who presumably retain
the original DDMMmm integer values before conversion. This would yield exact
positions for all records without reliance on the dead-reckoning smoother.

---

## Family 4 — XML submission pipeline {#sec-xml}

```
Vessel (logbook software)
  │
  │  XML via email / HTTPS
  ▼
afladagb_xml_mottaka (raw XML archive)
  │                     │
  │  Oracle import       │  run log
  ▼                     ▼
rafr_stofn          rafr_hledsla_log
rafr_afli           (import log)
rafr_toga …
  │
  │  monthly compliance check
  ▼
afladagb_skil
afladagb_skil_linur (compliance tracking)
```

### `afladagb_xml_mottaka` — raw XML archive

**483,738** rows; one row per submission. The `sending` column holds the
complete XML document as a character string. Metadata columns:

| Column | Description |
|---|---|
| `id` | Submission row id |
| `skip_nr` | Vessel number |
| `fra` / `til` | Submission period (start / end) |
| `londun_faersla` | Type flag: `"F"` = fishing, `"L"` = landing |
| `hugbunadur` | Logbook software name and version (58 distinct values; predominantly Trackwell variants) |
| `ret_val` | Processing return code |

Earliest valid `fra` timestamps are ~2004–2005; the series extends to 2021-05.
Rows with `fra = 0000-12-30` are Oracle null-date artefacts.

### `afladagb_schema` — XML schema versions

37 rows; stores the XML schema definitions used to validate submissions over
time. Links submissions to their schema version.

### `afladagb_skil` / `afladagb_skil_linur` — compliance records

Introduced January 2014 (per Oracle comment). `afladagb_skil` (9,961 rows) is
the master table — one row per vessel per submission period. `afladagb_skil_linur`
(46,561 rows) is the detail table tracking each fishing month within a period
(`man`, `ar`, `skil` = submission status). Both carry `snn`/`snt`/`sbn`/`sbt`
audit columns (stamped/approved by / timestamps).

### `afladagbok_afhending` — physical logbook delivery

801 rows. Records the handing-over of paper logbooks: vessel, logbook number
range, date, recipient name and address. Relevant for vessels that did not
participate in the digital system. A `tegund_bokar` (book type) field
distinguishes log types.

### `rafr_hledsla_log` / `rafr_hledsla_log_runs`

6,595,396 log rows across 456 import runs. Each run processes a batch of XML
submissions; log entries carry `visir`, `skip_nr`, severity (`INFO` / `ERROR`),
and Icelandic-language message text. Useful for diagnosing import failures;
only 15 rows relate to station-loading errors in the full archive.

### `mottaka` — early catch-receipt register

509,774 rows. Predates the XML system; records vessel-level catches by gear and
fishing period (`fra`, `til`). The `stada` column encodes processing status.
Temporal coverage begins ~1999–2000. Likely used as a cross-check against
reported logbook catch before the digital system was mature.

---

## Family 5 — Reference tables {#sec-reference}

### Species and gear

| Table | Rows | Description |
|---|---|---|
| `fiskar` | 68 | Species code → Icelandic name; `er_yfirflokkur` flags aggregate codes |
| `rafr_oll_stofngogn` | 408 | Lookup for electronic station data types (e.g. "Lína", "Net") |
| `rafr_stofngogn` | 13 | Station data category labels |

### Statistical areas

| Table | Rows | Description |
|---|---|---|
| `skikar` | 1,211 | Statistical area (skiki) codes and names |
| `skil` | 12,987 | Legacy paper submission records; includes `man`/`ar` (month/year) |
| `faer_reitur` | 37 | Statistical square lookup |
| `brei_reitir` | 28 | Broad area definitions |
| `brei_nofn` | 428 | Broad area names |
| `afli_skikar` | 1,470 | Catch aggregated by statistical area |
| `g_reitir` | 182 | Lumpsucker statistical squares |
| `hafsauga` | 266 | Reference list of statistical square codes (`reitur`) that fall in open ocean; used to flag offshore squares. Oracle comment: *"reiti sem eru út í hafsauga"* |
| `nordursvaedi` / `sudursvaedi` | 99 / 92 | Northern / southern area codes |
| `eyjar` | 49 | Island coordinates (DDMM); used to flag near-island positions |
| `land` | 22 | Country codes |
| `stadur` | 96 | Harbour and place codes |

### Quality and ratings

`einkunn` (129,952 rows) stores vessel-level quality ratings (`einkunn`
column) by vessel, year-month (`arman`), and fish type (`fteg`). Used
historically to assess logbook reporting quality — likely not current.

### `togarar` — trawler register

318 rows. A vessel register restricted to trawlers (`nafn_vorpu` = trawl gear
name). Not linked to the core logbook tables by a foreign key; rather a
standing reference for which vessels operated which gear over time.

---

## Family 6 — Specialist subsystems

### Lumpsucker (grasleppa)

A self-contained sub-schema for lumpsucker egg fishery, which has a distinct
seasonal and spatial pattern (coastal, spring, fixed stations).

| Table | Rows | Description |
|---|---|---|
| `g_stofn` | 3,275 | Lumpsucker stations; `skipnr`, `vear`, `stadur`, `gerd_neta`, `moskvi`, `tunnur`, `hrogn` |
| `g_sokn` | 79,880 | Lumpsucker effort by station |
| `grasl_stofn` | 9,633 | Later-era lumpsucker stations |
| `grasl_sokn` | 239,254 | Later-era lumpsucker effort |
| `grasl_verkun` | 4 | Processing records |
| `grasleppureitur` | 359 | Lumpsucker-specific statistical squares |

### Lobster (humar)

`humar` (11,965 rows): trap-based lobster records with columns `visir`,
`gildrur` (traps), `klst` (hours), `breidd`/`lengd` (DDMM position).
Structurally parallel to `gildra` but maintained as a separate table,
reflecting the lobster fishery's distinct regulatory framework.

### Catch processing summaries (`urv_*`)

The `urv_` (úrvinnsla = processing) tables store aggregated catch and
effort by statistical area after editorial review. Gear-specific:

These tables contain a **wide-format denormalized layout**: individual species
columns (`torskur`, `ysa`, `ufsi`, `karfi`, …) rather than the normalised
one-row-per-species structure of `afli`. Each row also carries full station
metadata (`skipnr`, `breidd`, `lengd`, `reitur`, `togtimi`, …), making them
self-contained aggregation tables. `urv_botnv` coverage begins ~1988 (`lar`
column). These are likely the source for official catch statistics by area
rather than the raw `stofn` / `afli` records.

| Table | Rows | Gear |
|---|---|---|
| `urv_botnv` | 2,596,988 | Demersal trawl |
| `urv_dragn` | 999,262 | Seine net |
| `urv_net` | — | Gillnet |
| `urv_lina` | — | Longline |
| `urv_handf` | — | Handline |
| `urv_raekja` | — | Shrimp trawl |

### `steini` — foreign vessel records

74,425 rows. Records trawl operations by vessels flagged as foreign
(`thjod = 6`), with `breidd`/`lengd` in DDMM format, tow duration, and
aggregated catch. Temporal coverage 1995–onwards. Not linked by `visir`
to the main `stofn` table.

### Advisory office (`ra_*`)

Three small tables: `ra_umsoknir_skip` (1,494 rows), `ra_umsoknir_umbodsmenn`
(20 rows), `ra_email_skeyti` (137,391 rows). These appear to relate to an
advisory or quota-application process (`umsoknir` = applications) rather than
logbook operations, and are not linked by `visir` to the fishing tables.

---

## Key join paths {#sec-joins}

The `visir` key links tables within each layer. The consolidation flow runs
upward: XML submissions are ingested into the `rafr_*` staging tables and
simultaneously written to the master tables with negative `visir` values.
The logger tables sit alongside the staging layer and are not consolidated
into a paper-era counterpart.

```
afladagb_xml_mottaka
       │  xml_sending_visir
       ▼
 rafr_stofn ──────────────────► stofn  (negative visir rows)
 rafr_afli  ──────────────────► afli
 rafr_toga  ──────────────────► toga
 rafr_lineha ─────────────────► lineha
 …
 sjalfvirkir_maelar  (visir ←── rafr_stofn; no master-table counterpart)
```

The `stofn.visir` sign convention:

| Sign | Source |
|---|---|
| positive | Paper-era record (transcribed from physical logbook) |
| negative (2003–2020) | Electronic-era record loaded via `rafr_*` staging pipeline |
| negative (2020–2022) | Electronic-era record direct-loaded into `stofn`, bypassing `rafr_*` |

**Minor hand-editing.** A small fraction (~0.7%) of `rafr_stofn` records
that were consolidated into `stofn` show differences in `lengd`, `breidd`,
or `vedags` between the two tables. `stofn` holds the corrected values;
`rafr_stofn` preserves the original electronic submission. The master tables
are therefore the authoritative source for fishing data.

---

## Database artefacts and caveats

- **Null-date artefacts.** Oracle stores NULL datetime as `1899-12-30`; a
  handful of rows in `rafr_stofn`, `rafr_sjalfvirkir_maelar`, and
  `sjalfvirkir_maelar` carry this sentinel. Far-future dates (~2090–2099) also
  appear and likely represent open-ended validity periods.

- **`eytt` (deleted) flag.** `rafr_stofn` carries an `eytt` column (0/1).
  16,912 rows have `eytt = 1` (officially deleted in Oracle). Of these, 327
  passed through the consolidation into `stofn` and appear in the current
  `data/afli/station.parquet` and `data/afli/fishing_sample.parquet` outputs
  (none carry catch). The convert script (`01_afli_convert.R`) now propagates
  this as an `eytt_deleted` logical column: `TRUE` = deleted in Oracle,
  `FALSE` = active in rafr_ era (2003–2020), `NA` = paper era or Phase 4
  direct-load.

- **`snt`/`sng`/`sbt`/`sbg` audit columns.** Nearly all electronic tables
  carry these four columns (stamped-name/time, approved-name/time). They record
  who entered and who approved each row but are not needed for catch analysis.

- **Views (`v_*`).** Five view definitions are present in the dump
  (`v_afli`, `v_stofn`, `v_toga`, `v_net`, `v_rafr_skil`) but were exported
  without data (`num_rows = NA`). They likely represent commonly-used joins
  or filtered subsets used in the Oracle application layer.
