---
title: "Historical back-entries in afli/stofn"
---

## Background

The `afli` schema (`stofn` table) covers Icelandic fisheries logbooks from
approximately 1950 to the present, with ~6.9 million station records after
removing [monthly aggregate records](afli-tables.qmd#aggregate-records).
Because the system was introduced progressively — by gear type, by vessel, and
by voluntary adoption — the earliest records raise an obvious question: which
pre-1975 records are **genuine contemporaneous entries** and which are
**retrospective back-entries** of historical data digitised long after the
fishing events?

This document traces the investigation and explains the derivation of the
`back_entry` flag in `fishing_sample.parquet`.

---

## Setup

```{r}
library(whack)
library(geo)
library(tidyverse)
library(nanoparquet)

dictionary <- read_parquet("data/dictionary.parquet") |> filter(schema == "afli")

source <-
  read_parquet("data-raw/data-dump/afli/stofn.parquet") |>
  wk_translate(dictionary)

# Remove monthly aggregate records (see afli-tables.qmd)
non_agg <- source |>
  filter(
    is.na(comment) |
    (!comment %in% c("Frá OPS$SIGFUS.GAMALL_HUMAR, samantekt",
                     "Frá OPS$SIGFUS.GOMUL_RAEKJA, samantekt") &
     !str_starts(comment, "Frá afli.gomul_dragnot, samantekt"))
  )
```

---

## Step 1 — Where does recording actually start?

The first step is to look at the pre-1975 record count by gear type and date.

```{r}
#| fig-cap: "Pre-1975 station counts by gear. Each panel is one `gid` (gear code). The x-axis spans 1950–1975."
non_agg |>
  count(gid, date) |>
  filter(year(date) < 1975) |>
  ggplot(aes(date, n)) +
  geom_point(size = 0.2) +
  facet_wrap(~ gid, scales = "free_y") +
  labs(x = NULL, y = "stations per day",
       title = "Pre-1975 stations by gear (non-aggregate)")
```

The plot reveals two patterns:

- **Most gear types show a clear recording onset around 1970–1972**, with volume
  increasing steadily from there.
- **Isolated points appear as far back as the early 1950s** — lone dots far from
  the main data cloud.

These isolated early points are the candidates for date entry errors or
back-entries.

---

## Step 2 — Is `.sid` a proxy for data-entry time?

In Oracle-based systems, primary keys are often drawn from a shared sequence
that increments with each insert. If `.sid` tracks insertion order, a genuine
old record should have a *low* `.sid`; a modern record with a mistyped year
(or a retrospective batch insert) would have a *high* `.sid* paired with an
old `date`.

The plot below checks this for all records with `date < 1965`:

```{r}
#| fig-cap: "`.sid` value vs year(date) for all records with date before 1965. High `.sid` + old date = strong evidence of late entry."
non_agg |>
  filter(.sid > 0, !is.na(date), year(date) < 1965) |>
  ggplot(aes(date, .sid)) +
  geom_point(size = 0.8, alpha = 0.6) +
  scale_y_continuous(labels = scales::label_comma()) +
  facet_wrap(~ gid, scales = "free") +
  labs(x = "year(date)", y = ".sid",
       title = "Pre-1965 records: date vs .sid",
       subtitle = "High .sid + old date = likely late entry")
```

All three gear panels show `.sid` values in the **tens of millions** for dates
in 1950–1964. Given that the main recording era (1973–2000) sits roughly in
the 2M–50M `.sid` range, these records were clearly inserted long after the
fishing events.

---

## Step 3 — Is the relationship between `.sid` and date broadly monotone?

To see the full picture, we bin `.sid` into 200 equal-count quantile groups and
plot the date distribution within each bin.

```{r}
#| fig-cap: "Date distribution by `.sid` quantile bin (200 bins). Black = median, grey ribbon = p05–p95, red = p01, blue = p99. A well-behaved contemporaneous system would show a smooth rising curve throughout."
non_agg |>
  filter(.sid > 0, !is.na(date)) |>
  mutate(
    sid_bin   = ntile(.sid, 200),
    year_date = year(date)
  ) |>
  group_by(sid_bin) |>
  summarise(
    sid_med   = median(.sid),
    yr_p01    = quantile(year_date, 0.01),
    yr_p05    = quantile(year_date, 0.05),
    yr_p50    = median(year_date),
    yr_p95    = quantile(year_date, 0.95),
    yr_p99    = quantile(year_date, 0.99),
    .groups   = "drop"
  ) |>
  ggplot(aes(sid_med)) +
  geom_ribbon(aes(ymin = yr_p05, ymax = yr_p95), alpha = 0.2) +
  geom_line(aes(y = yr_p50)) +
  geom_line(aes(y = yr_p01), colour = "firebrick", linewidth = 0.4) +
  geom_line(aes(y = yr_p99), colour = "steelblue", linewidth = 0.4) +
  scale_x_continuous(labels = scales::label_comma()) +
  labs(x = ".sid", y = "year(date)",
       title = "Date distribution by .sid bin",
       subtitle = "Black = median · ribbon = p05–p95 · red = p01 · blue = p99")
```

The median (black) rises broadly with `.sid` — as expected. But the **red p01
line shows a series of dips** at various `.sid` ranges: small clusters of
records with dates far older than anything else in their bin. These are the
back-entry batches.

---

## Step 4 — Characterising the pre-1980 blocks

Filtering to all records with `date < 1980` and sorting by `.sid`, we group
consecutive `.sid` values (gap < 5000) into blocks.

```{r}
pre80 <- non_agg |>
  filter(.sid > 0, !is.na(date), year(date) < 1980) |>
  arrange(.sid) |>
  mutate(
    sid_gap   = .sid - lag(.sid),
    new_block = is.na(sid_gap) | sid_gap > 5000
  ) |>
  mutate(block_id = cumsum(new_block))

catch_raw <- read_parquet("data-raw/data-dump/afli/afli.parquet")

pre80 |>
  group_by(block_id) |>
  summarise(
    n_station  = n(),
    sid_min    = min(.sid),
    sid_max    = max(.sid),
    yr_min     = year(min(date)),
    yr_max     = year(max(date)),
    n_vid      = n_distinct(vid),
    gids       = paste(sort(unique(gid)), collapse = ","),
    .groups    = "drop"
  ) |>
  left_join(
    catch_raw |>
      filter(visir %in% (pre80 |> pull(.sid))) |>
      count(visir) |>
      left_join(pre80 |> select(.sid, block_id), by = c("visir" = ".sid")) |>
      group_by(block_id) |>
      summarise(n_catch_rows = sum(n), .groups = "drop"),
    by = "block_id"
  ) |>
  mutate(
    interpretation = case_when(
      sid_max < 2500000                    ~ "genuine contemporaneous",
      n_station < 100                      ~ "isolated (single-vessel)",
      .default                             = "organised back-entry"
    )
  ) |>
  knitr::kable(
    col.names = c("Block", "n stations", ".sid min", ".sid max",
                  "yr min", "yr max", "n vessels", "gids",
                  "n catch rows", "Interpretation")
  )
```

Key observations:

- **Blocks 1–2** (`.sid` < 2.34M): Genuine contemporaneous records — Blocks 1
  covers the earliest longline pioneers (1969–1979, 15 vessels); Block 2 is
  the main trawl roll-out (1973–1979, 93 vessels, 338k stations).
- **All other blocks** (`.sid` > 2.5M): Organised back-entries — deliberately
  digitised historical data with catch records attached and multi-vessel
  coverage. The largest are the shrimp batch (44k stations, gid 14) and the
  multi-gear batch (13k stations, gids 1/2/5/7/8 going back to 1950).
- **No evidence of random date-entry typos at any meaningful scale.** Every
  cluster of anomalous dates is an organised retrospective batch.

---

## Step 5 — Why a data-driven flag fails

Several algorithmic approaches were tested (bin-median, bin-p90, running
maximum). All failed in at least one case because:

1. Large back-entry blocks (e.g., 44k shrimp records) can dominate their `.sid`
   bin, pulling the reference percentile toward the back-entry dates and
   preventing detection.
2. The `.sid` space is **not globally monotone with respect to date**: records
   with `.sid` 200k–2M have dates from 1988–2019, so a running-max frontier
   immediately overshoots and flags all subsequent blocks including the
   genuine ones.

---

## The `back_entry` flag

The flag uses the single clean boundary identified above:

```r
back_entry = !is.na(date) & year(date) < 1980 & .sid > 2500000
```

- **Genuine early records** (blocks 1–2, `.sid` ≤ 2.34M): correctly left as
  `FALSE`.
- **All 11 back-entry / isolated blocks** (`.sid` > 2.5M, pre-1980 dates):
  correctly flagged `TRUE`.
- **Post-1979 data**: `FALSE` by definition — no back-entry batches were
  detected beyond 1979.

Total records flagged: **95,743** (~1.7% of the station table).

This flag is intended for sensitivity analyses. Analysts working on pre-1980
time series should consider whether to include or exclude back-entered data,
as completeness and precision may differ from contemporaneous records.

---

## Validation

```{r}
non_agg |>
  filter(.sid > 0, !is.na(date)) |>
  mutate(back_entry = !is.na(date) & year(date) < 1980 & .sid > 2500000) |>
  count(back_entry, yr_range = if_else(year(date) < 1980, "pre-1980", "1980+")) |>
  knitr::kable(col.names = c("back_entry", "era", "n"))
```

Zero post-1979 records are flagged (by construction); 338,957 genuine early
pre-1980 records are correctly preserved as `FALSE`.
