---
title: "Convert script audit: station record loss"
date: 2026-04-11
---

## Overview

Both `scripts/01_adb_convert.R` and `scripts/01_fs_afladagbok_convert.R` shared
two bugs that caused a large fraction of fishing stations to be silently dropped
from the output `station.parquet` files.  The bugs were identified by comparing
the `adb` script output against a brute-force reference implementation
(`scripts/01_adb_convert_brute-force.R`), then the same patterns were checked in
the `fs_afladagbok` script.

| Schema | Stations before fix | Stations after fix | Recovered |
|---|---:|---:|---:|
| `adb` | 455,897 | 628,130 | 172,233 |
| `fs_afladagbok` | 350,172 | 469,881 | 119,709 |

---

## Bug 1 — `bind_rows` of inner joins drops stations without aux records

### What happened

Both scripts assembled the `station` table by splitting the base station set
into gear-class blocks, inner-joining each block to a gear-specific auxiliary
parquet (trawl widths, hook counts, etc.), and `bind_rows`-ing the results:

```r
# BEFORE (both scripts)
station <- bind_rows(static, mobile, seine, traps)
#   where each of the above is:  base |> [optional filter] |> inner_join(aux_table)
```

Any station whose `.sid` is absent from **all** auxiliary tables is silently
dropped.  This is not a rare edge case — 25–27% of stations have no matching
gear detail record, particularly in earlier years when those tables were not
yet populated upstream.

### Why stations are absent from aux tables

Absence is a genuine upstream data sparsity issue, not a processing error.  For
the `adb` schema, the trawl auxiliary table (`trawl_and_seine_net_v`) is
completely unpopulated for gid 6 trips in 2020–2021:

```{r}
#| label: tbl-aux-coverage
#| tbl-cap: "gid 6 coverage in trawl_and_seine_net_v by year"
library(nanoparquet)
library(tidyverse)

station_v <- read_parquet("data-raw/data-dump/adb/station_v.parquet")
trawl     <- read_parquet("data-raw/data-dump/adb/trawl_and_seine_net_v.parquet")

station_v |>
  filter(gear_no == 6) |>
  mutate(in_trawl = station_id %in% trawl$station_id,
         year     = year(fishing_start)) |>
  count(year, in_trawl) |>
  tidyr::spread(in_trawl, n, fill = 0) |>
  rename(not_in_trawl = `FALSE`, in_trawl = `TRUE`) |>
  mutate(pct_missing = round(100 * not_in_trawl / (not_in_trawl + in_trawl), 1)) |>
  filter(between(year, 2019, 2026)) |>
  knitr::kable(col.names = c("Year", "Not in trawl aux", "In trawl aux", "% missing"))
```

### Fix

Build compact effort tables for each gear class (inner-joining the aux table
only to compute effort), combine them, and **left-join onto base** so all
stations are retained regardless of aux coverage:

```r
# AFTER (both scripts, illustrated for one gear class)
mobile_aux <-
  read_parquet("...trawl_and_seine_net_v.parquet") |>
  wk_translate(dictionary) |>
  inner_join(base |> select(.sid, gid, t1, t2), by = ".sid") |>
  mutate(towtime = as.numeric(difftime(t2, t1, units = "mins")),
         effort = ..., effort_unit = ...) |>
  select(.sid, effort, effort_unit, towtime, gear_width)

station <-
  base |>
  left_join(bind_rows(mobile_aux, dredge_aux, static_aux, trap_aux, seine_aux),
            by = ".sid") |>
  arrange(date, .sid, t0, t1, t2)
```

Stations without an aux record retain all base columns; effort columns are `NA`.

---

## Bug 2 — `date` derived from a mostly-NA time column

### What happened

Both scripts computed `date` from the middle time column (`t1`), which maps
to `tow_start` (adb) and `milli_timi` / intermediate time (fs_afladagbok).
Both columns are sparsely populated:

| Schema | Column behind `t1` | NA rate |
|---|---|---:|
| `adb` | `tow_start` | ~42 % |
| `fs_afladagbok` | `milli_timi` | 77.6 % |

In the `fs_afladagbok` case this caused **70 % of the output station records to
have `NA` dates**, making year-based filtering and aggregation essentially
useless.

```{r}
#| label: tbl-date-na
#| tbl-cap: "NA date rate before and after fix"
tibble::tribble(
  ~Schema,          ~`Before (NA %)`, ~`After (NA %)`,
  "adb",            "~42",            "0",
  "fs_afladagbok",  "69.9",           "0"
) |>
  knitr::kable()
```

### Fix

Use `coalesce()` across all three time columns, falling back to the start time
(`t0`) which is always populated:

```r
# BEFORE
mutate(date = as_date(t1))

# AFTER
mutate(date = as_date(coalesce(t0, t1, t2)))
```

---

## Additional bugs fixed in `01_adb_convert.R`

### Missing `dredge_v` aux table

The `adb` script had no gear block for `dredge_v.parquet`.  This table covers
gid 5, 6, 7, 51, and 53 stations that are absent from `trawl_and_seine_net_v`
(the two tables are mutually exclusive — zero station_id overlaps):

```{r}
#| label: tbl-dredge
#| tbl-cap: "dredge_v gear code coverage"
dredge <- read_parquet("data-raw/data-dump/adb/dredge_v.parquet")
station_v |>
  filter(station_id %in% dredge$station_id) |>
  count(gear_no) |>
  knitr::kable(col.names = c("gid", "n"))
```

### Wrong gear code filters

The `traps` block filtered on `gid %in% c(18, 39)` — codes from the legacy
`afli` schema.  In the `adb` system trap records carry gid 51 (`trap_v` holds
428 such stations).  The `static` block covered only gid 1–3; gid codes 11,
25, 29, 91, and 92 also appear in `line_and_net_v` and were excluded.

### `difftime(t2, t2, ...)` typo

The `mobile` block computed `towtime = as.numeric(difftime(t2, t2, units = "mins"))`,
yielding zero for every record.  The fix is `difftime(t2, t1, ...)`.

### Dead code

`dragnot_varpa <- read_parquet(...)` at the top of the effort section in
`01_fs_afladagbok_convert.R` read the file into a variable that was never
used.  The `knitr::kable()` diagnostic call that reported the bug was also
removed.

---

## Year-by-year comparison

### `adb` schema

```{r}
#| label: tbl-adb-years
#| tbl-cap: "adb station counts by year: brute-force reference vs fixed script"
library(nanoparquet)
library(tidyverse)

bind_rows(
  read_parquet("data/adb/station_new-brute-force.parquet") |>
    mutate(year = year(date), source = "brute-force"),
  read_parquet("data/adb/station.parquet") |>
    mutate(year = year(date), source = "fixed")
) |>
  count(year, source) |>
  filter(between(year, 2019, 2026)) |>
  tidyr::spread(source, n, fill = 0) |>
  mutate(diff = `fixed` - `brute-force`) |>
  knitr::kable()
```

### `fs_afladagbok` schema

```{r}
#| label: tbl-fs-years
#| tbl-cap: "fs_afladagbok station counts by year after fix"
read_parquet("data/fs_afladagbok/station.parquet") |>
  mutate(year = year(date)) |>
  count(year) |>
  filter(between(year, 2016, 2026)) |>
  knitr::kable()
```
