---
title: "Datacall spatial strategy"
subtitle: "Assigning 566M AIS pings to ICES areas, depth classes, and seabed habitats"
date: today
format:
  html:
    code-fold: true
    toc: true
---

## Background

The ICES VMS datacall requires each AIS ping to be assigned to three spatial
reference objects:

| Object | Variable | Features |
|---|---|---|
| ICES statistical areas | `ices_area` | 66 |
| ICES/GEBCO depth classes | `depth_class` | 9 |
| EmodNet seabed habitats | `MSFD_BBHT` | 2.18 M |

The AIS dataset (`data/trail`) contains ~566 million pings spanning 2007–2026.
A naive spatial join of this many points against these polygon layers is not
feasible — for reasons that differ across the three objects and that took some
time to fully diagnose.

This document describes what was tried, why each approach failed or was
abandoned, and what was ultimately adopted. Reproducible examples of all
approaches are collected in the appendix.

## The three objects and their problems

### ICES statistical areas (`ia`)

66 polygons, **1.45 million vertices**. The polygons trace national coastlines
at sub-metre detail in places — mean vertex spacing 63–68 m in the densest
areas (Irish Sea, Celtic Sea). This is finer than AIS positional accuracy
(10–100 m) and entirely unnecessary for assigning statistical area labels.
The vertex density makes any direct per-ping geometry evaluation unworkable.

### ICES/GEBCO depth classes

9 polygons, **7.59 million vertices** — ~843,000 per feature. Same structural
problem as `ia`, only more concentrated.

### EmodNet seabed habitats (`eusm`)

2.18 million polygons, ~107 million total vertices. The per-feature vertex
count is actually low (~5 vertices each), but the feature count makes any
per-feature operation impractical. Inspection of individual features reveals
the origin: cell dimensions consistently 25–115 m, exactly 5 vertices per
feature — the unmistakable signature of a raster polygonized into rectangles.
The raster, had it been retained, would have been trivial to query.

## The solution

All three objects are rasterized to the **same 0.005° template** covering the
ICES domain bounding box, then each raster cell centre is assigned an H3
resolution-9 cell ID. The three resulting `(h3 → label)` lookup tables are
joined into a single parquet file.

```{r}
#| eval: false
library(sf)
library(terra)
library(tidyverse)
library(h3)

zoom <- 9L

ia_sf      <- read_sf("data-raw/aux/ices_areas.fgb")
ices_gebco <- readRDS("data-raw/aux/ICES_GEBCO.rds")

# Common raster template: ICES domain at 0.005° (~500 m)
r <- rast(ext(vect(ia_sf)), resolution = 0.005, crs = "EPSG:4326")
```

A single helper handles all three objects identically:

```{r}
#| eval: false
# Encode a character column as integer, rasterize to template, extract H3
# lookup. Returns a two-column tibble: h3, <label_col>.
rast_to_h3 <- function(sf_obj, label_col, raster_template, res) {
  levels_tbl <- tibble(label = sort(unique(sf_obj[[label_col]]))) |>
    mutate(id = row_number())
  sf_obj <- sf_obj |>
    mutate(id = levels_tbl$id[match(.data[[label_col]], levels_tbl$label)])
  r_out <- rasterize(vect(sf_obj), raster_template, field = "id", fun = "min") |>
    trim()
  as.data.frame(r_out, xy = TRUE, na.rm = TRUE) |>
    as_tibble() |>
    rename(lon = x, lat = y, id = 3) |>
    left_join(levels_tbl, by = "id") |>
    mutate(h3 = geo_to_h3(data.frame(lat, lon), res = res)) |>
    count(h3, label) |>
    slice_max(n, n = 1, with_ties = FALSE, by = h3) |>
    select(h3, label) |>
    rename(!!label_col := label)
}

# ICES statistical areas
ia_h3 <- rast_to_h3(ia_sf, "Area_Full", r, zoom) |>
  rename(ices_area = Area_Full)

# ICES/GEBCO depth classes
gebco_h3 <- rast_to_h3(ices_gebco, "depth", r, zoom) |>
  rename(depth_class = depth)

# eusm — validate and crop to ICES domain first
sf::sf_use_s2(FALSE)
eusm <- readRDS("data-raw/aux/eusm.rds") |>
  rename(geometry = Shape) |>
  filter(MSFD_BBHT != "") |>
  st_make_valid() |> st_make_valid()
eusm <- eusm[st_is_valid(eusm$geometry), ] |>
  st_crop(st_bbox(ia_sf)) |>
  st_cast("MULTIPOLYGON")

eusm_h3 <- rast_to_h3(eusm, "MSFD_BBHT", r, zoom)

# Combine into single lookup
h3_lookup <-
  ia_h3 |>
  full_join(gebco_h3, by = "h3") |>
  full_join(eusm_h3,  by = "h3")

nanoparquet::write_parquet(h3_lookup, "data/aux/h3_lookup.parquet")
```

The DuckDB join to tag AIS pings — one join, three variables:

```{r}
#| eval: false
library(duckdbfs)

trail   <- open_dataset("data/trail")
h3_look <- open_dataset("data/aux/h3_lookup.parquet")

trail |>
  mutate(h3 = h3_latlng_to_cell_string(lat, lon, 9L)) |>
  left_join(h3_look, by = "h3") |>
  count(ices_area, depth_class, MSFD_BBHT) |>
  collect()
```

### Why the same raster template for all three?

An earlier design used `polyfill()` for `ia` and `ICES_GEBCO` (exact polygon
coverage, no resolution artefact) and rasterize only for `eusm`. This was
rejected on **consistency** grounds: the same 0.005° resolution and the same
H3 centroid extraction logic applies uniformly to all three. The theoretical
precision advantage of polyfill is irrelevant at this scale — AIS positional
uncertainty is 10–100 m, results are aggregated to 0.5° ICES rectangles, and
a ~500 m boundary artefact affects a negligible fraction of pings on area
edges. Furthermore, polyfill proved slow in practice — see Appendix A.3.

### Why not work directly in H3 space?

An H3-first approach would generate all H3 res-9 cells covering the ICES
domain upfront, extract their centre coordinates, and do point-in-polygon
lookups against each layer. The problem is scale: the ICES domain bounding
box at H3 resolution 9 (~0.105 km² per cell) contains on the order of
hundreds of millions of cells, most of which cover open ocean. Generating
them, extracting centres, and running point-in-polygon joins against complex
polygon layers reintroduces the spatial join problem. `terra::rasterize()`
avoids this because it uses scan-line rasterization — a fundamentally
different and much faster algorithm than point-in-polygon testing.

---

## Appendix: approaches tried

### A.1 Direct DuckDB spatial join

The natural starting point. Loads the polygon layer as a spatial dataset and
uses `duckdbfs::spatial_join()` to tag each ping.

```{r}
#| eval: false
library(sf)
library(duckdbfs)

ia_sf <- read_sf("data-raw/aux/ices_areas.fgb")
tmp   <- tempfile(fileext = ".fgb")
write_sf(ia_sf, tmp, driver = "FlatGeobuf", quiet = TRUE)
ia_tbl <- open_dataset(tmp, format = "sf")

trail <- open_dataset("data/trail")

# Does not complete — 1.45M vertices in R-tree, tested against 2.9M pings/year
trail |>
  filter(year == 2007) |>
  mutate(geom = ST_POINT(lon, lat)) |>
  duckdbfs::spatial_join(ia_tbl, by = "st_intersects", join = "left") |>
  count(Area_Full) |>
  collect()
```

**Result**: does not complete on a single year (2.9M pings). The R-tree must
evaluate each ping against geometries with thousands of vertices per polygon.

### A.2 `ms_simplify` + direct spatial join

`rmapshaper::ms_simplify()` reduces vertices using the Visvalingam-Whyatt
algorithm while preserving shared boundaries between adjacent polygons.

```{r}
#| eval: false
library(rmapshaper)

ia_ms <- ia_sf |>
  st_transform(3035) |>
  ms_simplify(keep = 0.05, keep_shapes = TRUE) |>
  st_cast("MULTIPOLYGON") |>
  st_transform(4326)

cat("Vertices before:", nrow(st_coordinates(ia_sf)), "\n")
cat("Vertices after: ", nrow(st_coordinates(ia_ms)), "\n")

tmp_ms <- tempfile(fileext = ".fgb")
write_sf(ia_ms, tmp_ms, driver = "FlatGeobuf", quiet = TRUE)

system.time(
  open_dataset("data/trail") |>
    filter(year == 2007) |>
    mutate(geom = ST_POINT(lon, lat)) |>
    duckdbfs::spatial_join(open_dataset(tmp_ms, format = "sf"),
                           by = "st_intersects", join = "left") |>
    count(Area_Full) |>
    collect()
)
```

**Result**: 95% vertex reduction (1.45M → ~73K), join completes in ~9.5
s/year (~3 min for full dataset). Rejected because `ms_simplify` moves
coastline boundaries in arbitrary directions including seaward — near-shore
pings from Icelandic coastal and lumpfish fisheries could be silently dropped.
For `ICES_GEBCO` (7.59M vertices), `ms_simplify` with `sys = FALSE` raises
`Error: R character strings are limited to 2^31-1 bytes` — the GeoJSON
serialisation of the object exceeds the 2 GB R string limit even before
simplification begins.

### A.3 H3 `polyfill()` per feature

`h3::polyfill()` fills each polygon with all H3 cells whose centres fall
inside it, building a hash-joinable lookup without any geometry at query time.

```{r}
#| eval: false
library(h3)

zoom <- 9L

# Helper: polyfill one sf row, return data frame. NULL if no cells (feature
# smaller than one H3 cell at this resolution).
polyfill_feature <- function(sf_row, label_col, label_val, res) {
  cells <- polyfill(sf_row, res = res)
  if (length(cells) == 0) return(NULL)
  data.frame(h3 = cells, label = label_val) |>
    setNames(c("h3", label_col))
}

# polyfill() requires an sf object (not a bare sfg geometry), and does not
# accept MULTIPOLYGON in h3 v3.7.2 — cast to POLYGON first
ia_poly <- ia_sf |> st_cast("POLYGON")

ia_h3_polyfill <- map(
  seq_len(nrow(ia_poly)),
  \(i) polyfill_feature(ia_poly[i, ], "ices_area", ia_poly$Area_Full[i], zoom)
) |> bind_rows()
```

**Result**: `polyfill()` at resolution 9 fills each H3 cell (~174 m edge)
individually. Large ICES areas covering hundreds of thousands of km² contain
millions of H3 cells. Processing all 66 areas did not complete in reasonable
time. "Few features" and "fast" are not the same thing when the features are
large. Rejected.

### A.4 Buffer + simplify → direct spatial join

Buffer each polygon 500 m outward (metric CRS), then simplify to 500 m
tolerance. Shifting coastal boundaries inland excludes on-land wacky
coordinates while retaining legitimate near-shore pings; simplification is
geometrically consistent with the buffer shift; internal area-to-area
boundaries (straight lines) are unaffected.

```{r}
#| eval: false
ia_buf <- ia_sf |>
  st_transform(3035) |>
  st_buffer(dist = 500, nQuadSegs = 4) |>
  st_simplify(dTolerance = 500, preserveTopology = TRUE) |>
  st_cast("MULTIPOLYGON") |>
  st_transform(4326)

cat("Vertices before:", nrow(st_coordinates(ia_sf)), "\n")
cat("Vertices after: ", nrow(st_coordinates(ia_buf)), "\n")

write_sf(ia_buf, "data/aux/ia_buf.fgb", driver = "FlatGeobuf", quiet = TRUE)

system.time(
  open_dataset("data/trail") |>
    filter(year == 2007) |>
    mutate(geom = ST_POINT(lon, lat)) |>
    duckdbfs::spatial_join(open_dataset("data/aux/ia_buf.fgb", format = "sf"),
                           by = "st_intersects", join = "left") |>
    count(Area_Full) |>
    collect()
)
```

**Result**: 96% vertex reduction (1.45M → ~52K), join completes in ~2 s/year
(~1 min for full dataset). This works, but was superseded by the raster
approach for two reasons: (1) it requires per-object tuning (buffer distance,
simplification tolerance), and (2) it cannot be applied to `eusm` (2.18M
features — `st_simplify` does not preserve shared boundaries between features,
and `ms_simplify` hits the 2 GB string limit).

### A.5 H3 polygon centroid lookup for `eusm`

Since `eusm` is a complete tessellation (one polygon per ~100 m grid cell),
compute the H3 cell for each polygon's centroid and build a lookup. Avoids
any per-ping geometry operations.

```{r}
#| eval: false
eusm_centroid_h3 <- eusm |>
  st_centroid() |>
  mutate(
    lon = st_coordinates(geometry)[, 1],
    lat = st_coordinates(geometry)[, 2]
  ) |>
  st_drop_geometry() |>
  mutate(h3 = geo_to_h3(data.frame(lat, lon), res = zoom)) |>
  count(h3, MSFD_BBHT) |>
  slice_max(n, n = 1, with_ties = FALSE, by = h3) |>
  select(h3, MSFD_BBHT)
```

**Result**: Runs, but produces incorrect assignments for large irregular
polygons (e.g. deep bathyal rock classes covering hundreds of km²) whose
centroid may fall in a geometrically unrepresentative part of the polygon, or
even outside it for concave shapes. The raster approach samples the polygon
uniformly across its entire area, producing a more robust majority vote.
Rejected.

### A.6 `eusm` direct simplification

`ms_simplify` with `sys = TRUE` (routes through system mapshaper, bypassing
the 2 GB R string limit):

```{r}
#| eval: false
library(rmapshaper)

eusm_simple <- eusm |>
  st_transform(3035) |>
  ms_simplify(keep = 0.25, keep_shapes = TRUE,
              snap_interval = 200,
              explode = TRUE,
              sys = TRUE) |>
  st_transform(4326) |>
  st_cast("MULTIPOLYGON")

cat("Vertices after simplification:", nrow(st_coordinates(eusm_simple)), "\n")
```

**Result**: even with `sys = TRUE`, returns `Error: R character strings are
limited to 2^31-1 bytes`. The 107M-vertex object cannot be serialised to
GeoJSON for mapshaper even via the system pathway. `st_simplify` is not a
substitute because it processes each feature independently, breaking shared
boundaries between adjacent habitat polygons. Rejected.

EOQMD 2>&1
