---
title: "Linking Landings to Logbook Trips"
date: today
---

## Goal

Attach a landing identifier (`.lid` from `aflagrunnur`) to each logbook trip (`.tid`
in `data/merged/trip.parquet`), enabling downstream catch-weight reconciliation between
the two systems. The natural join key is vessel × landing date × port.

## Background {#sec-background}

Understanding the match results requires knowing how the two systems evolved.

**The `afli` era (pre-2021).** Electronic logbooks were introduced during the `afli`
schema phase, not at the transition to `fs_afladagbok`. The government contracted a
single company (Trackwell) to handle all e-logbook submissions, and larger vessels
progressively adopted this system. Smaller vessels — those targeting lumpfish, coastal
jig fisheries, and similar — remained on **paper** throughout the `afli` era. A single
government employee was responsible for manually entering paper logbooks and loading
both paper and electronic records into the legacy database structure (`stofn`, `toga`,
`lineha`, …). The steady improvement in logbook-landing match rates from 2008 to 2018
is at least partly attributable to his active curation. His retirement around 2019–2020
is a plausible explanation for the subsequent decline.

**Spring 2021: market opening.** Until early 2021, Trackwell held an exclusive
government contract. Opening the market to multiple vendors was abrupt — a structural
break in data quality, not a gradual phase-in. This "wild west" period is reflected in
inconsistent harbour codes, incomplete timing records, and other QC issues that
characterise the `fs_afladagbok` era. Crucially, the overlap period between `afli` and
`fs_afladagbok` (2021–2022) falls squarely within this window, which is why the merge
rule **`afli` wins over `fs_afladagbok`** for duplicate trips is a quality decision, not
a statement about paper vs. electronic recording.

**Harbour coding.** `fs_afladagbok` stores harbour as `hafnarnumer_id` (a sequential
1–140 index into the `aflagrunnur` lookup), while `afli` stores `hafnarnumer` directly
(the numeric code in `aflagrunnur`). Both are bridged to a common port identifier (`pid`)
via `data/ports/ports.gpkg`, eliminating the schema-specific coding difference before
matching.

## Data {#sec-data}

```{r}
#| label: setup
library(arrow)
library(sf)
library(duckdbfs)
library(tidyverse)
```

```{r}
#| label: load-harbours
# Port reference: pid + port name, keyed on hid (= hafnarnumer in aflagrunnur)
# ports <- read_sf("data/ports/ports.gpkg") |> st_drop_geometry()
```

```{r}
#| label: load-land
# Landing records: one row per (vid × date × pid × port); catch summed.
# Multiple .lids on the same key are collapsed to min(.lid); n_.lid records how
# many were merged.
# here stuff a bit premature - should really do the first step in the 
#. landings project
ports <-
  read_parquet("data/ports/hafnarnumerakerfid.parquet")

land <-
  read_parquet("data-dump/landings/agf/aflagrunnur.parquet") |>
  # drop out foreigners, for now
  filter(skip_numer > 0, !between(skip_numer, 3700, 4999)) |>
  # align with the last fs_logbook dump
  filter(londun_hefst <= ymd("2025-12-12")) |>
  left_join(ports,
            by = join_by(hafnarnumer == hid)) |>
  # left_join(ports, by = join_by(hafnarnumer == hid)) |>
  select(.lid = londun_id, date = londun_hefst, vid = skip_numer,
         hid = hafnarnumer, port, catch = magn_oslaegt) |>
  group_by(.lid, vid, date, hid, port) |>
  summarise(catch = sum(catch, na.rm = TRUE), .groups = "drop") |>
  filter(catch > 0) |>
  group_by(vid, date, hid, port) |>
  summarise(.lid = min(.lid), catch = sum(catch), n_.lid = n(), .groups = "drop") |>
  mutate(date = as.Date(date))
```

`n_.lid > 1` flags keys where multiple landing records were collapsed into one. Those
represent either split deliveries or gear-disaggregated entries in the source database.

```{r}
#| label: tbl-n-lid
#| tbl-cap: "Proportion of landing keys with multiple source records."
land |>
  count(n_.lid) |>
  mutate(pct = round(100 * n / sum(n), 1))
```

```{r}
#| label: load-trip
# Trips with total logbook catch; harbour translated to hid/port for both schemas.
trip <-
  read_parquet("data/logbooks/trip.parquet") |>
  mutate(date = as_date(T2)) |>
  filter(date >= ymd("2007-09-01"), date <= ymd("2025-12-12")) |>
  filter(vid > 0, !between(vid, 3700, 4999)) |>
  left_join(
    read_parquet("data/logbooks/catch.parquet") |>
      left_join(
        read_parquet("data/logbooks/station.parquet") |> select(.sid, .tid, schema),
      ) |>
      filter(!is.na(.tid)) |>
      group_by(.tid) |>
      summarise(catch = sum(as.numeric(catch), na.rm = TRUE), .groups = "drop"),
    by = ".tid"
  ) |>
  filter(catch > 0) |>
  select(.tid, vid, date, hid = hid2, port, catch, schema)
```

## Matching {#sec-matching}

### Strategy {#sec-strategy}

The join key is `vid × date × port (hid)`. Known sources of mismatch:

- **Date lag** — `londun_hefst` in the landing register may lag the logbook's `T2` by a
  day or two due to processing or recording delays.
- **Port discrepancy** — planned vs. actual landing port sometimes differ.

A three-step cascade, accepting only unambiguous matches (exactly one `.lid` per `.tid`)
at each step:

| Step | Keys | Notes |
|---|---|---|
| 1 | `vid + date + hid + port` | exact; unambiguous |
| 2 | `vid + date` | port dropped; unambiguous |
| 3 | `vid + date ± 1 day + hid + port` | date relaxed by one day; unambiguous |

### Cascade function {#sec-cascade-fn}

```{r}
#| label: cascade-fn
# One cascade step: inner-join, keep only unambiguous matches (1 .lid per .tid).
try_match <- function(unmatched, land_tbl, by, type) {
  unmatched |>
    inner_join(land_tbl, by = by) |>
    group_by(.tid) |>
    filter(n_distinct(.lid) == 1) |>
    slice(1) |>
    ungroup() |>
    mutate(match_type = type)
}
```

### Cascade {#sec-cascade}

```{r}
#| label: cascade
m1 <- try_match(
  trip, 
  land, 
  c("vid", "date", "hid", "port"), 
  "01 exact"
)


# ±1 day: expand land by shifting dates one day in each direction
land_pm1 <- bind_rows(
  land |> mutate(date = date - 1),
  land |> mutate(date = date + 1)
)

m2 <- try_match(
  trip |> filter(!.tid %in% m1$.tid),
  land_pm1 |> filter(!.lid %in% m1$.lid),
  c("vid", "date", "hid", "port"), 
  "02 exact ±1 day"
)

m3 <- try_match(
  trip |> filter(!.tid %in% c(m1$.tid, m2$.tid)),
  land |> filter(!.lid %in% c(m1$.lid, m2$.lid)),
  c("vid", "date"), 
  "03 vid_date"
)



lid_map <- bind_rows(m1, m2, m3) |>
  select(.tid, .lid, match_type)
```

```{r}
#| label: tbl-cascade
#| tbl-cap: "Match counts by step and schema."
trip |>
  left_join(lid_map, by = ".tid") |>
  mutate(match_type = replace_na(match_type, "04 unmatched")) |>
  count(schema, match_type) |>
  group_by(schema) |>
  mutate(pct = round(100 * n / sum(n), 1)) |>
  arrange(schema, match_type)
```

```{r}
#| label: trip-with-lid
# Attach .lid and match_type to the trip table
trip <- trip |> left_join(lid_map, by = ".tid")
```

## Annual coverage {#sec-annual}

Both the logbook system and the landing register are intended to function as **complete
censuses**: every fishing trip should produce a logbook entry, and every delivery should
appear in the landing register. The figures below show how far from that ideal we are and
where the gaps lie.

```{r}
#| label: annual-setup
logbook_yr <- trip |>
  mutate(year = year(date),
         match_type = replace_na(match_type, "04 unmatched")) |>
  filter(year >= 2008, year <= 2025) |>
  group_by(year, match_type) |>
  summarise(n = n(), catch_kt = sum(catch, na.rm = TRUE) / 1e6, .groups = "drop")

land_yr <- land |>
  left_join(lid_map |> select(.lid, match_type), by = ".lid") |>
  mutate(year = year(date),
         match_type = replace_na(match_type, "04 unmatched")) |>
  filter(year >= 2008, year <= 2025) |>
  group_by(year, match_type) |>
  summarise(n = n(), land_kt = sum(catch, na.rm = TRUE) / 1e6, .groups = "drop")
```

```{r}
#| label: fig-annual-trips
#| fig-cap: "Annual trip counts in the logbook (lower) and landing records (upper), 2008–2025. Blue tones: matched; orange: matched on vessel + date only; red: unmatched."
bind_rows(
  logbook_yr |> select(year, match_type, n) |> mutate(source = "Logbook trips"),
  land_yr    |> select(year, match_type, n) |> mutate(source = "Landing records")
) |>
  ggplot(aes(year, n / 1e3, fill = match_type)) +
  geom_col(position = position_stack(reverse = TRUE)) +
  facet_wrap(~source, ncol = 1) +
  labs(x = NULL, y = "Thousands of records") +
  scale_fill_brewer(palette = "Set1")
```

```{r}
#| label: fig-annual-weight
#| fig-cap: "Annual catch weight in the logbook (lower) and landing weight (upper), 2008–2025."
bind_rows(
  logbook_yr |> select(year, match_type, kt = catch_kt) |> mutate(source = "Logbook catch"),
  land_yr    |> select(year, match_type, kt = land_kt)  |> mutate(source = "Landing weight")
) |>
  ggplot(aes(year, kt, fill = match_type)) +
  geom_col(position = position_stack(reverse = TRUE)) +
  facet_wrap(~source, ncol = 1) +
  scale_fill_brewer(palette = "Set1") +
  labs(x = NULL, y = "kt (thousand tonnes)")
```

### What the figures show

**Trip counts (@fig-annual-trips).** In the `afli` era (2008–2020), logbook and landing
totals are broadly similar in scale — on the order of 40–65k records per year on each
side — but a non-trivial fraction on each side fails to find a counterpart. On the
**landing side**, roughly 15–20% of records per year have no matching logbook trip. On the
**logbook side**, roughly 5% of trips have no matching landing. For a system that is
supposed to be a complete census on both sides, these gaps are surprisingly persistent.

The unmatched landings belong disproportionately to fisheries that have historically
operated outside the routine logbook system: lumpfish (Grásleppunet / Rauðmaganet),
coastal pole-and-line (sjóstöng), and a small number of seaweed deliveries (which have
no logbook requirement). These are small-vessel fisheries that remained on paper logbooks
throughout the `afli` era, relying on manual digitisation by a single government employee
whose retirement around 2019–2020 coincides with a visible increase in unmatched landings.

From 2021 onward the picture shifts: logbook records decline as `afli` winds down, and
the proportion of unmatched logbook trips grows — consistent with the data-quality
deterioration of the "wild west" period following the market opening.

**Catch weight (@fig-annual-weight).** Landing weight is systematically higher than
logbook catch in most years. Part of this gap is structural: landing records exist for
fisheries (lumpfish, seaweed, shellfish) that are either absent from or severely
under-represented in the logbooks. In pelagic fisheries, catch transfers between vessels
at sea mean that a catching vessel's logbook records the full haul while the weight
appears under the receiving vessel's landing — not a data quality issue, but a structural
feature.

## Trip-level weight comparison {#sec-weight}

For matched trips, `catch` in the logbook and `catch` in `land` are both in kg,
enabling a direct comparison.

```{r}
#| label: tbl-weight-ratio
#| tbl-cap: "Ratio of logbook catch to landing weight for matched trips, by schema and match type. Trips with catch > 0 on both sides."
trip |>
  filter(!is.na(.lid)) |>
  inner_join(land |> select(.lid, land_catch = catch), by = ".lid") |>
  filter(catch > 0, land_catch > 0) |>
  mutate(ratio = catch / land_catch) |>
  group_by(schema, match_type) |>
  summarise(
    n      = n(),
    p25    = round(quantile(ratio, 0.25), 3),
    median = round(median(ratio), 3),
    p75    = round(quantile(ratio, 0.75), 3),
    mean   = round(mean(ratio), 3),
    .groups = "drop"
  )
```

The median ratio close to 1.0 confirms that logbook `catch` and landing `catch` are in
the same units (kg) throughout and that the overall agreement for matched trips is
genuine. Elevated **mean** values relative to the median reflect a right tail driven by
pelagic trips where catch is transferred to a receiving vessel: the catching vessel's
logbook records the full haul but no corresponding landing appears under its own
identifier.

## Coverage: the view from landings {#sec-reverse}

The forward analysis asks which trips lack a landing. Reversing the question — which
landings lack a logbook trip? — is more informative for identifying systematic gaps.
This section focuses on 2008–2020, when `afli` is the primary schema.

```{r}
#| label: reverse-setup
land_afli <- land |>
  #filter(date >= ymd("2008-01-01"), date <= ymd("2020-12-31")) |>
  mutate(year = year(date), matched = .lid %in% lid_map$.lid)

cat("Landings 2008-2020:", nrow(land_afli), "\n")
cat("Matched to logbook:", sum(land_afli$matched),
    sprintf("(%.1f%%)\n", 100 * mean(land_afli$matched)))
```

### By year

```{r}
#| label: fig-coverage-year
#| fig-cap: "Fraction of landings matched to a logbook trip, by year. Dashed lines mark the curator retirement (~2019–2020) and market opening (spring 2021)."
land_afli |>
  count(year, matched) |>
  group_by(year) |>
  mutate(pct = 100 * n / sum(n)) |>
  filter(matched) |>
  ggplot(aes(year, pct)) +
  geom_col() +
  geom_vline(xintercept = c(2019.5, 2021), linetype = "dashed", colour = "grey50") +
  labs(x = NULL, y = "Landings matched (%)")
```

Coverage improved from ~73% in 2009–2010 to ~87% by 2017–2018, then fell to ~72% in
2020. The arc is more plausibly explained by institutional factors — the curator and the
market opening — than by any change in fishing behaviour or vessel compliance.

### By vessel

```{r}
#| label: tbl-vessel-coverage
#| tbl-cap: "Landing frequency and logbook coverage by vessel coverage group (vessels with ≥10 landings, 2008–2020)."
by_vessel <- land_afli |>
  group_by(vid) |>
  summarise(
    n_landings  = n(),
    n_matched   = sum(matched),
    pct_matched = round(100 * n_matched / n_landings, 1),
    .groups = "drop"
  ) |>
  filter(n_landings >= 10)

low_cov_vids  <- by_vessel |> filter(pct_matched <  50) |> pull(vid)
high_cov_vids <- by_vessel |> filter(pct_matched >= 90) |> pull(vid)

by_vessel |>
  mutate(coverage_group = case_when(
    pct_matched <  50 ~ "< 50%",
    pct_matched <  90 ~ "50–90%",
    TRUE              ~ "≥ 90%"
  )) |>
  group_by(coverage_group) |>
  summarise(
    n_vessels        = n(),
    median_landings  = median(n_landings),
    total_landings   = sum(n_landings),
    pct_of_unmatched = round(
      100 * sum(n_landings - n_matched) / sum(land_afli$matched == FALSE), 1
    ),
    .groups = "drop"
  )
```

The distribution is strongly bimodal. Low-coverage vessels (<50%) have a median of
roughly 4–5× fewer landings than well-covered vessels. This reflects the **paper vs.
electronic split**: larger, more active vessels adopted Trackwell e-logbooks early and
are consistently represented in `afli`; smaller vessels (lumpfish, coastal jig, sjóstöng)
remained on paper, relying on manual digitisation that declined with the curator's
retirement.

### By harbour

```{r}
#| label: tbl-harbour-coverage
#| tbl-cap: "Harbours with lowest logbook coverage (≥50 landings, 2008–2020)."
land_afli |>
  group_by(port) |>
  summarise(
    n_landings      = n(),
    pct_matched     = round(100 * mean(matched), 1),
    pct_low_cov_vid = round(100 * mean(vid %in% low_cov_vids), 1),
    .groups = "drop"
  ) |>
  filter(n_landings >= 50) |>
  arrange(pct_matched) |>
  slice_head(n = 12)
```

Most low-coverage harbours are explained by low-compliance vessels clustering there.
Harbours near zero coverage that are not explained by low-coverage vessels likely reflect
fisheries with no logbook requirement (seaweed, shellfish).

### By gear type

```{r}
#| label: tbl-gear-coverage
#| tbl-cap: "Logbook coverage by gear type (landings side, 2008–2020, ≥100 landings)."
d <- read_parquet("data-dump/landings/agf/aflagrunnur.parquet") |> 
  select(.lid = londun_id, date = londun_hefst,
         gid_land = veidarfaeri) |>
  distinct() |>
  #filter(date >= ymd("2008-01-01"), date <= ymd("2020-12-31")) |>
  right_join(land_afli |> select(.lid, matched), by = ".lid")

d |>
  group_by(gid_land) |>
  summarise(n = n(), pct_matched = round(100 * mean(matched), 1), .groups = "drop") |>
  filter(n >= 100) |>
  arrange(-pct_matched) |>
  select(gid_land, n, pct_matched) |> 
  knitr::kable()
d |> 
  mutate(year = year(date)) |> 
  count(year, gid_land, matched) |>
  ggplot(aes(year, n, fill= matched)) +
  geom_col() +
  facet_wrap(~ gid_land, scales = "free_y")
```

Two gear types are nearly absent from logbooks:

- **Sjóstöng** (`gid=19`, sea-rod): <1% coverage across ~20k landings — coastal
  pole-and-line, likely exempt from logbook requirements.
- **Grásleppunet / Rauðmaganet** (`gid=3/4`, lumpfish gillnets): 26% and 50% across
  ~60k combined landings. Both target lumpfish and are the largest single contributor to
  unmatched landings in absolute terms.

::: {.callout-warning}
### Lumpfish logbook records are unreliable

Grásleppunet (`gid=3`) and Rauðmaganet (`gid=4`) together represent ~60k landings
2008–2020 but match at only 26% and 50% respectively. The logbook records that do exist
for lumpfish are known to be problematic in other respects (seasonal timing, gear
reporting). **Do not use lumpfish logbook data in catch reconciliation without dedicated
QA.** Flagged as outstanding work.
:::

## Outstanding issues {#sec-outstanding}

::: {.callout-note}
### Collapsed landing records (`n_.lid > 1`)

Where multiple `.lid` values share the same `vid × date × hid × port` key, the smallest
`.lid` is retained and totals are summed. This occurs when a single delivery is recorded
as multiple rows (e.g., split by gear type) in `aflagrunnur`. Currently ~X% of keys are
affected; the collapsed record should be validated before downstream weight reconciliation.
:::

::: {.callout-warning}
### Duplicate trips in `fs_afladagbok`

Some `vid × date × port` combinations map to more than one `.tid` in the logbook — an
upstream completeness issue in `scripts/01_fs_afladagbok_convert.R`. Affects ~7,200
records. These receive the same `.lid` here; the root cause should be fixed in the
convert script.
:::

::: {.callout-note}
### Permanently unmatched trips (~2%)

Trips remaining after all three cascade steps have no landing record within ±1 day at
the same port and no unambiguous vessel+date match. These are genuine gaps — either
unreported landings, transfers to other vessels (pelagic), or logbook entries for catch
that was discarded or used as bait.
:::

