---
title: "Linking Landings to Logbook Trips"
date: today
---

## Goal

Attach a landing identifier (`.tid_land`) to each logbook trip, enabling
downstream reconciliation between the two systems. The match is
produced by the second half of `data/logbooks/logbooks.R` and stored as
`data/logbooks/lid_map.parquet` with columns `.tid`, `schema`, `.tid_land`,
`match_type`. The result is also written back into `data/logbooks/trip.parquet`.

## Data {#sec-data}

### Landing register

`data/landings/landings.parquet` and `data/landings/catch.parquet` are built by
`data/landings/landings.R` from the Oracle dump
`data-dump/landings/agf/aflagrunnur.parquet`. The source has one row per
species per delivery; `landings.R` groups to trip level:

- **`landings.parquet`** — one row per landing trip: `.tid` (= `min(.lid)`
  within `vid × date × hid`), `vid`, `date`, `hid`, `höfn`
- **`catch.parquet`** — one row per species per original landing record: `.tid`,
  `.lid` (original `londun_id`), `gid`, `.sid` (species), `.stid` (stock),
  `catch` (kg)

Where multiple `.lid` records share the same `vid × date × hid` (split
deliveries or gear-disaggregated entries), they are collapsed to a single `.tid`
with catch summed. In the matching script, `.tid` is renamed to `.tid_land` to
avoid collision with the logbook `.tid`.

### Logbook trips

`data/logbooks/trip.parquet` is the merged `afli` + `fs_afladagbok` output (see
[Merge](merge.qmd)). Trips with no catch and foreign vessels (vid 3700–4999)
are excluded before matching. The arrival port `hid` and its name `port` are
used as the port key.

## Matching strategy {#sec-matching}

The cascade works in eight steps of decreasing key specificity. Each step
operates only on trips and landings **not yet consumed** by earlier steps, and
accepts only **unambiguous** matches — exactly one `.tid_land` candidate per
`.tid`. The helper used throughout:

```{r}
#| eval: false
try_match <- function(unmatched, land_tbl, by, type) {
  unmatched |>
    inner_join(land_tbl, by = by) |>
    group_by(.tid) |>
    filter(n_distinct(.tid_land) == 1) |>
    slice(1) |>
    ungroup() |>
    mutate(match_type = type)
}
```

### Steps 1–5: key-based matching

The natural join key is `vid × date × hid`. Steps 1–5 progressively relax this
key, shifting the landing date window and dropping the port requirement.

```{r}
#| eval: false
# Step 1: exact key
m1 <- try_match(trip, land, c("vid", "date", "hid"), "01 exact")

# Step 2: date ±1 day, port still required
land_pm1 <- bind_rows(land |> mutate(date = date - 1),
                      land |> mutate(date = date + 1))
m2 <- try_match(..., land_pm1, c("vid", "date", "hid"), "02 exact ±1 day")

# Step 3: same day, port dropped
m3 <- try_match(..., land, c("vid", "date"), "03 vid_date")

# Step 4: ±1 day, port dropped
m4 <- try_match(..., land_pm1, c("vid", "date"), "04 vid_date ±1")

# Step 5: ±2 days, port dropped
land_pm2 <- bind_rows(land |> mutate(date = date - 2),
                      land |> mutate(date = date + 2))
m5 <- try_match(..., land_pm2, c("vid", "date"), "05 vid_date ±2")
```

**Why ±2 days?** The table below shows how signal-to-ambiguity degrades as the
window widens:

| Extension | New unambiguous matches | New ambiguous (discarded) | Ratio |
|---|---|---|---|
| ±1 day | ~21,600 | — | — |
| ±2 days | ~3,400 | ~300 | 11:1 |
| ±3 days | ~600 | ~300 | ~2:1 |

All 3,400 trips resolved at ±2 use exactly a 2-day offset, pointing to a
systematic recording lag (landing registered the day after `T2`) rather than
anything correlated with trip length. ±3 is not worth the added ambiguity.

### Steps 6–7: catch-weight matching {#sec-weight-match}

Trips still unmatched after step 5 may be ambiguous — matching multiple
landings on `vid × date`. Steps 6 and 7 revisit those trips and use the
logbook total catch weight as a tiebreaker. A match is accepted when:

1. `abs(log(catch_logbook / catch_landing)) ≤ tol` — the weight ratio is within
   tolerance.
2. The log-ratio gap between best and second-best candidate exceeds
   `WEIGHT_MIN_GAP = log(1.5)` — one candidate is clearly closer than the rest.

Demersal and pelagic trips are treated separately:

- **Step 6 (demersal):** tolerance `log(1.3)` (~30%). Demersal trips land their
  full catch, so a close weight match is a reliable signal.
- **Step 7 (pelagic):** tolerance `log(2.0)` (~100%). Pelagic trips routinely
  transfer catch at sea, so the logbook total rarely matches any single landing
  record. This step yields near-zero matches but is retained so tolerances can
  be tuned without restructuring the cascade.

```{r}
#| eval: false
try_match_weight <- function(unmatched, land_tbl, consumed_land, tol, gap, type) {
  unmatched |>
    inner_join(land_tbl |> filter(!.tid_land %in% consumed_land),
               by = c("vid", "date")) |>
    mutate(score = abs(log(catch / catch_land))) |>
    filter(is.finite(score), score <= tol) |>
    group_by(.tid) |>
    arrange(score, .by_group = TRUE) |>
    filter(n() == 1 | (score[2] - score[1]) >= gap) |>
    slice(1) |>
    ungroup() |>
    mutate(match_type = type)
}
```

The pelagic flag is derived as `gid %in% pelagic_gids` rather than via a join
on `gear_mapping`, which has multiple rows per `gid` across old/new code
versions.

### Step 8: wrong-date detection {#sec-wrong-date}

`.tid` values follow rough chronological order within each of three ranges:
positive `afli`, negative `afli`, and `fs_afladagbok`. An `approxfun` is fitted
per range using matched trips as the reference curve (`tid_num → expected_date`).
Unmatched trips where `|recorded_date − expected_date| > 35 days` are flagged as
likely data-entry errors (month or year transposition); ~2,068 such trips are
searched for a landing within ±3 days of the expected date, accepting only
unambiguous hits.

Yield is small (~36 matches), but confidence is high: two independent signals
agree — a large `.tid`/date mismatch and a unique landing on the expected date.
The `date_error_days` column on unmatched trips in `lid_map.parquet` is the
primary value of this step: a QA diagnostic for identifying suspect entries.

## Results {#sec-results}

```{r}
#| label: setup
library(nanoparquet)
library(tidyverse)
trip    <- read_parquet("data/logbooks/trip.parquet")
lid_map <- read_parquet("data/logbooks/lid_map.parquet")
```

### Match counts by step

```{r}
#| label: tbl-cascade
#| tbl-cap: "Matches per step and cumulative coverage."
lid_map |>
  count(match_type) |>
  mutate(cumulative = cumsum(n),
         pct        = round(100 * n / nrow(trip |> filter(catch > 0)), 1))
```

### Annual coverage

```{r}
#| label: fig-annual-trips
#| fig-cap: "Annual logbook trip counts by match outcome, 2007–2025."
trip |>
  filter(catch > 0, date >= ymd("2007-09-01"), date <= ymd("2025-12-12")) |>
  mutate(match_type = replace_na(match_type, "09 unmatched"),
         year = year(date)) |>
  count(year, match_type) |>
  ggplot(aes(year, n / 1e3, fill = match_type)) +
  geom_col(position = position_stack(reverse = TRUE)) +
  scale_fill_brewer(palette = "Set1") +
  labs(x = NULL, y = "Thousands of trips", fill = NULL)
```

### Trip-level weight agreement

For matched trips, logbook `catch` and landing catch are both in kg.

```{r}
#| label: tbl-weight-ratio
#| tbl-cap: "Ratio of logbook catch to landing weight for matched trips."
land <-
  read_parquet("data/landings/landings.parquet") |>
  filter(vid > 0, date >= ymd("2007-09-01"), date <= ymd("2025-12-12")) |>
  rename(.tid_land = .tid) |>
  left_join(
    read_parquet("data/landings/catch.parquet") |>
      group_by(.tid) |>
      summarise(land_catch = sum(catch), .groups = "drop") |>
      rename(.tid_land = .tid),
    by = ".tid_land"
  )

trip |>
  filter(!is.na(.tid_land)) |>
  inner_join(land |> select(.tid_land, land_catch), by = ".tid_land") |>
  filter(catch > 0, land_catch > 0) |>
  mutate(ratio = catch / land_catch) |>
  group_by(schema, match_type) |>
  summarise(
    n      = n(),
    p25    = round(quantile(ratio, 0.25), 3),
    median = round(median(ratio), 3),
    p75    = round(quantile(ratio, 0.75), 3),
    mean   = round(mean(ratio), 3),
    .groups = "drop"
  )
```

The median ratio close to 1.0 confirms that both systems record catch in kg and
that agreement for matched trips is genuine. Elevated means relative to the
median reflect pelagic trips where catch is transferred to a receiving vessel.

## Coverage: the view from landings {#sec-reverse}

Reversing the question — which landings lack a logbook trip? — is more
informative for identifying systematic gaps.

```{r}
#| label: fig-coverage-year
#| fig-cap: "Fraction of landings matched to a logbook trip by year, 2007–2025. Dashed lines: curator retirement (~2019–2020) and market opening (spring 2021)."
land |>
  filter(vid > 0, date >= ymd("2007-09-01"), date <= ymd("2025-12-12")) |>
  mutate(year    = year(date),
         matched = .tid_land %in% lid_map$.tid_land) |>
  count(year, matched) |>
  group_by(year) |>
  mutate(pct = 100 * n / sum(n)) |>
  filter(matched) |>
  ggplot(aes(year, pct)) +
  geom_col() +
  geom_vline(xintercept = c(2019.5, 2021), linetype = "dashed", colour = "grey50") +
  labs(x = NULL, y = "Landings matched (%)")
```

Coverage improved from ~73% in 2009–2010 to ~87% by 2017–2018, then fell to
~72% in 2020. The arc is better explained by institutional factors (the
curator's retirement and the market opening) than by changes in fishing
behaviour.

### By gear type

```{r}
#| label: tbl-gear-coverage
#| tbl-cap: "Logbook coverage by gear type on the landings side (≥100 landings)."
land_catch_gid <- read_parquet("data/landings/catch.parquet") |>
  select(.tid, gid) |>
  distinct() |>
  rename(.tid_land = .tid)

land |>
  filter(vid > 0, date >= ymd("2007-09-01"), date <= ymd("2025-12-12")) |>
  mutate(matched = .tid_land %in% lid_map$.tid_land) |>
  left_join(land_catch_gid, by = ".tid_land") |>
  group_by(gid) |>
  summarise(n = n(), pct_matched = round(100 * mean(matched), 1), .groups = "drop") |>
  filter(n >= 100) |>
  arrange(-pct_matched) |>
  knitr::kable()
```

Two gear types are nearly absent from logbooks:

- **Sjóstöng** (`gid=19`, sea-rod): <1% coverage across ~20k landings — coastal
  pole-and-line, likely exempt from logbook requirements.
- **Grásleppunet / Rauðmaganet** (`gid=3/4`, lumpfish gillnets): 26% and 50%
  match rates across ~60k combined landings. The largest single contributor to
  unmatched landings in absolute terms.

::: {.callout-warning}
### Lumpfish logbook records are unreliable

`gid=3` and `gid=4` together represent ~60k landings 2008–2020 but match at
only 26% and 50% respectively. Logbook records that do exist for lumpfish are
known to be problematic (seasonal timing, gear reporting). **Do not use lumpfish
logbook data in catch reconciliation without dedicated QA.**
:::

## Outstanding issues {#sec-outstanding}

::: {.callout-note}
### Collapsed landing records

Where multiple `.lid` values share the same `vid × date × hid`, they are
collapsed to a single `.tid` (= `min(.lid)`) with catch summed. This occurs for
split deliveries or gear-disaggregated entries in `aflagrunnur`. The collapsed
record should be validated before downstream weight reconciliation.
:::

::: {.callout-note}
### Wrong-date QA

~2,068 unmatched trips carry dates >35 days from their `.tid`-interpolated
expected date. The `date_error_days` column in `lid_map.parquet` quantifies the
deviation and can be used to prioritise manual review.
:::

::: {.callout-note}
### Permanently unmatched trips

Trips remaining after all eight steps have no landing record within the search
window. Causes include: unreported landings, catch transferred to other vessels
(pelagic), catch discarded or used as bait, and logbook entries for fisheries
with no landing requirement.
:::
