---
title: "Merging logbook data sources"
subtitle: "Rationale, method, and quality assessment"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
execute:
  warning: false
---

```{r}
#| label: load-packages
library(tidyverse)
library(nanoparquet)
library(gt)
```

```{r}
#| label: load-data
afli_trip    <- read_parquet("data/afli/trip.parquet")
afli_station <- read_parquet("data/afli/station.parquet")
afli_fs      <- read_parquet("data/afli/fishing_sample.parquet")
afli_catch   <- read_parquet("data/afli/catch.parquet")

fs_trip    <- read_parquet("data/fs_afladagbok/trip.parquet")
fs_station <- read_parquet("data/fs_afladagbok/station.parquet")
fs_fs      <- read_parquet("data/fs_afladagbok/fishing_sample.parquet")

merged_trip    <- read_parquet("data/merged/trip.parquet")
merged_station <- read_parquet("data/merged/station.parquet")
merged_fs      <- read_parquet("data/merged/fishing_sample.parquet")
merged_catch   <- read_parquet("data/merged/catch.parquet")
```

# Overview

The `data/merged/` parquet files are the primary output of this project: a
single unified set of `trip`, `station`, `fishing_sample`, and `catch` tables
spanning the full historical record of Icelandic fishing logbooks. They are
produced by `scripts/02_merge.R`.

Two source schemas are combined:

| Schema | Script | Temporal coverage | Character |
|---|---|---|---|
| **afli** | `01_afli_convert.R` | ~1950 – present | Legacy Oracle logbook system; primary historical record |
| **fs_afladagbok** | `01_fs_afladagbok_convert.R` | 2020 – 2025-12 | Static dump from Fiskistofa; canonical digital logbook format |

A third schema (`adb`, the live electronic feed) exists in `data/adb/` but is
excluded from the merge; its gear codes are old-schema values that require
lossy conversion, and its coordinates are less thoroughly validated.

# Source priority

## Principle

Where both schemas cover the same vessel-day, a single source is chosen for
all stations on that day. The decision granularity is the **vessel-day**
(`vid × date`) to avoid partial-day splits where one source has some hauls
and the other has others.

**Priority order:**

1. **afli** — used unconditionally where present.
2. **fs_afladagbok** — used for vessel-days not present in `afli`.

## Why afli takes priority

The key reason is **mobile gear timing precision**. In `afli`, tow-start time
(`t2`) is derived from the on-bottom clock (`ibotni`), an HHMM integer entered
by the skipper to record when the gear reached the seabed. Tow-end (`t3`) is
then `t2 + towtime`. In `fs_afladagbok`, `t2` and `t3` are not recorded for
OTB or OTM; duration is instead computed as `t4 − t1` (end-of-operation minus
start-of-operation), which includes deployment and hauling time in addition to
actual tow time.

```{r}
#| label: tbl-timing-comparison
#| tbl-cap: "Timing column availability (%) by schema and gear class"

focus_gears <- c("OTB", "OTM", "DRB", "SDN", "LLS", "GNS", "LHM", "FPO")

merged_fs |>
  filter(gear %in% focus_gears) |>
  group_by(Schema = schema, `Gear` = gear) |>
  summarise(
    Stations     = n(),
    `t1 (%)` = round(mean(!is.na(t1)) * 100, 1),
    `t2 (%)` = round(mean(!is.na(t2)) * 100, 1),
    `t3 (%)` = round(mean(!is.na(t3)) * 100, 1),
    `t4 (%)` = round(mean(!is.na(t4)) * 100, 1),
    `duration_m (%)` = round(mean(!is.na(duration_m)) * 100, 1),
    .groups = "drop"
  ) |>
  arrange(Gear, Schema) |>
  gt() |>
  fmt_number(columns = Stations, use_seps = TRUE, decimals = 0) |>
  tab_style(
    style = cell_fill(color = "#fff3cd"),
    locations = cells_body(
      columns = c(`t2 (%)`),
      rows = Schema == "afli" & Gear %in% c("OTB", "OTM") & `t2 (%)` > 50
    )
  )
```

For bottom trawl (OTB), `afli` achieves ~86% `t2` coverage (on-bottom clock)
versus 0% in `fs_afladagbok`. The `duration_m` in `afli` is the recorded tow
time on the seabed; in `fs_afladagbok` it is the full operation window from
warp-entry to hauling completion, which is systematically longer.

A secondary reason is historical span: `afli` covers ~1950 to the present,
and the merged output inherits that full historical depth from it directly.

# Coverage

## Vessel-day overlap

```{r}
#| label: tbl-coverage
#| tbl-cap: "Vessel-day coverage in the overlap period (2020–2025)"

DATE_MIN <- as.Date("2020-01-01")
FS_MAX   <- as.Date("2025-12-31")

afli_vd <- afli_station |>
  left_join(afli_trip |> select(.tid, vid), by = ".tid") |>
  filter(between(date, DATE_MIN, FS_MAX)) |>
  distinct(vid, date)

fs_vd <- fs_station |>
  filter(between(date, DATE_MIN, FS_MAX)) |>
  left_join(fs_trip |> select(.tid, vid), by = ".tid") |>
  distinct(vid, date)

tibble(
  Metric = c(
    "afli vessel-days (2020–2025)",
    "fs_afladagbok vessel-days (2020–2025)",
    "Vessel-days in both schemas",
    "fs_afladagbok vessel-days not in afli (tier-2 contribution)"
  ),
  Count = c(
    nrow(afli_vd),
    nrow(fs_vd),
    nrow(semi_join(fs_vd, afli_vd, by = c("vid", "date"))),
    nrow(anti_join(fs_vd, afli_vd, by = c("vid", "date")))
  )
) |>
  mutate(Count = scales::comma(Count)) |>
  gt()
```

# Implementation

## Date bounds

| Source | Lower bound | Upper bound | Rationale |
|---|---|---|---|
| `afli` | none | 2026-12-31 | Reliable across the full historical period |
| `fs_afladagbok` | 2020-01-01 | 2025-12-31 | Static dump; coverage ends 2025-12 |

## Decision rule

```r
# Tier 1: afli vessel-day index
afli_vid_date <- afli_station |>
  left_join(afli_trip |> select(.tid, vid), by = ".tid") |>
  distinct(vid, date)

# Tier 2: fs_afladagbok vessel-days not covered by afli
fs_only_station <- fs_station |>
  left_join(fs_trip |> select(.tid, vid), by = ".tid") |>
  anti_join(afli_vid_date, by = c("vid", "date")) |>
  select(-vid)
```

## Record counts

```{r}
#| label: tbl-record-counts
#| tbl-cap: "Record counts in the merged output"

# catch.parquet has no schema column; look up via .sid membership
fs_sids <- merged_station |> filter(schema == "fs_afladagbok") |> pull(.sid)

tibble(
  Table = c("station", "fishing_sample", "trip", "catch"),
  afli = c(
    nrow(merged_station |> filter(schema == "afli")),
    nrow(merged_fs      |> filter(schema == "afli")),
    nrow(merged_trip    |> filter(schema == "afli")),
    nrow(merged_catch) - sum(merged_catch$.sid %in% fs_sids)
  ),
  `fs-only` = c(
    nrow(merged_station |> filter(schema == "fs_afladagbok")),
    nrow(merged_fs      |> filter(schema == "fs_afladagbok")),
    nrow(merged_trip    |> filter(schema == "fs_afladagbok")),
    sum(merged_catch$.sid %in% fs_sids)
  ),
  Merged = c(
    nrow(merged_station),
    nrow(merged_fs),
    nrow(merged_trip),
    nrow(merged_catch)
  )
) |>
  mutate(across(where(is.numeric), scales::comma)) |>
  gt()
```

## Year-by-year source breakdown

```{r}
#| label: fig-yearly-source
#| fig-cap: "Station counts by source, 2018–2025"
#| fig-height: 4

merged_station |>
  mutate(yr = lubridate::year(date)) |>
  filter(between(yr, 2018, 2025)) |>
  count(yr, schema) |>
  ggplot(aes(yr, n, fill = schema)) +
  geom_col() +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(breaks = 2018:2025) +
  labs(x = NULL, y = "Stations", fill = "Schema")
```

`afli` dominates through 2021 and drops sharply from 2022 as the legacy
system was wound down. `fs_afladagbok` fills the gap from 2021 onwards,
with near-complete coverage from 2023.

## `.tid` type harmonisation

`afli` trip identifiers (`.tid`) are numeric Oracle sequence values.
`fs_afladagbok` `.tid` values are also numeric but from a different sequence.
Both are coerced to character in the merged output so the column type is
uniform. Downstream joins should key on `(schema, .sid)` or `(schema, .tid)`
to avoid cross-schema collisions.

# Timing quality

Timing columns are in `fishing_sample`; `station` holds only the spatial and
date envelope. The two schemas use different time columns depending on gear
class:

| Schema | Mobile gear | Static gear |
|---|---|---|
| `afli` | `t2` tow-start (on-bottom clock); `t3` = `t2` + tow minutes | `t3` retrieval start; `t4` retrieval end |
| `fs_afladagbok` | `t1` deployment start; `t4` operation end (`t2`, `t3` absent for OTB/OTM) | `t1` deployment start; `t3` retrieval start; `t4` retrieval end |

## t1 ≤ t4 ordering violations

```{r}
#| label: tbl-timing-violations
#| tbl-cap: "t1 > t4 ordering violations by schema (fishing_sample)"

merged_fs |>
  filter(!is.na(t1), !is.na(t4)) |>
  group_by(Schema = schema) |>
  summarise(
    `Rows with t1 and t4` = n(),
    `t1 > t4 violations`  = sum(t1 > t4),
    `Violation rate (%)`  = round(mean(t1 > t4) * 100, 3),
    .groups = "drop"
  ) |>
  gt() |>
  fmt_number(columns = 2:3, use_seps = TRUE, decimals = 0)
```

Violations in `afli` are confined to records with negative `.sid` values —
known garbage entries inherited from the Oracle source. Filter to
`.sid > 0` to exclude them before timing-sensitive analyses.

## Duration coverage by gear

```{r}
#| label: tbl-duration-coverage
#| tbl-cap: "duration_m availability (%) and median value (hours) by schema and gear"

merged_fs |>
  filter(gear %in% c("OTB", "OTM", "DRB", "SDN", "LLS", "GNS", "GND", "LHM", "FPO")) |>
  group_by(Schema = schema, Gear = gear) |>
  summarise(
    Stations       = n(),
    `duration available (%)` = round(mean(!is.na(duration_m)) * 100, 1),
    `median duration (h)`    = round(median(duration_m, na.rm = TRUE) / 60, 1),
    .groups = "drop"
  ) |>
  arrange(Gear, Schema) |>
  gt() |>
  fmt_number(columns = Stations, use_seps = TRUE, decimals = 0)
```

`afli` SDN (Dragnót) shows low duration availability (~11%) because that gear
class records a tow count rather than timestamps; effort for SDN is per
setting. The `afli` FPO (traps) rate of ~46% reflects sparse soak-time
recording in the legacy system.

# Annual catch totals

```{r}
#| label: fig-annual-catch
#| fig-cap: "Annual catch totals in the merged dataset by source"
#| fig-height: 4

merged_catch |>
  left_join(merged_station |> select(.sid, date, schema), by = join_by(.sid, schema)) |>
  mutate(yr = lubridate::year(date)) |>
  filter(between(yr, 1990, 2025)) |>
  group_by(yr, schema) |>
  summarise(catch_kt = sum(catch, na.rm = TRUE) / 1e6, .groups = "drop") |>
  ggplot(aes(yr, catch_kt, fill = schema)) +
  geom_col() +
  labs(x = NULL, y = "Catch (kt)", fill = "Schema")
```

```{r}
#| label: tbl-annual-catch-recent
#| tbl-cap: "Annual catch totals, 2018–2025 (thousands of tonnes)"

merged_catch |>
  left_join(merged_station |> select(.sid, date, schema),by = join_by(.sid, schema)) |>
  mutate(yr = lubridate::year(date)) |>
  filter(between(yr, 2018, 2025)) |>
  group_by(yr, schema) |>
  summarise(catch_kt = round(sum(catch, na.rm = TRUE) / 1e6, 0), .groups = "drop") |>
  pivot_wider(names_from = schema, values_from = catch_kt, values_fill = 0L) |>
  rename(Year = yr) |>
  mutate(Total = afli + fs_afladagbok) |>
  gt() |>
  fmt_number(columns = where(is.numeric), use_seps = TRUE, decimals = 0)
```

# Known limitations

1. **`.sid` and `.tid` collisions**: `afli` and `fs_afladagbok` can share the
   same numeric identifiers by coincidence. The `schema` column disambiguates
   provenance; downstream joins should always key on `(schema, .sid)` rather
   than `.sid` alone.

2. **Residual t1 ≤ t4 violations**: see @tbl-timing-violations. Small in
   absolute terms and confined to known garbage records in `afli`.
   Filter `.sid > 0` before timing-sensitive analyses.

3. **Mobile gear timing in fs_afladagbok**: `duration_m` for OTB/OTM is
   `t4 − t1` (full operation time), not pure tow time on the seabed.
   This is systematically longer than the tow-time recorded in `afli`.
   In the merged output, OTB/OTM vessel-days covered by `afli` use the more
   precise measure; `fs_afladagbok`-only records use the coarser estimate.

4. **DRB duration in fs_afladagbok suspiciously short**: median ~18 minutes
   for dredge operations, which is atypically brief. How `upphaf_timi` and
   `lok_timi` are populated for plow gear in the Fiskistofa system is not
   fully understood.

5. **~54 k OTB stations (2021–2022) missing effort data**: these `fs_afladagbok`
   records have no matching row in `ws_dragnot_varpa`, so `effort_unit = NA`
   and `effort = NA`. This is a completeness gap in the source data, not a
   processing error.

6. **Wacky coordinate error in afli logger positions**: automatic logger
   positions (`rafr_sjalfvirkir_maelar.parquet`) are affected by a systematic
   DDMMmm→DDMMss misinterpretation. A recovery algorithm is documented in
   `wacky_recovery.qmd` but has not yet been applied to the merged output.
