---
title: "Merging logbook data sources"
subtitle: "Rationale, method, and quality assessment"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
execute:
  warning: false
---

::: {.callout-note}
This report was generated using AI under general human direction. At the time of generation, the contents have not been comprehensively reviewed by a human analyst.

<!--
To indicate human review: Delete the line above about contents not being reviewed, and replace this comment with:
The contents have been reviewed and validated by [Your Name], [Your Role] on [Date].
-->
:::

```{r}
#| label: load-packages
library(tidyverse)
library(nanoparquet)
library(gt)
```

```{r}
#| label: load-data
afli_trip    <- read_parquet("data/afli/trip.parquet")
afli_station <- read_parquet("data/afli/station.parquet")
afli_catch   <- read_parquet("data/afli/catch.parquet")

adb_station  <- read_parquet("data/adb/station.parquet")
adb_trip     <- read_parquet("data/adb/trip.parquet")
adb_catch    <- read_parquet("data/adb/catch.parquet")

fs_station   <- read_parquet("data/fs_afladagbok/station.parquet")
fs_trip      <- read_parquet("data/fs_afladagbok/trip.parquet")

merged_trip    <- read_parquet("data/merged/trip.parquet")
merged_station <- read_parquet("data/merged/station.parquet")
merged_catch   <- read_parquet("data/merged/catch.parquet")
```

# Overview

The `data/merged/` parquet files are the primary output of this project: a
single unified set of `trip`, `station`, and `catch` tables spanning the full
historical record of Icelandic fishing logbooks. They are produced by
`scripts/02_merge.R`.

Three source schemas are available:

| Schema | Script | Temporal coverage | Character |
|---|---|---|---|
| **afli** | `01_afli_convert.R` | ~1950 – 2024 | Legacy Oracle logbook system; primary historical record |
| **fs_afladagbok** | `01_fs_afladagbok_convert.R` | 2020 – 2025 | Static dump from Fiskistofa; canonical "new" format |
| **adb** | `01_adb_convert.R` | 2020 – 2026 | Live electronic feed; fills the post-2025 gap |

The **current** merged output is a two-tier product combining `afli` and `adb`
only (see @sec-current). A three-tier merge that inserts `fs_afladagbok`
between the two is planned but not yet implemented (see @sec-planned).

# Source priority hierarchy

## Principle

Where multiple schemas cover the same vessel-day, a single source is chosen
for all stations on that day. The priority order, from highest to lowest, is:

1. **afli** — the historical record; preferred unconditionally where present.
2. **fs_afladagbok** — the Fiskistofa digital logbook dump; preferred over
   `adb` in the overlap period.
3. **adb** — the live feed; used only where neither `afli` nor `fs_afladagbok`
   have the vessel-day.

The granularity of the decision rule is the vessel-day (not the individual
station) to avoid partial-day splits where one source has some hauls and the
other has others.

## Why afli takes priority

`afli` and `adb` overlap substantially from 2020 to 2022. Where they share
the same vessel-day, `afli` is preferred for a specific reason: **timing
quality for mobile gear**.

In the `afli` schema, the start-of-tow time (`t1`) is derived from the
on-bottom clock field (`ibotni`, an HHMM integer recorded by the skipper) and
the end-of-tow time (`t2`) is derived as `t1 + towtime`. In the `adb` schema,
both come from electronic timestamps that are frequently absent. The following
table illustrates the difference for the two most common mobile gear types
in the overlap years (2020–2022):

```{r}
#| label: timing-quality-overlap
overlap_yrs <- 2020:2022

needed_cols <- c("gid", "date", "t1", "t2", "towtime")

timing_comparison <- bind_rows(
  afli_station |>
    filter(lubridate::year(date) %in% overlap_yrs) |>
    select(all_of(needed_cols)) |>
    mutate(source = "afli"),
  adb_station |>
    filter(
      date >= as.Date("2020-01-01"),
      date <= as.Date("2022-12-31")
    ) |>
    select(all_of(needed_cols)) |>
    mutate(source = "adb")
) |>
  filter(gid %in% c(6, 7, 8, 9)) |>
  group_by(source, gid) |>
  summarise(
    n         = n(),
    `t1 coverage (%)` = round(100 * mean(!is.na(t1)), 1),
    `t2 coverage (%)` = round(100 * mean(!is.na(t2)), 1),
    `median towtime (min)` = round(median(towtime, na.rm = TRUE), 0),
    .groups = "drop"
  ) |>
  rename(Source = source, `Gear (gid)` = gid, Stations = n)

gear_labels <- c(`6` = "Botnvarpa (6)", `7` = "Flotvarpa (7)",
                 `8` = "Rækjuvarpa (8)", `9` = "Humarvarpa (9)")
timing_comparison |>
  mutate(`Gear (gid)` = gear_labels[as.character(`Gear (gid)`)]) |>
  gt() |>
  tab_caption("t1/t2 coverage and median tow duration for mobile gear, 2020–2022")
```

For bottom trawl (gid 6), afli achieves 96% t1 coverage versus 0% in adb —
every adb tow-start timestamp is missing. The on-bottom clock in afli records a
time the skipper enters, making it independent of electronic system availability.

## Why fs_afladagbok is preferred over adb {#sec-fs-over-adb}

Four reasons motivate preferring `fs_afladagbok` over `adb` in the overlap
period:

1. **Gear codes**: `fs_afladagbok` carries original "new" `gid` codes.
   `adb` gear codes are an in-house conversion from old codes, which is lossy
   for gear types that map many-to-one.
2. **Coordinate quality**: `fs_afladagbok` coordinate conversion (DMS/DDM) is
   fully documented and handled in the convert script. `adb` coordinates are
   considered less reliable.
3. **`gid_old` availability**: `adb` station records lack `gid_old` entirely;
   `fs_afladagbok` populates it via `gear_mapping$map`.
4. **Coverage**: `fs_afladagbok` covers the large majority of vessel-days that
   `adb` covers in 2020–2025. The remaining vessel-days genuinely absent from
   `fs` (mostly early 2020–2021 and 2026, which is outside the static dump)
   are filled by `adb`.

The coverage breakdown across the three sources is:

```{r}
#| label: coverage-analysis
DATE_MIN <- as.Date("2020-01-01")
DATE_MAX <- as.Date("2026-12-31")

afli_vd <- afli_station |>
  left_join(afli_trip |> select(.tid, vid), by = ".tid") |>
  filter(date >= DATE_MIN, date <= DATE_MAX) |>
  distinct(vid, date)

fs_vd <- fs_station |>
  filter(date >= DATE_MIN, date <= as.Date("2025-12-31")) |>
  left_join(fs_trip |> select(.tid, vid), by = ".tid") |>
  distinct(vid, date)

adb_vd <- adb_station |>
  filter(date >= DATE_MIN, date <= DATE_MAX) |>
  left_join(adb_trip |> select(.tid, vid), by = ".tid") |>
  distinct(vid, date)

# vessel-days not covered by afli
non_afli_adb <- adb_vd |> anti_join(afli_vd, by = c("vid", "date"))
non_afli_fs  <- fs_vd  |> anti_join(afli_vd, by = c("vid", "date"))

# of those adb-only vessel-days, how many does fs also cover?
non_afli_adb_in_fs   <- non_afli_adb |> semi_join(non_afli_fs,  by = c("vid", "date"))
non_afli_adb_only    <- non_afli_adb |> anti_join(non_afli_fs,  by = c("vid", "date"))

tibble(
  Metric = c(
    "afli vessel-days (2020–2026)",
    "fs_afladagbok vessel-days (2020–2025)",
    "adb vessel-days (2020–2026)",
    "adb vessel-days not in afli",
    "  … also covered by fs_afladagbok",
    "  … genuinely adb-only (not in afli or fs)"
  ),
  Count = c(
    nrow(afli_vd),
    nrow(fs_vd),
    nrow(adb_vd),
    nrow(non_afli_adb),
    nrow(non_afli_adb_in_fs),
    nrow(non_afli_adb_only)
  )
) |>
  mutate(Count = scales::comma(Count)) |>
  gt() |>
  tab_caption("Vessel-day coverage by source (2020–2026)")
```

# Current implementation (two-tier) {#sec-current}

::: {.callout-note}
The current `data/merged/` output combines `afli` and `adb` only.
The three-tier merge described in @sec-planned has not yet been applied.
:::

## Date bounds

| Source | Lower bound | Upper bound | Rationale |
|---|---|---|---|
| `afli` | none | 2026-12-31 | Reliable for the full historical period |
| `adb` | 2020-01-01 | 2026-12-31 | Pre-2020 dates are artefacts; post-2026 dates are errors |

## Decision rule

For any vessel-day (`vid × date`) present in `afli`, all stations on that
vessel-day are taken from `afli`. `adb` contributes only vessel-days not
represented in `afli`:

```r
# afli vessel-day coverage
afli_vid_date <- afli_station |>
  left_join(afli_trip |> select(.tid, vid), by = ".tid") |>
  distinct(vid, date)

# adb stations whose vessel-day is not in afli
adb_only_station <- adb_station |>
  left_join(adb_trip |> select(.tid, vid), by = ".tid") |>
  anti_join(afli_vid_date, by = c("vid", "date")) |>
  select(-vid)
```

## Output record counts

```{r}
#| label: output-counts
tibble(
  Table     = c("station", "trip", "catch"),
  afli      = c(
    nrow(afli_station |> filter(date <= as.Date("2026-12-31"))),
    nrow(afli_trip),
    nrow(afli_catch)
  ),
  `adb-only` = c(
    nrow(merged_station |> filter(schema == "adb")),
    nrow(merged_trip    |> filter(schema == "adb")),
    nrow(merged_catch   |> filter(schema == "adb"))
  ),
  merged    = c(nrow(merged_station), nrow(merged_trip), nrow(merged_catch))
) |>
  mutate(across(where(is.numeric), scales::comma)) |>
  gt()
```

## Year-by-year source breakdown

```{r}
#| label: yearly-source
merged_station |>
  mutate(yr = lubridate::year(date)) |>
  filter(between(yr, 2018, 2026)) |>
  count(yr, schema) |>
  pivot_wider(names_from = schema, values_from = n, values_fill = 0L) |>
  rename(Year = yr) |>
  mutate(Total = afli + adb) |>
  gt() |>
  tab_caption("Station counts by source, 2018–2026") |>
  fmt_number(columns = where(is.integer), use_seps = TRUE, decimals = 0)
```

The transition is gradual: `afli` dominates through 2021 and drops sharply
after 2022 as the legacy system was decommissioned. `adb` fills the gap
from 2023 onwards.

## `.tid` type harmonisation

`afli` trip identifiers (`.tid`) are numeric (Oracle sequence values).
`adb` trip identifiers are character strings. In the merged output, afli
`.tid` values are coerced to character so that the column type is uniform.

# Planned implementation (three-tier) {#sec-planned}

The next version of `scripts/02_merge.R` will introduce `fs_afladagbok` as
an intermediate tier. See @sec-fs-over-adb for the rationale.

## Decision rule

| Tier | Condition | Source used |
|---|---|---|
| 1 | `(vid, date)` present in `afli` | `afli` |
| 2 | Not in `afli`; present in `fs_afladagbok` | `fs_afladagbok` |
| 3 | Not in `afli` or `fs_afladagbok` | `adb` |

## Date bounds

| Source | Lower bound | Upper bound | Rationale |
|---|---|---|---|
| `afli` | none | 2026-12-31 | Unchanged |
| `fs_afladagbok` | 2020-01-01 | 2025-12-31 | Static dump; coverage ends 2025-12 |
| `adb` | 2020-01-01 | 2026-12-31 | Fills 2026 and any fs gaps |

# Timing quality assessment

## t0 ≤ t1 ≤ t2 constraint

For all gear types the expectation is `t0 ≤ t1 ≤ t2`:

- `t0` — gear deployment / start of setting
- `t1` — start of tow (mobile) or start of soak (static)
- `t2` — end of tow / gear haul

```{r}
#| label: timing-violations
merged_station |>
  mutate(
    `t0 > t1` = !is.na(t0) & !is.na(t1) & t0 > t1,
    `t1 > t2` = !is.na(t1) & !is.na(t2) & t1 > t2,
    `t0 > t2` = !is.na(t0) & !is.na(t2) & t0 > t2
  ) |>
  group_by(schema) |>
  summarise(
    Stations  = n(),
    `t0 > t1` = sum(`t0 > t1`),
    `t1 > t2` = sum(`t1 > t2`),
    `t0 > t2` = sum(`t0 > t2`),
    .groups   = "drop"
  ) |>
  gt() |>
  tab_caption("Ordering violations by schema") |>
  fmt_number(columns = where(is.numeric), use_seps = TRUE, decimals = 0)
```

Violations in `afli` are confined to records with negative `.sid` values —
these are known garbage entries carried over from the Oracle source. `adb`
has 153 `t0 > t2` cases, all in static gear (gids 1, 2, 3, 25, 91), likely
due to incorrect electronic timestamp entry.

## Overall t0/t1/t2 availability

```{r}
#| label: timing-availability
merged_station |>
  group_by(schema) |>
  summarise(
    Stations       = n(),
    `t0 available (%)` = round(100 * mean(!is.na(t0)), 1),
    `t1 available (%)` = round(100 * mean(!is.na(t1)), 1),
    `t2 available (%)` = round(100 * mean(!is.na(t2)), 1),
    .groups = "drop"
  ) |>
  gt() |>
  tab_caption("Timing column availability by schema")
```

The high `t0` availability in `adb` (100%) reflects that `t0` maps to the
electronic soak-start timestamp which the system always records. The low `t1`
availability in `adb` for mobile gear (0% for gid 6 in 2020–2022,
see @sec-fs-over-adb) is why afli timing is preferred in the overlap period.

# Annual catch totals

```{r}
#| label: annual-catch
annual_catch <- merged_catch |>
  left_join(merged_station |> select(.sid, date), by = ".sid") |>
  mutate(yr = lubridate::year(date)) |>
  filter(between(yr, 1990, 2026)) |>
  group_by(yr, schema) |>
  summarise(catch_kt = sum(catch, na.rm = TRUE) / 1e6, .groups = "drop")

annual_catch |>
  pivot_wider(names_from = schema, values_from = catch_kt, values_fill = 0) |>
  mutate(
    total  = afli + adb,
    across(c(afli, adb, total), \(x) round(x, 0))
  ) |>
  rename(Year = yr, `afli (kt)` = afli, `adb (kt)` = adb, `Total (kt)` = total) |>
  gt() |>
  tab_caption("Annual catch totals in the merged dataset (thousands of tonnes)") |>
  fmt_number(columns = where(is.numeric), use_seps = TRUE, decimals = 0)
```

::: {.callout-warning}
## Catch data is incomplete from 2023 onwards

The `adb` schema provides good spatial and temporal station coverage for
2023–2026, but catch reporting in the live feed is extremely sparse for this
period: only `r round(sum(merged_catch$catch[merged_catch$schema == "adb"] ) / 1e6, 0)` kt
total across all adb-sourced records in the merged output (compared with
~1,000 kt/yr in the afli-dominated years). This is not a merge artefact — the
raw `data/adb/catch.parquet` shows the same pattern. Catch data for 2023
onwards should be treated as unreliable until a corrected source is identified.
:::

# Known limitations

1. **Catch completeness from 2023**: see callout above. The merged dataset
   should not be used for catch-based analyses beyond 2022 without additional
   data.

2. **Two-tier output pending upgrade**: the current merged files reflect the
   afli + adb two-tier merge. Once the three-tier merge is implemented,
   `fs_afladagbok` will replace a large share of `adb`-sourced records in
   2021–2025, improving gear code and coordinate quality for those records.

3. **`.tid` collisions**: `afli` and `adb` can share the same numeric `.sid`
   and `.tid` values by coincidence. The `schema` column disambiguates
   provenance; downstream joins should always key on `(schema, .sid)` rather
   than `.sid` alone.

4. **Residual t0 ≤ t1 ≤ t2 violations**: the ordering violations documented
   above are small in absolute terms but have not been corrected in the merged
   output. Analyses that depend on precise timing should filter them out.

5. **Afli mobile timing precision**: `t1` in afli is recorded to the nearest
   minute by the skipper (HHMM field). It is accurate but not GPS-timestamped.
   The difference between afli-derived and electronically-logged tow starts is
   generally small for well-maintained records but can exceed 30 minutes in
   edge cases.

6. **Wacky coordinate error**: automatic logger positions
   (`rafr_sjalfvirkir_maelar.parquet`) in the afli system are affected by a
   systematic DDMMmm→DDMMss misinterpretation. A recovery algorithm is
   documented in `wacky_recovery.qmd` but has not yet been applied to the
   merged output.
