---
title: "macOS Disk Space — Diagnostics and Recovery"
date: 2026-04-27
---

Notes from a session where the ASTD ZIP → Parquet pipeline failed mid-run due to
a full disk. Covers how to find what is consuming space, why `du` and `df` often
disagree on macOS, and quick wins for recovering space.

---

## Checking free space

`df -h` reports usage at the filesystem/volume level.
The `-h` flag gives human-readable sizes (GB/TB).

```{bash}
#| eval: false
df -h /Users/einarhj   # or just df -h to see all volumes
```

Key columns: **Size**, **Used**, **Avail**, **Capacity**.
On a full disk you will see `Capacity: 100%` with only a few hundred MB **Avail**.

::: {.callout-warning}
Even a few hundred MB free is functionally full on macOS.
APFS reserves space for metadata and large writes will fail before
`Avail` reaches zero.
:::

---

## Finding large directories

### Top-level scan

`du -sh` gives a summary size for each argument; `sort -rh` sorts by
human-readable size descending.

```{bash}
#| eval: false
# One level deep under a directory
du -sh ~/stasi/*/  2>/dev/null | sort -rh | head -20
```

### Recursive drill-down

`--max-depth=N` limits recursion.  Start shallow and drill into the
largest directories.

```{bash}
#| eval: false
# Home directory, 6 levels deep — catches most project trees
du -h ~ --max-depth=6 2>/dev/null | sort -rh | head -30
```

### Stay on one filesystem

The `-x` flag (single-device) stops `du` from crossing mount points
(e.g., into `/Volumes/...`), avoiding inflated numbers.

```{bash}
#| eval: false
du -shx /Users/ /Applications/ /Library/ /opt/ /private/ 2>/dev/null \
  | sort -rh
```

---

## Why `du` and `df` disagree on macOS

On this machine the Data volume showed **969.9 GB** used (via `diskutil`),
but all `du` scans together only accounted for **~283 GB**.
The ~687 GB gap is explained by:

| Source | Explanation |
|---|---|
| **iCloud Drive local cache** | Files synced to iCloud but also cached locally; count toward APFS usage but many paths are SIP-protected (`Operation not permitted`). |
| **Photos library** | Similarly protected; can be hundreds of GB if Photos is set to keep originals. |
| **Sandboxed app containers** | `~/Library/Containers/`, `~/Library/Group Containers/` — macOS SIP prevents `du` from reading most of these. |
| **Purgeable space** | APFS marks some cached/iCloud-backed content as "purgeable"; it shows as used in `df` but can be reclaimed by the OS on demand. |
| **Trash** | Files deleted in Finder but not emptied; `~/.Trash` is also protected from terminal `du`. |

::: {.callout-note}
The only tool that sees the true breakdown — including purgeable content —
is **System Settings → General → Storage**.
Use it to identify and offload iCloud content, empty Trash, and review
large app data before going further.
:::

### Inspecting the APFS container directly

```{bash}
#| eval: false
# Shows actual bytes consumed per APFS volume (more honest than df)
diskutil apfs list disk3
```

Look for the **Capacity Consumed** line on each volume.

---

## NTFS volumes are read-only on macOS

External drives formatted as **NTFS** are mounted read-only by default on macOS.
No third-party driver = no writes.

```{bash}
#| eval: false
# Check the filesystem type and protocol of a mounted volume
diskutil info /Volumes/myDrive | grep -E "Protocol|Type|Solid State"
```

**Options:**

- **exFAT** — reformatting as exFAT gives native read/write on both macOS and
  Windows with no drivers needed. Best choice for data archive drives that don't
  need NTFS-specific features.
- **Paragon NTFS / Tuxera NTFS** — paid kernel extensions; seamless but adds
  a dependency.
- macOS experimental NTFS write (fstab trick) — unreliable; not recommended
  for large files.

---

## Quick wins for recovering space

### Homebrew download cache

```{bash}
#| eval: false
# Removes cached downloads and old formula versions
brew cleanup
# Preview first (dry run):
brew cleanup -n
```

Typically frees 2–5 GB.

### Empty Trash from Finder

`Cmd+Shift+Delete` → Empty Trash, or right-click Trash icon → Empty.
Terminal cannot access `~/.Trash` without Full Disk Access.

### Offload iCloud Drive

In **System Settings → Apple ID → iCloud → iCloud Drive → Optimise Mac Storage**,
enable "Optimise Mac Storage" to keep only recent files locally.

### Move raw data archives to external drive

Once a conversion pipeline has run to completion, raw input files
(ZIP archives, Oracle dumps, etc.) can be moved to an external drive.
Update path variables in your scripts to point to the new location.

---

## Pipeline failure: "No space left on device"

The ASTD ZIP → Parquet script (`data-raw/ais/astd_zip-to-parquet.R`) failed
when the disk filled mid-run.  Two bugs were also found and fixed.

### Root cause

DuckDB's `COPY ... TO` writes an output parquet; if the disk fills
during the write, the file is left **partially written** but present on disk.
On the next run the script's "already exists" check sees the partial file
and silently skips it — so the corrupt file persists indefinitely.

Additionally, each ZIP is first extracted to a temp directory under `/tmp`
via `unzip(zip, exdir = tmp)`.  When `/tmp` is on the same volume as the
output (it is on macOS), a full disk causes `dir.create(tmp)` to fail
silently, `unzip` then errors ("exdir does not exist"), and the loop hits
`next` **without recording the failure in `results`**.
The summary log therefore reported "Failed: 2" while silently dropping
141 months.

### Fix 1: detect temp-dir creation failure

```{r}
#| eval: false
tmp    <- tempfile(pattern = "astd_")
dir_ok <- dir.create(tmp)

if (!dir_ok || !dir.exists(tmp)) {
  warning("Cannot create temp dir for ", basename(zip),
          " — likely no disk space. Stopping pipeline.")
  results[[i]] <- data.frame(yyyymm = ym, ok = FALSE, seconds = 0)
  break   # stop the loop rather than silently skipping all remaining months
}
```

Key changes:

- Capture the return value of `dir.create()` (it returns `FALSE` on failure, not an error).
- Record the failure in `results[[i]]` so it appears in the summary.
- Use `break` instead of `next` — a failed temp dir almost certainly means
  no disk space, so there is no point continuing to the next file.

### Fix 2: delete partial output on DuckDB failure

```{r}
#| eval: false
ok <- tryCatch({
  dbExecute(con, sql)
  TRUE
}, error = function(e) {
  warning("FAILED on ", basename(zip), ": ", conditionMessage(e))
  # Remove any partial parquet left behind by the failed write.
  # Without this, the "already exists" guard would skip it on re-run.
  if (file.exists(out_file)) unlink(out_file)
  FALSE
})
```

### Recovery procedure after a disk-full run

```{bash}
#| eval: false
# 1. Check which parquets exist and their sizes.
#    A suspiciously small file compared to neighbours is likely corrupt.
ls -lh data-raw/ais/astd/year=2014/month=01/

# 2. Delete any partial output manually (if the fixed script wasn't the one
#    that wrote it, it won't have been auto-deleted).
rm data-raw/ais/astd/year=2014/month=01/astd_201401.parquet

# 3. Free up space (move/delete source ZIPs, run brew cleanup, empty Trash).

# 4. Re-run the script — the "already exists" guard means only missing
#    months will be processed.
```

### Verifying the output after a re-run

```{r}
#| eval: false
library(duckdbfs)

# Row counts by year/month — gaps indicate missing or failed months
open_dataset("data-raw/ais/astd") |>
  count(year, month) |>
  arrange(year, month) |>
  collect() |>
  print(n = Inf)
```
