---
title: "Recovering wacky coordinates"
author: "Einar Hjörleifsson"
date: today
---

```{r setup}
#| include: false
library(tidyverse)
library(duckdbfs)
```

## Background

The automatic logger data (`rafr_sjalfvirkir_maelar.parquet`) stores vessel positions
in columns `lengd` (longitude) and `breidd` (latitude) as decimal degrees. However,
an investigation documented in `wackytracks.qmd` revealed that many recorded positions
are implausible — vessel tracks exhibit characteristic zigzag patterns that do not
reflect real movement.

The root cause is a **systematic conversion error**: the original positions were
encoded as integers in the format **DDMMmm** (degrees–minutes–decimal-minutes,
where the last two digits represent hundredths of a minute, ranging 00–99). These
were incorrectly converted to decimal degrees using code that treated the last two
digits as *seconds* (valid range 00–59). When the decimal-minute value was ≥ 60,
the conversion rolled the minute counter over and subtracted 60, placing the reported
position up to ~0.4 nautical miles away from the true location.

The stored decimal-degree values reflect this erroneous conversion. This document
describes how to detect which records are affected, recover the true coordinates,
and resolve ambiguous cases using reported speed and heading.

::: {.callout-important}
## Confidence summary (development subset, 2017)

Every affected record can be assigned one of three recovery-confidence levels based
on the extracted decimal-minutes value ($ss$) for each coordinate:

| Level | Criterion | Positional uncertainty | Share |
|:--|:--|:--|--:|
| **High** | $ss \in [40,59]$ for **both** coordinates | Exact (unique solution) | ~4% |
| **Partial** | $ss \in [40,59]$ for **one** coordinate | ~300–740 m in one direction | ~33% |
| **Low** | $ss \in [0,39]$  for **both** coordinates | ~300–740 m in both directions | ~63% |

High-confidence records are recovered exactly. Partial and low-confidence records
are resolved using the greedy forward-backward smoother (reported speed and heading),
which reduces median track-prediction error by ~21% and median turn angle by 33–48%
compared with simple recovery alone.
:::

---

## Functions

All utility functions are collected here for easy reuse.

```{r funs}
# ── Coordinate conversion ──────────────────────────────────────────────────

# Correct DDMMmm integer → decimal degrees (or inverse)
geo_convert <- function(x, inverse = FALSE, na = FALSE) {
  if (!inverse) {
    i   <- sign(x); x <- abs(x)
    min <- (x / 100) - trunc(x / 10000) * 100
    if (na) min <- ifelse(min > 60, NA, min)
    return((i * (x + (200 / 3) * min)) / 10000)
  } else {
    i  <- sign(x); x <- abs(x)
    p1 <- floor(x)
    p2 <- floor((x - p1) * 60)
    p3 <- round((x - p1 - p2 / 60) * 100 * 60)
    return(i * (p1 * 10000 + p2 * 100 + p3))
  }
}

# ── Wacky-recovery functions ───────────────────────────────────────────────

# Extract the decimal-minutes value (ss) hidden in a wacky decimal degree.
# This reverses the erroneous seconds-based conversion.
extract_ss <- function(w) {
  w  <- abs(w)
  dd <- floor(w)
  mm <- floor((w - dd) * 60)
  ss <- round(((w - dd) * 60 - mm) * 60)
  ss
}

# Interpretation A: non-wacky recovery (treat extracted ss as decimal minutes)
recover_coords <- function(w, sign_w = 1) {
  w  <- abs(w)
  dd <- floor(w)
  mm <- floor((w - dd) * 60)
  ss <- round(((w - dd) * 60 - mm) * 60)
  sign_w * (dd + mm / 60 + ss / 6000)
}

# Interpretation B: wacky recovery (original dm was ss + 60, minute over-counted)
recover_coords_wacky <- function(w, sign_w = 1) {
  w  <- abs(w)
  dd <- floor(w)
  mm <- floor((w - dd) * 60)
  ss <- round(((w - dd) * 60 - mm) * 60)
  sign_w * (dd + (mm - 1) / 60 + (ss + 60) / 6000)
}

# Classify confidence for a single record given its ss values.
# Returns "high" (both coords exact), "partial" (one coord exact),
# or "low" (both coords carry residual ambiguity).
classify_confidence <- function(ss_lon, ss_lat) {
  dplyr::case_when(
    ss_lon >= 40 & ss_lat >= 40 ~ "high",
    ss_lon  < 40 & ss_lat  < 40 ~ "low",
    TRUE                         ~ "partial"
  )
}

# ── Geodetic helpers ───────────────────────────────────────────────────────

# Haversine distance in metres between two WGS-84 points
haversine_m <- function(lon1, lat1, lon2, lat2) {
  R    <- 6371000
  phi1 <- lat1 * pi / 180; phi2 <- lat2 * pi / 180
  dphi <- (lat2 - lat1) * pi / 180
  dlam <- (lon2 - lon1) * pi / 180
  a    <- sin(dphi / 2)^2 + cos(phi1) * cos(phi2) * sin(dlam / 2)^2
  2 * R * asin(sqrt(a))
}

# Forward bearing in degrees (0–360) from point 1 to point 2
bearing_deg <- function(lon1, lat1, lon2, lat2) {
  phi1 <- lat1 * pi / 180; phi2 <- lat2 * pi / 180
  dlam <- (lon2 - lon1) * pi / 180
  y    <- sin(dlam) * cos(phi2)
  x    <- cos(phi1) * sin(phi2) - sin(phi1) * cos(phi2) * cos(dlam)
  (atan2(y, x) * 180 / pi + 360) %% 360
}

# Project a position forward: speed (knots) × elapsed time (mins) on bearing (°)
project_pos <- function(lon, lat, speed_kt, bearing_deg, dt_min) {
  dist_m <- speed_kt * (1852 / 60) * dt_min
  R      <- 6371000
  phi1   <- lat * pi / 180
  lam1   <- lon * pi / 180
  brg    <- bearing_deg * pi / 180
  d_R    <- dist_m / R
  phi2   <- asin(sin(phi1) * cos(d_R) + cos(phi1) * sin(d_R) * cos(brg))
  lam2   <- lam1 + atan2(sin(brg) * sin(d_R) * cos(phi1),
                         cos(d_R) - sin(phi1) * sin(phi2))
  list(lon = lam2 * 180 / pi, lat = phi2 * 180 / pi)
}

# ── Main pipeline functions ────────────────────────────────────────────────

# Add recovery columns and confidence classification to a data frame that has
# columns `lengd` (wacky longitude) and `breidd` (wacky latitude).
add_recovery_cols <- function(df) {
  df |>
    dplyr::mutate(
      ss_lon     = extract_ss(lengd),
      ss_lat     = extract_ss(breidd),
      confidence = classify_confidence(ss_lon, ss_lat),
      # Interpretation A: simple (non-wacky) recovery
      lon_A      = recover_coords(lengd,  sign(lengd)),
      lat_A      = recover_coords(breidd, sign(breidd)),
      # Interpretation B: wacky recovery (only differs when ss < 40)
      lon_B      = dplyr::if_else(ss_lon < 40,
                                  recover_coords_wacky(lengd,  sign(lengd)), lon_A),
      lat_B      = dplyr::if_else(ss_lat < 40,
                                  recover_coords_wacky(breidd, sign(breidd)), lat_A)
    )
}

# Greedy forward-propagation resolver for a single track.
# Input: data frame for one vessel, sorted by timi, with columns
#   lon_A, lat_A, lon_B, lat_B, skip_hradi, skip_stefna, timi.
# Output: same data frame with added lon_r, lat_r (resolved coordinates).
resolve_track <- function(df) {
  n     <- nrow(df)
  lon_r <- numeric(n)
  lat_r <- numeric(n)
  lon_r[1] <- df$lon_A[1]
  lat_r[1] <- df$lat_A[1]

  for (i in seq_len(n - 1)) {
    dt <- as.numeric(difftime(df$timi[i + 1], df$timi[i], units = "mins"))
    # Fall back to interpretation A on large gaps or missing sensor values
    if (is.na(dt) || dt <= 0 || dt > 30 ||
        is.na(df$skip_hradi[i]) || is.na(df$skip_stefna[i])) {
      lon_r[i + 1] <- df$lon_A[i + 1]
      lat_r[i + 1] <- df$lat_A[i + 1]
      next
    }
    pred <- project_pos(lon_r[i], lat_r[i],
                        df$skip_hradi[i], df$skip_stefna[i], dt)
    dA   <- haversine_m(pred$lon, pred$lat, df$lon_A[i + 1], df$lat_A[i + 1])
    dB   <- haversine_m(pred$lon, pred$lat, df$lon_B[i + 1], df$lat_B[i + 1])
    if (is.na(dA) || is.na(dB) || dA <= dB) {
      lon_r[i + 1] <- df$lon_A[i + 1]
      lat_r[i + 1] <- df$lat_A[i + 1]
    } else {
      lon_r[i + 1] <- df$lon_B[i + 1]
      lat_r[i + 1] <- df$lat_B[i + 1]
    }
  }
  dplyr::mutate(df, lon_r = lon_r, lat_r = lat_r)
}
```

```{r data-load}
#| include: false
# Load development subset once here so it is available to all sections,
# including Confidence classification (which precedes Worked example).
siriti_raw <-
  open_dataset("data-raw/data-dump/afli/rafr_sjalfvirkir_maelar.parquet") |>
  mutate(year = year(timi)) |>
  select(year, visir, timi, lengd, breidd, skip_hradi, skip_stefna) |>
  filter(year   == 2017,
         breidd >  65.9,  breidd < 66.10,
         lengd  > -27,    lengd  < -26.4) |>
  collect() |>
  arrange(visir, timi)

siriti_resolved <-
  siriti_raw |>
  add_recovery_cols() |>
  group_by(visir) |>
  group_modify(~ resolve_track(.x)) |>
  ungroup()
```

---

## Coordinate recovery

### Extraction and classification

From any wacky decimal-degree value $w$ we can extract the three integer components
that were embedded in the original DDMMmm encoding:

$$
dd = \lfloor |w| \rfloor, \quad
mm = \lfloor (|w| - dd) \times 60 \rfloor, \quad
ss = \operatorname{round}\bigl(((|w| - dd) \times 60 - mm) \times 60\bigr)
$$

$ss$ is the original decimal-minutes value that was mishandled. The corrected
coordinate is $\mathrm{sign}(w)\times(dd + mm/60 + ss/6000)$.

The ambiguity arises because the wacky conversion is not injective when $ss \ge 60$:

| Extracted $ss$ | Possible original $dm$ | Confidence |
|:-:|:-:|:--|
| 40–59 | Only $dm = ss$ | **High** — wacky case impossible |
| 0–39  | $dm = ss$ *or* $dm = ss + 60$ | **Ambiguous** |

Two candidate positions follow:

- **Interpretation A** (non-wacky): $\text{sign}(w)\times(dd + mm/60 + ss/6000)$  
- **Interpretation B** (wacky): $\text{sign}(w)\times(dd + (mm-1)/60 + (ss+60)/6000)$

The two differ by a **constant 0.00667°** per coordinate (~300 m east–west and ~740 m
north–south at 66°N), so the positional ambiguity is bounded.

```{r classify-example}
#| eval: false
# Apply to a loaded data frame (requires lengd, breidd columns with wacky values)
dat <- add_recovery_cols(dat)

# Classification summary
dat |>
  count(confidence) |>
  mutate(pct = round(100 * n / sum(n), 1))
```

In the 2017 subset used for development (a ~0.6° × 0.2° area off the Icelandic
coast, ~10 000 records):

| Confidence | Description | % |
|:--|:--|--:|
| High | Both coordinates exact | 4 |
| Partial | One coordinate exact, one ambiguous | 33 |
| Low | Both coordinates ambiguous | 63 |

::: {.callout-note}
The low fraction of high-confidence records (~4%) is expected. Assuming the
original decimal-minutes are approximately uniformly distributed over 00–99, only
the 20-in-100 values in [40–59] survive per coordinate, giving ~4% jointly.
The diagnostic histogram of $ss$ values confirms this: density approximately
doubles below 40, because the [0–39] bin contains folded-over wacky cases stacked
on top of genuine non-wacky ones.
:::

---

## Confidence classification {#sec-classification}

### Three-tier scheme

Each record receives a confidence level based on the extracted $ss$ value for
longitude and latitude independently:

| Level | $ss_\text{lon}$ | $ss_\text{lat}$ | Recoverable coordinates | Records (dev. subset) |
|:--|:--:|:--:|:--|--:|
| **High** | 40–59 | 40–59 | Both exact — wacky case impossible | ~4% |
| **Partial** | 40–59 | 0–39 | Longitude exact; latitude ±740 m | ~17% |
| **Partial** | 0–39 | 40–59 | Latitude exact; longitude ±300 m | ~16% |
| **Low** | 0–39 | 0–39 | Both ambiguous | ~63% |

The two partial sub-cases are combined under a single "partial" label because the
resolver treats each coordinate independently; a high-confidence coordinate in
either dimension eliminates its ambiguity entirely.

The ~300 m and ~740 m figures are the constant offset $\Delta = 1/150°$ expressed
in metres at 66°N for longitude and latitude respectively.

### Diagnostic

The fingerprint of the problem is visible in the distribution of extracted $ss$
values (@fig-ss-diag). A density approximately halves above $ss = 40$ for both
coordinates, consistent with roughly half the original decimal-minute values being
in the wacky range [60–99] and folding down into [0–39].

```{r fig-ss-diag}
#| fig-cap: "Distribution of extracted decimal-minutes (ss) for longitude (bottom) and latitude (top). The density step at ss = 40 (red dashed line) is the diagnostic signature of the wacky conversion: values in [0–39] accumulate contributions from both non-wacky (original dm ∈ [0,39]) and wacky (original dm ∈ [60,99], folded down) records, roughly doubling the count relative to [40–59]."
#| fig-height: 4
siriti_resolved |>
  select(ss_lon, ss_lat) |>
  pivot_longer(everything(), names_to = "coord", values_to = "ss") |>
  mutate(coord = recode(coord, ss_lon = "Longitude", ss_lat = "Latitude")) |>
  ggplot(aes(ss)) +
  theme_bw() +
  geom_histogram(binwidth = 1, colour = "white") +
  geom_vline(xintercept = 40, colour = "red", linetype = "dashed") +
  facet_wrap(~coord, ncol = 1) +
  labs(x = "Extracted ss (original decimal-minutes)", y = "Count")
```

### Spatial distribution

@fig-conf-spatial shows confidence levels plotted at their wacky (uncorrected)
positions. The high-confidence points (~4%) are scattered uniformly across the
area — consistent with the expectation that the confidence level is determined
solely by the last two digits of the original integer encoding, not by geography.

```{r fig-conf-spatial}
#| fig-cap: "Recorded (wacky) positions coloured by confidence level: red = high (~4%), orange = partial (~33%), grey = low (~63%). High-confidence points are dispersed uniformly, confirming that confidence is determined by the encoding digits, not by location."
ggplot(siriti_resolved, aes(lengd, breidd, colour = confidence)) +
  theme_bw() +
  geom_point(
    data = filter(siriti_resolved, confidence == "low"),
    colour = "grey80", size = 0.15, alpha = 0.5
  ) +
  geom_point(
    data = filter(siriti_resolved, confidence == "partial"),
    colour = "orange", size = 0.4, alpha = 0.7
  ) +
  geom_point(
    data = filter(siriti_resolved, confidence == "high"),
    colour = "red", size = 0.8
  ) +
  coord_quickmap() +
  labs(x = NULL, y = NULL) +
  guides(colour = "none")
```

---

## Resolving ambiguous records

### Approach: greedy forward propagation

For ambiguous records, the correct interpretation is chosen using the instantaneous
speed (`skip_hradi`, knots) and heading (`skip_stefna`, degrees) recorded alongside
each position. The idea is to predict where the vessel *should* be at the next
timestamp and pick whichever candidate position (A or B) is closer to that prediction.

**Algorithm:**

1. Sort each vessel's records by time (`timi`).
2. Initialise the first record with interpretation A.
3. For each subsequent record $i+1$:
   a. Project the current resolved position forward by $\Delta t$ minutes using
      `skip_hradi` and `skip_stefna` at record $i$.
   b. Compute the Haversine distance from the predicted position to candidates
      $A_{i+1}$ and $B_{i+1}$.
   c. Pick the closer candidate. Ties go to A.
   d. If $\Delta t > 30$ min or speed/heading is missing, fall back to A and
      restart the chain.

```{r resolve-example}
#| eval: false
# Full pipeline: load → add recovery columns → resolve tracks
dat_resolved <-
  open_dataset("rafr_sjalfvirkir_maelar.parquet") |>
  select(visir, timi, lengd, breidd, skip_hradi, skip_stefna) |>
  collect() |>
  add_recovery_cols() |>
  arrange(visir, timi) |>
  group_by(visir) |>
  group_modify(~ resolve_track(.x)) |>
  ungroup()
```

### Evaluation

The resolver was evaluated on the 2017 development subset (~22 800 records). For
each step between consecutive records, a predicted next position was computed from
the current resolved position plus `skip_hradi`/`skip_stefna`. The **prediction
residual** — Haversine distance between the prediction and the actual next resolved
position — was used as the quality metric.

| Method | Median residual |
|:--|--:|
| Simple recovery (A only) | 737 m |
| Greedy resolved | 578 m |
| Improvement | **21.5 %** |

The improvement holds across all confidence categories, including steps where both
endpoints are unambiguous (the benefit there comes from better-resolved neighbours).

::: {.callout-warning}
**Circularity caveat.** The evaluation metric (prediction residual) is the same
quantity the resolver minimises at each step, so the improvement is partly guaranteed
by construction. An independent validation — e.g. using track-smoothness or
comparing against a hold-out set of high-confidence anchor points — is warranted
before drawing strong conclusions about absolute accuracy.
:::

### Choice frequency

Among ambiguous records, the resolver selected:

- **Interpretation A** (non-wacky): 57 %  
- **Interpretation B** (wacky): 43 %

The near-50/50 split is consistent with the theoretical expectation for a roughly
uniform original decimal-minutes distribution.

---

## Worked example

The following code runs against a geographic and temporal subset of the full
dataset: vessels recorded in 2017 within a ~0.6° × 0.2° window off the
Icelandic coast (~22 800 records across ~600 trips). All steps are shown end-to-end.

### Load and process

```{r ex-load}
#| eval: false
siriti_raw <-
  open_dataset("data-raw/data-dump/afli/rafr_sjalfvirkir_maelar.parquet") |>
  mutate(year = year(timi)) |>
  select(year, visir, timi, lengd, breidd, skip_hradi, skip_stefna) |>
  filter(year   == 2017,
         breidd >  65.9,  breidd < 66.10,
         lengd  > -27,    lengd  < -26.4) |>
  collect() |>
  arrange(visir, timi)

siriti_resolved <-
  siriti_raw |>
  add_recovery_cols() |>
  group_by(visir) |>
  group_modify(~ resolve_track(.x)) |>
  ungroup()
```

### Classification summary

```{r ex-classify}
siriti_resolved |>
  count(confidence) |>
  mutate(pct = round(100 * n / sum(n), 1)) |>
  arrange(desc(pct)) |>
  rename(Confidence = confidence, N = n, `%` = pct)
```

See @sec-classification for the full confidence diagnostic, including the $ss$
histogram and spatial distribution plot.

### Track comparison: wacky vs resolved

A set of four trips with at least 20 recorded positions is used to compare the
raw wacky coordinates against the greedy-resolved track (@fig-track-compare).

```{r fig-track-compare}
#| fig-cap: "Wacky positions (grey dashed) versus greedy-resolved positions (solid) for four example trips. Point colour indicates confidence level: red = high (both coordinates exact), orange = partial (one coordinate exact), blue = low (both coordinates ambiguous)."
top_trips <- siriti_resolved |>
  count(visir) |>
  filter(n >= 20) |>
  slice_head(n = 4) |>
  pull(visir)

siriti_resolved |>
  filter(visir %in% top_trips) |>
  arrange(visir, timi) |>
  ggplot(aes(lon_r, lat_r)) +
  theme_bw() +
  geom_path(aes(lengd, breidd), colour = "grey60",
            linewidth = 0.4, linetype = "dashed") +
  geom_path(linewidth = 0.5) +
  geom_point(aes(colour = confidence), size = 1.5) +
  scale_colour_manual(
    values = c("high"    = "red",
               "partial" = "orange",
               "low"     = "steelblue")
  ) +
  facet_wrap(~visir, scales = "free") +
  coord_quickmap() +
  labs(x = NULL, y = NULL, colour = "Confidence")
```

### Residual improvement

For each step between consecutive records, the predicted next position is computed
by dead-reckoning from the current resolved position using `skip_hradi` and
`skip_stefna`. The prediction residual (Haversine distance to actual next position)
is compared between simple recovery (interpretation A only) and the greedy resolver
(@fig-ecdf).

```{r ex-eval}
# Compute step-level residuals under both methods
eval_steps <- siriti_resolved |>
  arrange(visir, timi) |>
  group_by(visir) |>
  mutate(
    dt_min     = as.numeric(difftime(lead(timi), timi, units = "mins")),
    next_lon_A = lead(lon_A), next_lat_A = lead(lat_A),
    next_lon_r = lead(lon_r), next_lat_r = lead(lat_r)
  ) |>
  ungroup() |>
  filter(!is.na(dt_min), dt_min > 0, dt_min <= 30,
         !is.na(skip_hradi), !is.na(skip_stefna))

pred_A <- Map(project_pos,
              eval_steps$lon_A, eval_steps$lat_A,
              eval_steps$skip_hradi, eval_steps$skip_stefna, eval_steps$dt_min)
pred_r <- Map(project_pos,
              eval_steps$lon_r, eval_steps$lat_r,
              eval_steps$skip_hradi, eval_steps$skip_stefna, eval_steps$dt_min)

eval_steps <- eval_steps |>
  mutate(
    res_A = haversine_m(sapply(pred_A, `[[`, "lon"), sapply(pred_A, `[[`, "lat"),
                        next_lon_A, next_lat_A),
    res_r = haversine_m(sapply(pred_r, `[[`, "lon"), sapply(pred_r, `[[`, "lat"),
                        next_lon_r, next_lat_r)
  ) |>
  filter(!is.na(res_A), !is.na(res_r))
```

```{r ex-eval-summary}
eval_steps |>
  summarise(
    `Simple recovery — median residual (m)` = round(median(res_A)),
    `Greedy resolved — median residual (m)`  = round(median(res_r)),
    `Improvement (%)`                        = round(
      (median(res_A) - median(res_r)) / median(res_A) * 100, 1)
  ) |>
  pivot_longer(everything(), names_to = "Metric", values_to = "Value")
```

```{r fig-ecdf}
#| fig-cap: "Empirical CDF of step-level prediction residuals. A higher curve indicates more steps with small residuals (better). The greedy-resolved track (red) is uniformly better than the simple recovery up to ~1500 m."
eval_steps |>
  select(res_A, res_r) |>
  pivot_longer(everything(),
               names_to = "Method", values_to = "residual_m",
               names_pattern = "res_(.*)") |>
  mutate(Method = recode(Method,
                         "A" = "Simple recovery (A only)",
                         "r" = "Greedy resolved")) |>
  ggplot(aes(pmin(residual_m, 3000), colour = Method)) +
  theme_bw() +
  stat_ecdf() +
  scale_x_continuous(breaks = seq(0, 3000, 500)) +
  labs(x = "Prediction residual (m, capped at 3 000)",
       y = "Cumulative proportion",
       colour = NULL)
```

---

## Suggestions for future work

### 1 — Independent validation via track smoothness {#sec-smoothness}

The evaluation in the worked example uses prediction residuals computed from
`skip_hradi` and `skip_stefna` — the same signals the resolver uses — so the
improvement is partly guaranteed by construction. An independent quality metric
is the **turn angle**: the angle between the incoming and outgoing bearing at each
recorded position. A vessel on a realistic course should exhibit small, smooth
direction changes; wacky positions create erratic spikes.

To use gear type as a stratification variable, join the trip table
(`rafr_stofn.parquet`, columns `visir`, `skipnr`, `veidarf`) before computing
turn angles. Two gears are present in the development subset: gear 1 and gear 6
(which accounts for 97% of records).

```{r sug1-setup}
stofn <- open_dataset("data-raw/data-dump/afli/rafr_stofn.parquet") |>
  select(visir, skipnr, veidarf) |>
  collect()

# Join gear info onto the resolved data
siriti_full2 <- siriti_resolved |>
  left_join(stofn, by = "visir")
```

```{r sug-compute}
#| include: false
# Pre-compute forward-backward smoother and turn angles so that both
# suggestion sections can reference the results in order.

adiff <- function(f, b)
  pmin(abs(f - (b + 180) %% 360), 360 - abs(f - (b + 180) %% 360))

bad_step <- function(dt, hr, st)
  isTRUE(is.na(dt)) || isTRUE(dt <= 0) || isTRUE(dt > 30) ||
    isTRUE(is.na(hr)) || isTRUE(is.na(st))

resolve_track_fb <- function(df) {
  n <- nrow(df)
  lon_f <- numeric(n); lat_f <- numeric(n)
  lon_b <- numeric(n); lat_b <- numeric(n)
  lon_f[1] <- df$lon_A[1]; lat_f[1] <- df$lat_A[1]
  for (i in seq_len(n - 1)) {
    dt <- as.numeric(df$timi[i+1] - df$timi[i], units = "mins")
    if (bad_step(dt, df$skip_hradi[i], df$skip_stefna[i])) {
      lon_f[i+1] <- df$lon_A[i+1]; lat_f[i+1] <- df$lat_A[i+1]; next
    }
    pred <- project_pos(lon_f[i], lat_f[i], df$skip_hradi[i], df$skip_stefna[i], dt)
    dA <- haversine_m(pred$lon, pred$lat, df$lon_A[i+1], df$lat_A[i+1])
    dB <- haversine_m(pred$lon, pred$lat, df$lon_B[i+1], df$lat_B[i+1])
    if (isTRUE(is.na(dA)) || isTRUE(is.na(dB)) || isTRUE(dA <= dB)) {
      lon_f[i+1] <- df$lon_A[i+1]; lat_f[i+1] <- df$lat_A[i+1]
    } else {
      lon_f[i+1] <- df$lon_B[i+1]; lat_f[i+1] <- df$lat_B[i+1]
    }
  }
  lon_b[n] <- df$lon_A[n]; lat_b[n] <- df$lat_A[n]
  for (i in seq(n, 2)) {
    dt <- as.numeric(df$timi[i] - df$timi[i-1], units = "mins")
    if (bad_step(dt, df$skip_hradi[i], df$skip_stefna[i])) {
      lon_b[i-1] <- df$lon_A[i-1]; lat_b[i-1] <- df$lat_A[i-1]; next
    }
    rev_hdg <- (df$skip_stefna[i] + 180) %% 360
    pred <- project_pos(lon_b[i], lat_b[i], df$skip_hradi[i], rev_hdg, dt)
    dA <- haversine_m(pred$lon, pred$lat, df$lon_A[i-1], df$lat_A[i-1])
    dB <- haversine_m(pred$lon, pred$lat, df$lon_B[i-1], df$lat_B[i-1])
    if (isTRUE(is.na(dA)) || isTRUE(is.na(dB)) || isTRUE(dA <= dB)) {
      lon_b[i-1] <- df$lon_A[i-1]; lat_b[i-1] <- df$lat_A[i-1]
    } else {
      lon_b[i-1] <- df$lon_B[i-1]; lat_b[i-1] <- df$lat_B[i-1]
    }
  }
  lon_r2 <- numeric(n); lat_r2 <- numeric(n)
  for (i in seq_len(n)) {
    fwd_A <- isTRUE(lon_f[i] == df$lon_A[i] && lat_f[i] == df$lat_A[i])
    bwd_A <- isTRUE(lon_b[i] == df$lon_A[i] && lat_b[i] == df$lat_A[i])
    if (fwd_A == bwd_A) {
      lon_r2[i] <- lon_f[i]; lat_r2[i] <- lat_f[i]
    } else {
      dA_tot <- haversine_m(lon_f[i], lat_f[i], df$lon_A[i], df$lat_A[i]) +
                haversine_m(lon_b[i], lat_b[i], df$lon_A[i], df$lat_A[i])
      dB_tot <- haversine_m(lon_f[i], lat_f[i], df$lon_B[i], df$lat_B[i]) +
                haversine_m(lon_b[i], lat_b[i], df$lon_B[i], df$lat_B[i])
      if (isTRUE(is.na(dA_tot)) || isTRUE(is.na(dB_tot)) || isTRUE(dA_tot <= dB_tot)) {
        lon_r2[i] <- df$lon_A[i]; lat_r2[i] <- df$lat_A[i]
      } else {
        lon_r2[i] <- df$lon_B[i]; lat_r2[i] <- df$lat_B[i]
      }
    }
  }
  dplyr::mutate(df, lon_fb = lon_r2, lat_fb = lat_r2)
}

siriti_fb <- siriti_full2 |>
  arrange(visir, timi) |>
  group_by(visir) |>
  group_modify(~ resolve_track_fb(.x)) |>
  ungroup()

turn_all <- siriti_fb |>
  arrange(visir, timi) |>
  group_by(visir) |>
  mutate(
    dt_fwd    = as.numeric(lead(timi) - timi, units = "mins"),
    dt_bwd    = as.numeric(timi - lag(timi),  units = "mins"),
    valid_seq = !is.na(dt_fwd) & dt_fwd > 0 & dt_fwd <= 30 &
                !is.na(dt_bwd) & dt_bwd > 0 & dt_bwd <= 30,
    turn_A  = adiff(bearing_deg(lon_A,  lat_A,  lead(lon_A),  lead(lat_A)),
                    bearing_deg(lon_A,  lat_A,  lag(lon_A),   lag(lat_A))),
    turn_r  = adiff(bearing_deg(lon_r,  lat_r,  lead(lon_r),  lead(lat_r)),
                    bearing_deg(lon_r,  lat_r,  lag(lon_r),   lag(lat_r))),
    turn_fb = adiff(bearing_deg(lon_fb, lat_fb, lead(lon_fb), lead(lat_fb)),
                    bearing_deg(lon_fb, lat_fb, lag(lon_fb),  lag(lat_fb)))
  ) |>
  ungroup() |>
  filter(valid_seq, !is.na(turn_A), !is.na(turn_r), !is.na(turn_fb))
```

```{r fig-turn}
#| fig-cap: "Distribution of turn angles (°) at each recorded position by gear type and recovery method. Smaller angles indicate smoother, more realistic tracks. The greedy-resolved and forward-backward methods both substantially reduce erratic direction changes compared with simple recovery."
turn_all |>
  select(veidarf, turn_A, turn_r, turn_fb) |>
  pivot_longer(-veidarf, names_to = "method", values_to = "turn") |>
  mutate(
    method  = factor(method,
                     levels = c("turn_A", "turn_r", "turn_fb"),
                     labels = c("Simple recovery", "Greedy (fwd)", "Fwd-bwd smoother")),
    veidarf = paste("Gear", veidarf)
  ) |>
  ggplot(aes(method, turn)) +
  theme_bw() +
  geom_violin(fill = "grey85", colour = NA) +
  geom_boxplot(width = 0.15, outlier.shape = NA) +
  facet_wrap(~veidarf) +
  coord_cartesian(ylim = c(0, 120)) +
  labs(x = NULL, y = "Turn angle (°)")
```

Median turn angles and improvements (independent of reported speed/heading):

```{r sug1-table}
turn_all |>
  group_by(`Gear` = paste("Gear", veidarf)) |>
  summarise(
    n                   = n(),
    `Simple (°)`        = round(median(turn_A),  1),
    `Greedy fwd (°)`    = round(median(turn_r),  1),
    `Fwd-bwd (°)`       = round(median(turn_fb), 1),
    `Fwd improv. (%)`   = round((median(turn_A) - median(turn_r))  / median(turn_A) * 100, 1),
    `Fwd-bwd improv. (%)` = round((median(turn_A) - median(turn_fb)) / median(turn_A) * 100, 1)
  )
```

Both methods reduce median turn angle by **33–48%** depending on gear, confirming
the improvement is genuine and not an artefact of using the same signal for
optimisation and evaluation.

---

### 2 — Forward–backward smoother {#sec-fb}

The greedy algorithm processes records in chronological order, so an incorrect
decision at record $i$ biases all subsequent records until the next time gap. A
**forward–backward smoother** runs two independent passes and combines them:

1. **Forward pass** — identical to the greedy resolver (prediction from speed/heading at $i$).
2. **Backward pass** — travels from end to start; at each step the reversed heading
   ($\theta + 180°$) is used to project backward.
3. **Combine** — if both passes agree, take the agreed interpretation; if they
   disagree, score candidates A and B by the sum of their distances to *both* the
   forward and backward predictions and take the minimum.

```{r sug2-fb-fun}
#| eval: false
bad_step <- function(dt, hr, st)
  isTRUE(is.na(dt)) || isTRUE(dt <= 0) || isTRUE(dt > 30) ||
    isTRUE(is.na(hr)) || isTRUE(is.na(st))

resolve_track_fb <- function(df) {
  n <- nrow(df)
  lon_f <- numeric(n); lat_f <- numeric(n)
  lon_b <- numeric(n); lat_b <- numeric(n)

  # Forward pass
  lon_f[1] <- df$lon_A[1]; lat_f[1] <- df$lat_A[1]
  for (i in seq_len(n - 1)) {
    dt <- as.numeric(df$timi[i+1] - df$timi[i], units = "mins")
    if (bad_step(dt, df$skip_hradi[i], df$skip_stefna[i])) {
      lon_f[i+1] <- df$lon_A[i+1]; lat_f[i+1] <- df$lat_A[i+1]; next
    }
    pred <- project_pos(lon_f[i], lat_f[i], df$skip_hradi[i], df$skip_stefna[i], dt)
    dA <- haversine_m(pred$lon, pred$lat, df$lon_A[i+1], df$lat_A[i+1])
    dB <- haversine_m(pred$lon, pred$lat, df$lon_B[i+1], df$lat_B[i+1])
    if (isTRUE(is.na(dA)) || isTRUE(is.na(dB)) || isTRUE(dA <= dB)) {
      lon_f[i+1] <- df$lon_A[i+1]; lat_f[i+1] <- df$lat_A[i+1]
    } else {
      lon_f[i+1] <- df$lon_B[i+1]; lat_f[i+1] <- df$lat_B[i+1]
    }
  }

  # Backward pass (reversed heading)
  lon_b[n] <- df$lon_A[n]; lat_b[n] <- df$lat_A[n]
  for (i in seq(n, 2)) {
    dt <- as.numeric(df$timi[i] - df$timi[i-1], units = "mins")
    if (bad_step(dt, df$skip_hradi[i], df$skip_stefna[i])) {
      lon_b[i-1] <- df$lon_A[i-1]; lat_b[i-1] <- df$lat_A[i-1]; next
    }
    rev_hdg <- (df$skip_stefna[i] + 180) %% 360
    pred <- project_pos(lon_b[i], lat_b[i], df$skip_hradi[i], rev_hdg, dt)
    dA <- haversine_m(pred$lon, pred$lat, df$lon_A[i-1], df$lat_A[i-1])
    dB <- haversine_m(pred$lon, pred$lat, df$lon_B[i-1], df$lat_B[i-1])
    if (isTRUE(is.na(dA)) || isTRUE(is.na(dB)) || isTRUE(dA <= dB)) {
      lon_b[i-1] <- df$lon_A[i-1]; lat_b[i-1] <- df$lat_A[i-1]
    } else {
      lon_b[i-1] <- df$lon_B[i-1]; lat_b[i-1] <- df$lat_B[i-1]
    }
  }

  # Combine
  lon_r2 <- numeric(n); lat_r2 <- numeric(n)
  for (i in seq_len(n)) {
    fwd_A <- isTRUE(lon_f[i] == df$lon_A[i] && lat_f[i] == df$lat_A[i])
    bwd_A <- isTRUE(lon_b[i] == df$lon_A[i] && lat_b[i] == df$lat_A[i])
    if (fwd_A == bwd_A) {
      lon_r2[i] <- lon_f[i]; lat_r2[i] <- lat_f[i]
    } else {
      dA_tot <- haversine_m(lon_f[i], lat_f[i], df$lon_A[i], df$lat_A[i]) +
                haversine_m(lon_b[i], lat_b[i], df$lon_A[i], df$lat_A[i])
      dB_tot <- haversine_m(lon_f[i], lat_f[i], df$lon_B[i], df$lat_B[i]) +
                haversine_m(lon_b[i], lat_b[i], df$lon_B[i], df$lat_B[i])
      if (isTRUE(is.na(dA_tot)) || isTRUE(is.na(dB_tot)) || isTRUE(dA_tot <= dB_tot)) {
        lon_r2[i] <- df$lon_A[i]; lat_r2[i] <- df$lat_A[i]
      } else {
        lon_r2[i] <- df$lon_B[i]; lat_r2[i] <- df$lat_B[i]
      }
    }
  }
  dplyr::mutate(df, lon_fb = lon_r2, lat_fb = lat_r2)
}
```

```{r sug2-fb-run}
# siriti_fb is computed in the hidden sug-compute chunk above.
# Agreement between forward-only and forward-backward among ambiguous records:
siriti_fb |>
  filter(confidence != "high") |>
  mutate(agree = (lon_r == lon_A) == (lon_fb == lon_A)) |>
  summarise(
    pct_agree  = round(100 * mean(agree, na.rm = TRUE), 1),
    n_disagree = sum(!agree, na.rm = TRUE)
  )
```

The two passes agreed on **78%** of ambiguous records. For the 22% where they
disagreed, the combined-distance tiebreaker resolved the conflict. The additional
smoothness gain over the greedy forward pass was modest (median turn angle improved
by a further 0–1.3° depending on gear — see @fig-turn), suggesting the greedy pass
already captures most of the resolvable signal. The forward–backward approach is
nevertheless preferable for long tracks where early errors could otherwise propagate.

```{r}
#| label: fig-fb-track
#| fig-cap: "Four example trips: grey dashed = simple A recovery; blue = greedy forward; green = forward-backward smoother. Differences are subtle but the smoother tends to correct isolated disagreements mid-track."
top_trips <- siriti_fb |> count(visir) |> filter(n >= 20) |>
  slice_head(n = 4) |> pull(visir)

siriti_fb |>
  filter(visir %in% top_trips) |>
  arrange(visir, timi) |>
  ggplot() +
  theme_bw() +
  geom_path(aes(lengd, breidd), colour = "grey60", linewidth = 0.3, linetype = "dashed") +
  geom_path(aes(lon_r,  lat_r),  colour = "steelblue", linewidth = 0.4) +
  geom_path(aes(lon_fb, lat_fb), colour = "darkgreen", linewidth = 0.6, alpha = 0.7) +
  facet_wrap(~visir, scales = "free") +
  coord_quickmap() +
  labs(x = NULL, y = NULL,
       subtitle = "Grey dashed = wacky · Blue = greedy fwd · Green = fwd-bwd")
```

---

### 3 — Scale to the full dataset {#sec-scale}

The development analysis covered a single 0.6° × 0.2° window in 2017 (~22 800
records). The full archive spans **2008–2020**, comprising roughly 20 million valid
records across ~600 000 fishing trips.

```{r sug3-scale}
open_dataset("data-raw/data-dump/afli/rafr_sjalfvirkir_maelar.parquet") |>
  mutate(year = year(timi)) |>
  filter(year >= 2008, year <= 2020) |>
  group_by(year) |>
  summarise(n_records = n(), n_trips = n_distinct(visir)) |>
  arrange(year) |>
  collect()
```

When scaling up, several of the additional variables become important:

**`veidarf` (gear type, from `rafr_stofn.parquet`).** Different gears produce
fundamentally different movement patterns: trawlers move on long straight hauls,
while gillnetters may drift or be stationary for extended periods. The resolver
performs best when the vessel is actively steaming; for slow or stationary records
the 300–740 m positional ambiguity exceeds a typical step distance and the
speed/heading signal provides little discriminating power. Stratifying evaluation
by gear will reveal where the method works well and where additional signals are
needed.

**`skipnr` (vessel identifier, from `rafr_stofn.parquet`).** Linking records to
individual vessels allows vessel-level quality checks: if a particular vessel
consistently produces wacky coordinates at certain speeds or positions, it may
indicate a vessel-specific recording issue rather than the general conversion
error described here.

**`vindhradi` / `vindstefna` (wind speed and direction).** Wind introduces a
systematic deviation between the vessel's reported heading and its actual
trajectory. A dead-reckoning prediction based on heading alone will over-predict
the leeway component. Incorporating wind as a correction — even crudely, as an
additive offset proportional to wind speed projected onto the cross-track axis —
could sharpen the prediction and improve the resolver's ability to distinguish A
from B in marginal cases.

A production pipeline would look broadly like:

```{r sug3-pipeline}
#| eval: false
dat_full <-
  open_dataset("data-raw/data-dump/afli/rafr_sjalfvirkir_maelar.parquet") |>
  mutate(year = year(timi)) |>
  filter(year >= 2008, year <= 2020) |>
  select(visir, timi, lengd, breidd, skip_hradi, skip_stefna,
         vindhradi, vindstefna) |>
  left_join(
    open_dataset("data-raw/data-dump/afli/rafr_stofn.parquet") |>
      select(visir, skipnr, veidarf),
    by = "visir"
  ) |>
  collect() |>
  add_recovery_cols() |>
  arrange(visir, timi) |>
  group_by(visir) |>
  group_modify(~ resolve_track_fb(.x)) |>   # or resolve_track() for speed
  ungroup()
```

::: {.callout-note}
For the full ~20M record dataset the per-record loop in `resolve_track_fb` will
be slow. Consider rewriting the inner loop in Rcpp, or parallelising over trips
with `furrr::future_map()` / `multidplyr`.
:::


---

## Appendix {.appendix}

### A — Mathematical detail {#sec-math}

**Original encoding.** Positions were stored as signed integers in the DDMMmm
format. For example, $-264312$ encodes 26°43′12″W, where the last two digits (12)
are *decimal minutes* (i.e. 0.12 min), not seconds.

The correct conversion to decimal degrees is

$$
x_{\text{true}} = \operatorname{sign}(n)\times\frac{|n| + \tfrac{200}{3}\,m}{10\,000},
\quad m = \frac{|n|}{100} - 100\left\lfloor\frac{|n|}{10\,000}\right\rfloor
$$

where $n$ is the DDMMmm integer and $m$ is the MM.mm part divided by 100.

**Erroneous conversion (wacky).** The conversion that was actually applied treated
the last two digits $ss$ as seconds:

$$
x_{\text{wacky}} = \begin{cases}
dd + \dfrac{mm}{60} + \dfrac{ss}{3600} & \text{if } ss < 60 \\[6pt]
dd + \dfrac{mm+1}{60} + \dfrac{ss-60}{3600} & \text{if } ss \ge 60
\end{cases}
$$

**Recovery.** Extracting $dd, mm, ss$ from $x_{\text{wacky}}$ via the inverse
seconds formula (see `extract_ss`) recovers the original $ss$ when $ss < 60$
(non-wacky) and recovers $ss - 60$ when $ss \ge 60$ (wacky, with $mm$ over-counted
by 1). The corrected coordinate is:

$$
x_A = \operatorname{sign}(w)\times\!\left(dd + \frac{mm}{60} + \frac{ss}{6000}\right)
\quad \text{(interpretation A, assumes non-wacky)}
$$

$$
x_B = \operatorname{sign}(w)\times\!\left(dd + \frac{mm-1}{60} + \frac{ss+60}{6000}\right)
\quad \text{(interpretation B, assumes wacky)}
$$

The difference $x_A - x_B$ is constant:

$$
\Delta = \frac{1}{60} - \frac{60}{6000} = \frac{1}{60} - \frac{1}{100} = \frac{1}{150} \approx 0.00\overline{6}°
$$

**Identifiability.** If $ss \in [40, 59]$, then $x_B$ would require the original
$dm = ss + 60 \in [100, 119]$, which is impossible (DDMMmm uses hundredths of a
minute, max 99). Hence $x_A$ is the unique correct interpretation — these are the
**high-confidence** records. For $ss \in [0, 39]$, both $dm = ss$ and $dm = ss + 60$
are valid, and the records are **ambiguous**.

---

### B — Diagnostic histogram {#sec-hist}

The histogram of extracted $ss$ values provides a visual fingerprint of the problem.
Under a uniform distribution of original decimal-minutes:

- $ss \in [0, 39]$: receives contributions from both non-wacky ($dm \in [0,39]$)
  and wacky ($dm \in [60,99]$, folded down by 60) cases → density ~2×
- $ss \in [40, 59]$: receives contributions only from non-wacky ($dm \in [40,59]$) → density ~1×

A step-change (roughly halving) at $ss = 40$ is therefore the expected signature
when ~50% of original dm values were in the wacky range. The 2017 data shows this
pattern clearly for both longitude and latitude.

```{r diag-hist}
#| eval: false
dat |>
  mutate(ss_lon = extract_ss(lengd), ss_lat = extract_ss(breidd)) |>
  pivot_longer(c(ss_lon, ss_lat), names_to = "coord", values_to = "ss") |>
  ggplot(aes(ss)) +
  geom_histogram(binwidth = 1) +
  facet_wrap(~coord, ncol = 1) +
  geom_vline(xintercept = c(40, 59), colour = "red", linetype = "dashed") +
  labs(x = "Extracted decimal-minutes (ss)",
       y = "Count",
       subtitle = "Density ~2× below ss = 40 confirms folding of wacky records")
```

---

### C — Algorithm detail {#sec-algo}

**Greedy forward propagation** processes each track sequentially in time. At step
$i \to i+1$:

1. **Predict** next position using the dead-reckoning formula (great-circle
   projection from current resolved position, speed $v_i$ in knots,
   heading $\theta_i$ in degrees, elapsed time $\Delta t$ in minutes):

$$
\phi_2 = \arcsin\!\left(\sin\phi_1\cos\frac{d}{R} + \cos\phi_1\sin\frac{d}{R}\cos\theta_i\right)
$$
$$
\lambda_2 = \lambda_1 + \arctan2\!\left(\sin\theta_i\sin\frac{d}{R}\cos\phi_1,\ \cos\frac{d}{R}-\sin\phi_1\sin\phi_2\right)
$$

where $d = v_i \times \tfrac{1852}{60} \times \Delta t$ (metres) and $R = 6\,371\,000$ m.

2. **Score** candidates $A_{i+1}$ and $B_{i+1}$ by Haversine distance to the prediction.

3. **Select** the closer candidate; ties go to A.

4. **Fallback** to A when $\Delta t > 30$ min or speed/heading is missing,
   resetting the propagation chain.

**Limitations:**

- Errors can propagate forward (a wrong early decision biases all downstream choices).
- The 30-minute gap threshold is arbitrary; shorter gaps reduce propagation length
  but increase the number of cold-starts.
- The algorithm makes no use of the recorded heading or speed at the *destination*
  record; incorporating that signal (e.g. weighting by heading consistency at both
  ends of the step) could improve resolution.
- Performance degrades for slow-moving or stationary vessels, where the step
  distance is smaller than the ~300–740 m ambiguity.
