---
title: "Wacky coordinate tracks — discovery, mechanism, and recovery"
author: "Einar Hjörleifsson"
date: today
---

<!-- ═══ Hidden setup: libraries, functions, and pre-computed data ══════════ -->

```{r setup}
#| include: false
library(tidyverse)
library(duckdbfs)
```

```{r funs}
#| include: false

# ── Coordinate conversion ─────────────────────────────────────────────────────

# Correct DDMMmm integer → decimal degrees (or inverse)
geo_convert <- function(x, inverse = FALSE, na = FALSE) {
  if (!inverse) {
    i   <- sign(x); x <- abs(x)
    min <- (x / 100) - trunc(x / 10000) * 100
    if (na) min <- ifelse(min > 60, NA, min)
    return((i * (x + (200 / 3) * min)) / 10000)
  } else {
    i  <- sign(x); x <- abs(x)
    p1 <- floor(x)
    p2 <- floor((x - p1) * 60)
    p3 <- round((x - p1 - p2 / 60) * 100 * 60)
    return(i * (p1 * 10000 + p2 * 100 + p3))
  }
}

# Simulate the erroneous wacky conversion (DDMMmm integer → wacky decimal)
geo_convert_wacky <- function(x) {
  x  <- abs(x)
  dd <- trunc(x / 1e4)
  mm <- trunc((x - dd * 1e4) / 1e2)
  ss <- x - trunc(x / 1e2) * 1e2
  mm <- ifelse(ss >= 60, mm + 1, mm)
  ss <- ifelse(ss >= 60, ss - 60, ss)
  dd + mm / 60 + ss / 3600
}

# ── Recovery functions ────────────────────────────────────────────────────────

# Extract the decimal-minutes value (ss) hidden inside a wacky decimal degree
extract_ss <- function(w) {
  w  <- abs(w)
  dd <- floor(w)
  mm <- floor((w - dd) * 60)
  round(((w - dd) * 60 - mm) * 60)
}

# Interpretation A: treat extracted ss as decimal minutes (non-wacky)
recover_coords <- function(w, sign_w = 1) {
  w  <- abs(w)
  dd <- floor(w)
  mm <- floor((w - dd) * 60)
  ss <- round(((w - dd) * 60 - mm) * 60)
  sign_w * (dd + mm / 60 + ss / 6000)
}

# Interpretation B: original dm was ss + 60, minute was over-counted (wacky)
recover_coords_wacky <- function(w, sign_w = 1) {
  w  <- abs(w)
  dd <- floor(w)
  mm <- floor((w - dd) * 60)
  ss <- round(((w - dd) * 60 - mm) * 60)
  sign_w * (dd + (mm - 1) / 60 + (ss + 60) / 6000)
}

classify_confidence <- function(ss_lon, ss_lat) {
  dplyr::case_when(
    ss_lon >= 40 & ss_lat >= 40 ~ "high",
    ss_lon  < 40 & ss_lat  < 40 ~ "low",
    TRUE                         ~ "partial"
  )
}

add_recovery_cols <- function(df) {
  df |>
    dplyr::mutate(
      ss_lon     = extract_ss(lengd),
      ss_lat     = extract_ss(breidd),
      confidence = classify_confidence(ss_lon, ss_lat),
      lon_A      = recover_coords(lengd,  sign(lengd)),
      lat_A      = recover_coords(breidd, sign(breidd)),
      lon_B      = dplyr::if_else(ss_lon < 40,
                                  recover_coords_wacky(lengd,  sign(lengd)), lon_A),
      lat_B      = dplyr::if_else(ss_lat < 40,
                                  recover_coords_wacky(breidd, sign(breidd)), lat_A)
    )
}

# ── Geodetic helpers ──────────────────────────────────────────────────────────

haversine_m <- function(lon1, lat1, lon2, lat2) {
  R    <- 6371000
  phi1 <- lat1 * pi / 180; phi2 <- lat2 * pi / 180
  dphi <- (lat2 - lat1) * pi / 180
  dlam <- (lon2 - lon1) * pi / 180
  a    <- sin(dphi / 2)^2 + cos(phi1) * cos(phi2) * sin(dlam / 2)^2
  2 * R * asin(sqrt(a))
}

bearing_deg <- function(lon1, lat1, lon2, lat2) {
  phi1 <- lat1 * pi / 180; phi2 <- lat2 * pi / 180
  dlam <- (lon2 - lon1) * pi / 180
  y    <- sin(dlam) * cos(phi2)
  x    <- cos(phi1) * sin(phi2) - sin(phi1) * cos(phi2) * cos(dlam)
  (atan2(y, x) * 180 / pi + 360) %% 360
}

project_pos <- function(lon, lat, speed_kt, hdg_deg, dt_min) {
  dist_m <- speed_kt * (1852 / 60) * dt_min
  R      <- 6371000
  phi1   <- lat * pi / 180
  lam1   <- lon * pi / 180
  brg    <- hdg_deg * pi / 180
  d_R    <- dist_m / R
  phi2   <- asin(sin(phi1) * cos(d_R) + cos(phi1) * sin(d_R) * cos(brg))
  lam2   <- lam1 + atan2(sin(brg) * sin(d_R) * cos(phi1),
                         cos(d_R) - sin(phi1) * sin(phi2))
  list(lon = lam2 * 180 / pi, lat = phi2 * 180 / pi)
}

# ── Track resolvers ───────────────────────────────────────────────────────────

resolve_track <- function(df) {
  n     <- nrow(df)
  lon_r <- numeric(n); lat_r <- numeric(n)
  lon_r[1] <- df$lon_A[1]; lat_r[1] <- df$lat_A[1]
  for (i in seq_len(n - 1)) {
    dt <- as.numeric(difftime(df$timi[i + 1], df$timi[i], units = "mins"))
    if (is.na(dt) || dt <= 0 || dt > 30 ||
        is.na(df$skip_hradi[i]) || is.na(df$skip_stefna[i])) {
      lon_r[i + 1] <- df$lon_A[i + 1]; lat_r[i + 1] <- df$lat_A[i + 1]; next
    }
    pred <- project_pos(lon_r[i], lat_r[i], df$skip_hradi[i], df$skip_stefna[i], dt)
    dA   <- haversine_m(pred$lon, pred$lat, df$lon_A[i + 1], df$lat_A[i + 1])
    dB   <- haversine_m(pred$lon, pred$lat, df$lon_B[i + 1], df$lat_B[i + 1])
    if (is.na(dA) || is.na(dB) || dA <= dB) {
      lon_r[i + 1] <- df$lon_A[i + 1]; lat_r[i + 1] <- df$lat_A[i + 1]
    } else {
      lon_r[i + 1] <- df$lon_B[i + 1]; lat_r[i + 1] <- df$lat_B[i + 1]
    }
  }
  dplyr::mutate(df, lon_r = lon_r, lat_r = lat_r)
}

bad_step <- function(dt, hr, st)
  isTRUE(is.na(dt)) || isTRUE(dt <= 0) || isTRUE(dt > 30) ||
    isTRUE(is.na(hr)) || isTRUE(is.na(st))

resolve_track_fb <- function(df) {
  n <- nrow(df)
  lon_f <- numeric(n); lat_f <- numeric(n)
  lon_b <- numeric(n); lat_b <- numeric(n)
  lon_f[1] <- df$lon_A[1]; lat_f[1] <- df$lat_A[1]
  for (i in seq_len(n - 1)) {
    dt <- as.numeric(df$timi[i + 1] - df$timi[i], units = "mins")
    if (bad_step(dt, df$skip_hradi[i], df$skip_stefna[i])) {
      lon_f[i + 1] <- df$lon_A[i + 1]; lat_f[i + 1] <- df$lat_A[i + 1]; next
    }
    pred <- project_pos(lon_f[i], lat_f[i], df$skip_hradi[i], df$skip_stefna[i], dt)
    dA <- haversine_m(pred$lon, pred$lat, df$lon_A[i + 1], df$lat_A[i + 1])
    dB <- haversine_m(pred$lon, pred$lat, df$lon_B[i + 1], df$lat_B[i + 1])
    if (isTRUE(is.na(dA)) || isTRUE(is.na(dB)) || isTRUE(dA <= dB)) {
      lon_f[i + 1] <- df$lon_A[i + 1]; lat_f[i + 1] <- df$lat_A[i + 1]
    } else {
      lon_f[i + 1] <- df$lon_B[i + 1]; lat_f[i + 1] <- df$lat_B[i + 1]
    }
  }
  lon_b[n] <- df$lon_A[n]; lat_b[n] <- df$lat_A[n]
  for (i in seq(n, 2)) {
    dt <- as.numeric(df$timi[i] - df$timi[i - 1], units = "mins")
    if (bad_step(dt, df$skip_hradi[i], df$skip_stefna[i])) {
      lon_b[i - 1] <- df$lon_A[i - 1]; lat_b[i - 1] <- df$lat_A[i - 1]; next
    }
    rev_hdg <- (df$skip_stefna[i] + 180) %% 360
    pred <- project_pos(lon_b[i], lat_b[i], df$skip_hradi[i], rev_hdg, dt)
    dA <- haversine_m(pred$lon, pred$lat, df$lon_A[i - 1], df$lat_A[i - 1])
    dB <- haversine_m(pred$lon, pred$lat, df$lon_B[i - 1], df$lat_B[i - 1])
    if (isTRUE(is.na(dA)) || isTRUE(is.na(dB)) || isTRUE(dA <= dB)) {
      lon_b[i - 1] <- df$lon_A[i - 1]; lat_b[i - 1] <- df$lat_A[i - 1]
    } else {
      lon_b[i - 1] <- df$lon_B[i - 1]; lat_b[i - 1] <- df$lat_B[i - 1]
    }
  }
  lon_r2 <- numeric(n); lat_r2 <- numeric(n)
  for (i in seq_len(n)) {
    fwd_A <- isTRUE(lon_f[i] == df$lon_A[i] && lat_f[i] == df$lat_A[i])
    bwd_A <- isTRUE(lon_b[i] == df$lon_A[i] && lat_b[i] == df$lat_A[i])
    if (fwd_A == bwd_A) {
      lon_r2[i] <- lon_f[i]; lat_r2[i] <- lat_f[i]
    } else {
      dA_tot <- haversine_m(lon_f[i], lat_f[i], df$lon_A[i], df$lat_A[i]) +
                haversine_m(lon_b[i], lat_b[i], df$lon_A[i], df$lat_A[i])
      dB_tot <- haversine_m(lon_f[i], lat_f[i], df$lon_B[i], df$lat_B[i]) +
                haversine_m(lon_b[i], lat_b[i], df$lon_B[i], df$lat_B[i])
      if (isTRUE(is.na(dA_tot)) || isTRUE(is.na(dB_tot)) || isTRUE(dA_tot <= dB_tot)) {
        lon_r2[i] <- df$lon_A[i]; lat_r2[i] <- df$lat_A[i]
      } else {
        lon_r2[i] <- df$lon_B[i]; lat_r2[i] <- df$lat_B[i]
      }
    }
  }
  dplyr::mutate(df, lon_fb = lon_r2, lat_fb = lat_r2)
}

adiff <- function(f, b)
  pmin(abs(f - (b + 180) %% 360), 360 - abs(f - (b + 180) %% 360))
```

```{r data-load}
#| include: false
# Development subset: 2017, small area off the Icelandic coast (~22 800 records)
siriti_raw <-
  open_dataset("data-raw/data-dump/afli/rafr_sjalfvirkir_maelar.parquet") |>
  mutate(year = year(timi)) |>
  select(year, visir, timi, lengd, breidd, skip_hradi, skip_stefna) |>
  filter(year   == 2017,
         breidd >  65.9,  breidd < 66.10,
         lengd  > -27,    lengd  < -26.4) |>
  collect() |>
  arrange(visir, timi)

stofn_gear <-
  open_dataset("data-raw/data-dump/afli/rafr_stofn.parquet") |>
  select(visir, skipnr, veidarf) |>
  collect()

siriti_resolved <-
  siriti_raw |>
  add_recovery_cols() |>
  group_by(visir) |>
  group_modify(~ resolve_track(.x)) |>
  ungroup()

siriti_fb <- siriti_resolved |>
  left_join(stofn_gear, by = "visir") |>
  arrange(visir, timi) |>
  group_by(visir) |>
  group_modify(~ resolve_track_fb(.x)) |>
  ungroup()

turn_all <- siriti_fb |>
  arrange(visir, timi) |>
  group_by(visir) |>
  mutate(
    dt_fwd    = as.numeric(lead(timi) - timi, units = "mins"),
    dt_bwd    = as.numeric(timi - lag(timi),  units = "mins"),
    valid_seq = !is.na(dt_fwd) & dt_fwd > 0 & dt_fwd <= 30 &
                !is.na(dt_bwd) & dt_bwd > 0 & dt_bwd <= 30,
    turn_A  = adiff(bearing_deg(lon_A,  lat_A,  lead(lon_A),  lead(lat_A)),
                    bearing_deg(lon_A,  lat_A,  lag(lon_A),   lag(lat_A))),
    turn_r  = adiff(bearing_deg(lon_r,  lat_r,  lead(lon_r),  lead(lat_r)),
                    bearing_deg(lon_r,  lat_r,  lag(lon_r),   lag(lat_r))),
    turn_fb = adiff(bearing_deg(lon_fb, lat_fb, lead(lon_fb), lead(lat_fb)),
                    bearing_deg(lon_fb, lat_fb, lag(lon_fb),  lag(lat_fb)))
  ) |>
  ungroup() |>
  filter(valid_seq, !is.na(turn_A), !is.na(turn_r), !is.na(turn_fb))

# One example vessel for the simulated-path figure
ex_visir <- siriti_raw |> count(visir) |> filter(n >= 30) |> slice(1) |> pull(visir)
```

---

## The problem

The automatic logger table `rafr_sjalfvirkir_maelar.parquet` stores vessel
positions as decimal-degree coordinates. When those positions are plotted as
tracks, many vessels show **characteristic zigzag patterns** that no vessel
could actually travel:

```{r fig-wacky-tracks}
#| fig-cap: "A sample of 10 000 position records from one small area (2017). Many tracks exhibit the erratic zigzag signature of the wacky conversion error."
#| fig-height: 5
siriti_raw |>
  slice_head(n = 1e4) |>
  ggplot(aes(lengd, breidd, group = visir)) +
  theme_bw() +
  geom_point(colour = "red", size = 0.25) +
  geom_path(linewidth = 0.2, alpha = 0.5) +
  coord_quickmap() +
  labs(x = NULL, y = NULL)
```

The error is not noise — it is **systematic and predictable**. The same pattern
appears in both `rafr_sjalfvirkir_maelar.parquet` and the older
`sjalfvirkir_maelar.parquet`; the two tables carry identical coordinate values
for all overlapping records.

---

## Root cause: a coordinate encoding error

Positions were originally encoded as **DDMMmm integers** — a six-digit format
where the last two digits represent hundredths of a minute (range 00–99), not
seconds (range 00–59). For example, the integer `642783` encodes
64°27′.83 N = 64.4638°N.

The logbook software (Trackwell / SeaData) converted these integers to decimal
degrees by **treating the last two digits as seconds**. When the decimal-minute
value was ≥ 60 — which is perfectly valid in DDMMmm but impossible as
seconds — the software rolled the minute counter up and subtracted 60:

$$
x_{\text{wacky}} = \begin{cases}
dd + \dfrac{mm}{60} + \dfrac{dm}{3600} & \text{if } dm < 60 \\[6pt]
dd + \dfrac{mm+1}{60} + \dfrac{dm-60}{3600} & \text{if } dm \ge 60
\end{cases}
$$

where $dm$ is the hundredths-of-a-minute value (the true last two digits of the
DDMMmm integer). The resulting displacement is up to **~0.4 NM** (≈ 740 m
north–south, ≈ 300 m east–west at 66°N) and occurs for roughly half of all
records — those whose original decimal-minute value was ≥ 60.

The following figure makes this concrete. A simulated straight tow path (green)
is encoded as DDMMmm integers and then decoded correctly (green) and wackily
(blue). Grey points mark positions where *both* coordinates happen to fall below
the 60 threshold and are therefore error-free. Yellow segments show the
displacement for each point that crosses the threshold.

```{r fig-simulated-path}
#| fig-cap: "Simulated true tow path (green dots/line) and corresponding wacky path (blue). Grey area = positions where both coordinates are error-free (original dm < 60 for both). Yellow segments connect each true point to its displaced wacky counterpart. Red = one actual recorded track from the data."
#| fig-height: 5

# Grey background: grid points that survive the ss < 60 test
safe_bg <- expand.grid(lon = seq(-27, -26.4, by = 0.0001),
                       lat = seq(65.9,  66.1, by = 0.0001)) |>
  as_tibble() |>
  mutate(xi = -geo_convert(lon, inverse = TRUE),
         yi =  geo_convert(lat, inverse = TRUE),
         xss = xi - trunc(xi / 1e2) * 1e2,
         yss = yi - trunc(yi / 1e2) * 1e2) |>
  filter(xss < 60 & yss < 60) |>
  sample_n(2.5e5)

# Simulated diagonal tow path
sim_path <- tibble(lon = seq(-27, -26.4, length.out = 50),
                   lat = seq(65.90, 66.10, length.out = 50)) |>
  mutate(xi   =  geo_convert(lon, inverse = TRUE),
         yi   =  geo_convert(lat, inverse = TRUE),
         lon  =  geo_convert(xi),
         lat  =  geo_convert(yi),
         lonw = -geo_convert_wacky(xi),
         latw =  geo_convert_wacky(yi))

# One real track
ex_track <- siriti_raw |> filter(visir == ex_visir)

safe_bg |>
  ggplot(aes(lon, lat)) +
  theme_bw() +
  geom_point(size = 0.5, colour = "grey85") +
  geom_segment(data = sim_path, aes(lon, lat, xend = lonw, yend = latw),
               colour = "orange", linewidth = 0.3) +
  geom_path(data = sim_path, aes(lon, lat), colour = "darkgreen") +
  geom_point(data = sim_path, aes(lon, lat), colour = "darkgreen", size = 1) +
  geom_path(data = sim_path, aes(lonw, latw), colour = "blue") +
  geom_point(data = sim_path, aes(lonw, latw), colour = "blue", size = 1) +
  geom_path(data = ex_track, aes(lengd, breidd), colour = "red", alpha = 0.6) +
  geom_point(data = ex_track, aes(lengd, breidd), colour = "red",
             size = 0.8, alpha = 0.6) +
  coord_quickmap() +
  labs(x = NULL, y = NULL)
```

---

## Affected data

### Which tables

The error is present in two Oracle tables, both dumped to parquet:

| Table | Rows | Period | Column names |
|:--|--:|:--|:--|
| `rafr_sjalfvirkir_maelar.parquet` | 23.6 M | 2008–2020 | `lengd`, `breidd` |
| `sjalfvirkir_maelar.parquet` | 28.2 M | older, wider | `lengd`, `breidd` |

The two tables carry **identical coordinate values** for all overlapping records
(685 656 shared `visir`). The error propagated from the sender, not from any
Oracle import step.

The station-level coordinates in `rafr_stofn.parquet` (`breidd`, `lengd`) are
**not affected** — they are stored as raw DDMM integers and are correctly
converted by `geo::geoconvert.1()` in the convert script.

### Evidence that the error is sender-side

The same XML submissions that populate `rafr_sjalfvirkir_maelar` also populate
`rafr_stofn` and are archived verbatim in `afladagb_xml_mottaka.parquet`. Within
each XML document, station positions appear in `STOFN/STADSETNING` as raw DDMM
integers (correct), while logger positions appear in
`SJALFVIRKIR_MAELAR/LOG/STADSETNING` as the pre-converted decimal degrees
(wacky). Oracle stored both fields verbatim — **no conversion was applied at
import**. The error was introduced before transmission.

### Record counts by year

```{r scale-table}
open_dataset("data-raw/data-dump/afli/rafr_sjalfvirkir_maelar.parquet") |>
  mutate(year = year(timi)) |>
  filter(year >= 2008, year <= 2020) |>
  group_by(year) |>
  summarise(n_records = n(), n_trips = n_distinct(visir)) |>
  arrange(year) |>
  collect()
```

---

## Recovery

### Classification

Every stored position can be assigned a recovery confidence based on a
**diagnostic value extracted from the wacky decimal itself**. Reversing the
erroneous seconds-based conversion yields an extracted $ss$ value (see
[Appendix — Mathematics](#sec-math)):

| $ss_\text{lon}$ | $ss_\text{lat}$ | Confidence | Recoverable? |
|:--:|:--:|:--|:--|
| 40–59 | 40–59 | **High** (~4%) | Both exact — $ss+60$ would exceed 99 |
| 40–59 | 0–39 | **Partial** (~17%) | Longitude exact; latitude ambiguous ±740 m |
| 0–39 | 40–59 | **Partial** (~16%) | Latitude exact; longitude ambiguous ±300 m |
| 0–39 | 0–39 | **Low** (~63%) | Both ambiguous |

High-confidence records have a unique correct interpretation. For low and partial
records, two candidate positions (A and B) exist, differing by a **constant
$1/150° \approx 0.0067°$** per coordinate.

```{r fig-ss-diag}
#| fig-cap: "Distribution of extracted ss values for the 2017 development subset. Density roughly doubles below ss = 40 (red line): the [0–39] bin accumulates both genuine non-wacky records and wacky records folded down from [60–99]. This step-change is the fingerprint of the error."
#| fig-height: 4
siriti_resolved |>
  select(ss_lon, ss_lat) |>
  pivot_longer(everything(), names_to = "coord", values_to = "ss") |>
  mutate(coord = recode(coord, ss_lon = "Longitude", ss_lat = "Latitude")) |>
  ggplot(aes(ss)) +
  theme_bw() +
  geom_histogram(binwidth = 1, colour = "white") +
  geom_vline(xintercept = 40, colour = "red", linetype = "dashed") +
  facet_wrap(~coord, ncol = 1) +
  labs(x = "Extracted ss (original decimal-minutes)", y = "Count")
```

High-confidence points are scattered uniformly across the area — confidence is
determined by the encoding digits, not by geography:

```{r fig-conf-spatial}
#| fig-cap: "Wacky positions coloured by confidence. Red = high (~4%), orange = partial (~33%), grey = low (~63%). The uniform scatter of high-confidence points confirms that confidence is driven by the encoding, not by location."
ggplot(siriti_resolved, aes(lengd, breidd)) +
  theme_bw() +
  geom_point(data = filter(siriti_resolved, confidence == "low"),
             colour = "grey80", size = 0.15, alpha = 0.5) +
  geom_point(data = filter(siriti_resolved, confidence == "partial"),
             colour = "orange", size = 0.4, alpha = 0.7) +
  geom_point(data = filter(siriti_resolved, confidence == "high"),
             colour = "red", size = 0.8) +
  coord_quickmap() +
  labs(x = NULL, y = NULL)
```

### Resolving ambiguous records

For ambiguous records, the correct interpretation is chosen using the
instantaneous speed (`skip_hradi`, knots) and heading (`skip_stefna`, degrees)
recorded alongside each position. The **greedy forward propagation** algorithm
dead-reckons where the vessel *should* be at the next timestamp and picks
whichever candidate (A or B) is closer to that prediction. A
**forward–backward smoother** runs two independent passes and resolves
disagreements by picking the candidate minimising the sum of distances to both
predictions (see [Appendix — Algorithm](#sec-algo)).

Among ambiguous records in the development subset, the resolver selected
interpretation A (non-wacky) 57% of the time and B (wacky) 43% of the time —
consistent with the near-uniform distribution of original decimal-minute values.

### Performance

Performance was assessed using step-level prediction residuals and turn angles
— both independent of the resolver's optimisation target (see [Appendix —
Evaluation detail](#sec-eval)).

```{r perf-table}
#| tbl-cap: "Median turn angle (°) by gear type and method. Smaller = smoother track. Gear 6 = bottom trawl, Gear 1 = longline."
turn_all |>
  group_by(`Gear` = paste("Gear", veidarf)) |>
  summarise(
    n                     = n(),
    `Simple (°)`          = round(median(turn_A),  1),
    `Greedy fwd (°)`      = round(median(turn_r),  1),
    `Fwd-bwd (°)`         = round(median(turn_fb), 1),
    `Fwd improv. (%)`     = round((median(turn_A) - median(turn_r))  / median(turn_A) * 100, 1),
    `Fwd-bwd improv. (%)` = round((median(turn_A) - median(turn_fb)) / median(turn_A) * 100, 1)
  )
```

```{r fig-turn}
#| fig-cap: "Turn-angle distributions by gear and recovery method. The greedy and forward-backward resolvers substantially reduce erratic direction changes relative to simple recovery (A only)."
turn_all |>
  select(veidarf, turn_A, turn_r, turn_fb) |>
  pivot_longer(-veidarf, names_to = "method", values_to = "turn") |>
  mutate(
    method  = factor(method,
                     levels = c("turn_A", "turn_r", "turn_fb"),
                     labels = c("Simple recovery", "Greedy (fwd)", "Fwd-bwd smoother")),
    veidarf = paste("Gear", veidarf)
  ) |>
  ggplot(aes(method, turn)) +
  theme_bw() +
  geom_violin(fill = "grey85", colour = NA) +
  geom_boxplot(width = 0.15, outlier.shape = NA) +
  facet_wrap(~veidarf) +
  coord_cartesian(ylim = c(0, 120)) +
  labs(x = NULL, y = "Turn angle (°)")
```

### Track comparison

```{r fig-track-compare}
#| fig-cap: "Four example trips: grey dashed = raw wacky positions; solid line = forward-backward resolved; points coloured by confidence."
top_trips <- siriti_fb |>
  count(visir) |>
  filter(n >= 20) |>
  slice_head(n = 4) |>
  pull(visir)

siriti_fb |>
  filter(visir %in% top_trips) |>
  arrange(visir, timi) |>
  ggplot(aes(lon_fb, lat_fb)) +
  theme_bw() +
  geom_path(aes(lengd, breidd), colour = "grey60",
            linewidth = 0.4, linetype = "dashed") +
  geom_path(linewidth = 0.5) +
  geom_point(aes(colour = confidence), size = 1.5) +
  scale_colour_manual(values = c(high = "red", partial = "orange",
                                 low = "steelblue")) +
  facet_wrap(~visir, scales = "free") +
  coord_quickmap() +
  labs(x = NULL, y = NULL, colour = "Confidence")
```

---

## AIS-based recovery (alternative approach)

An independent recovery strategy is implemented in `scripts/01_siritar.R`.
Rather than resolving ambiguous positions algorithmically, this script
**interpolates corrected positions directly from AIS tracks** (`data/trail`,
copied from `../fishydata`), which are free of the DDMMmm encoding error.

The approach:

1. Load both GPS logger tables (`sjalfvirkir_maelar`, `rafr_sjalfvirkir_maelar`)
   and rename columns to a unified schema (`w_lon`/`w_lat` retain the original
   wacky values).
2. For each year 2008–2023, load AIS records for the vessels present in the GPS
   data and combine the two sources in time order.
3. Use `approx(..., method = "linear", rule = 1)` to interpolate AIS lon/lat
   onto every GPS timestamp. AIS positions serve as anchor points; GPS rows
   receive interpolated coordinates. Where AIS does not bracket a GPS timestamp
   (`rule = 1`), the corrected position is `NA`.
4. Filter back to GPS-source rows only and write the result to
   `data/afli/sensor_XXX.parquet`.

This approach is conceptually simpler than the forward–backward resolver and
eliminates the ~63% ambiguous-record problem entirely, at the cost of depending
on AIS coverage. Rows where AIS is absent or sparse will have `NA` lon/lat.

**`dt_sec` interpretation.** The column records the time gap between adjacent
rows in the combined gps + AIS sequence *before* AIS rows are filtered out.
It therefore approximates the **interpolation interval** around each GPS
timestamp — how far apart the bracketing AIS anchors were in time. Larger
`dt_sec` means the interpolated position is less constrained.

**`NA` positions.** `approx(..., rule = 1)` suppresses extrapolation. GPS
records that fall outside the time range of available AIS data for a vessel
will have `lon = lat = NA`. The scale of this gap is not yet quantified.

**Status:** schema duplication bug fixed; output renamed to `sensor_GPS.parquet`.
Remaining: run at full scale; characterise NA coverage; write to `data/afli/`.

---

## Outstanding work

- **Run `scripts/01_siritar.R` at full scale** — write corrected parquet to
  `data/afli/sensor_GPS.parquet` and notify `../fishydata`.
- **Characterise AIS coverage gaps** — quantify how many sensor records end up
  with `NA` corrected positions (AIS does not bracket their timestamps).
- **Consider contacting Trackwell/SeaData** for the original DDMMmm integer
  values — this would give exact positions for all records without any coverage
  dependency.
- **Clarify `logbook/` schema exposure** — determine whether the experimental
  `logbook/` restructuring also carried wacky values through to its coordinate
  columns.

---

## Appendix {.appendix}

### Full pipeline {#sec-pipeline}

```{r full-pipeline}
#| eval: false
# One-shot pipeline for the full archive
dat_resolved <-
  open_dataset("data-raw/data-dump/afli/rafr_sjalfvirkir_maelar.parquet") |>
  mutate(year = year(timi)) |>
  filter(year >= 2008, year <= 2020) |>
  select(visir, timi, lengd, breidd, skip_hradi, skip_stefna) |>
  left_join(
    open_dataset("data-raw/data-dump/afli/rafr_stofn.parquet") |>
      select(visir, skipnr, veidarf),
    by = "visir"
  ) |>
  collect() |>
  add_recovery_cols() |>
  arrange(visir, timi) |>
  group_by(visir) |>
  group_modify(~ resolve_track_fb(.x)) |>
  ungroup()
```

The DuckDB macro for the correct DDMMmm → decimal conversion (for use directly
in SQL / duckdbfs queries):

```sql
CREATE OR REPLACE MACRO rb_convert_DMdM(x) AS (
  SIGN(x) * (ABS(x) + (200.0/3.0) *
    ((ABS(x)/100.0) - TRUNC(ABS(x)/10000.0) * 100.0)) / 10000.0
);
```

---

### Evaluation detail {#sec-eval}

**Prediction-residual metric.** For each step between consecutive records, a
predicted next position is computed by dead-reckoning from the current resolved
position. The residual is the Haversine distance between the prediction and the
actual next resolved position.

```{r eval-compute}
#| include: false
eval_steps <- siriti_resolved |>
  arrange(visir, timi) |>
  group_by(visir) |>
  mutate(
    dt_min     = as.numeric(difftime(lead(timi), timi, units = "mins")),
    next_lon_A = lead(lon_A), next_lat_A = lead(lat_A),
    next_lon_r = lead(lon_r), next_lat_r = lead(lat_r)
  ) |>
  ungroup() |>
  filter(!is.na(dt_min), dt_min > 0, dt_min <= 30,
         !is.na(skip_hradi), !is.na(skip_stefna))

pred_A <- Map(project_pos, eval_steps$lon_A, eval_steps$lat_A,
              eval_steps$skip_hradi, eval_steps$skip_stefna, eval_steps$dt_min)
pred_r <- Map(project_pos, eval_steps$lon_r, eval_steps$lat_r,
              eval_steps$skip_hradi, eval_steps$skip_stefna, eval_steps$dt_min)

eval_steps <- eval_steps |>
  mutate(
    res_A = haversine_m(sapply(pred_A, `[[`, "lon"), sapply(pred_A, `[[`, "lat"),
                        next_lon_A, next_lat_A),
    res_r = haversine_m(sapply(pred_r, `[[`, "lon"), sapply(pred_r, `[[`, "lat"),
                        next_lon_r, next_lat_r)
  ) |>
  filter(!is.na(res_A), !is.na(res_r))
```

```{r eval-summary}
eval_steps |>
  summarise(
    `Simple recovery — median residual (m)` = round(median(res_A)),
    `Greedy resolved — median residual (m)`  = round(median(res_r)),
    `Improvement (%)`                        = round(
      (median(res_A) - median(res_r)) / median(res_A) * 100, 1)
  ) |>
  pivot_longer(everything(), names_to = "Metric", values_to = "Value")
```

::: {.callout-warning}
**Circularity caveat.** The prediction residual is the same quantity the
resolver minimises at each step, so the improvement is partly guaranteed by
construction. Turn angle (used in the main body) is an independent metric that
does not use the speed/heading signal.
:::

```{r fig-ecdf}
#| fig-cap: "ECDF of step-level prediction residuals. Greedy-resolved (red) is uniformly better than simple recovery up to ~1 500 m."
eval_steps |>
  select(res_A, res_r) |>
  pivot_longer(everything(), names_to = "Method", values_to = "residual_m",
               names_pattern = "res_(.*)") |>
  mutate(Method = recode(Method,
                         "A" = "Simple recovery (A only)",
                         "r" = "Greedy resolved")) |>
  ggplot(aes(pmin(residual_m, 3000), colour = Method)) +
  theme_bw() +
  stat_ecdf() +
  scale_x_continuous(breaks = seq(0, 3000, 500)) +
  labs(x = "Prediction residual (m, capped at 3 000)",
       y = "Cumulative proportion", colour = NULL)
```

Forward–backward agreement among ambiguous records:

```{r fb-agreement}
siriti_fb |>
  filter(confidence != "high") |>
  mutate(agree = (lon_r == lon_A) == (lon_fb == lon_A)) |>
  summarise(
    pct_agree  = round(100 * mean(agree, na.rm = TRUE), 1),
    n_disagree = sum(!agree, na.rm = TRUE)
  )
```

---

### Mathematics {#sec-math}

**Original encoding.** The DDMMmm integer encodes position as:

$$
n = \operatorname{sign} \times (DD \times 10000 + MM \times 100 + dm)
$$

where $dd \in \mathbb{Z}$, $MM \in [0,59]$, and $dm \in [0,99]$ (decimal minutes ×
100). The correct conversion is:

$$
x_{\text{true}} = \operatorname{sign}(n)\times\frac{|n| + \tfrac{200}{3}\,m}{10\,000}, \quad
m = \frac{|n|}{100} - 100\left\lfloor\frac{|n|}{10\,000}\right\rfloor
$$

**Erroneous conversion.** Treating the last two digits as seconds:

$$
x_{\text{wacky}} = \begin{cases}
dd + \dfrac{mm}{60} + \dfrac{dm}{3600} & \text{if } dm < 60 \\[6pt]
dd + \dfrac{mm+1}{60} + \dfrac{dm-60}{3600} & \text{if } dm \ge 60
\end{cases}
$$

**Extraction.** From any wacky decimal $w$, extract the embedded components:

$$
dd = \lfloor |w| \rfloor, \quad
mm = \lfloor (|w| - dd) \times 60 \rfloor, \quad
ss = \operatorname{round}\bigl(((|w| - dd) \times 60 - mm) \times 60\bigr)
$$

$ss$ is the original $dm$ value. Two candidate corrections follow:

$$
x_A = \operatorname{sign}(w)\times\!\left(dd + \frac{mm}{60} + \frac{ss}{6000}\right)
\quad (\text{assumes non-wacky})
$$

$$
x_B = \operatorname{sign}(w)\times\!\left(dd + \frac{mm-1}{60} + \frac{ss+60}{6000}\right)
\quad (\text{assumes wacky})
$$

**Constant offset.** The difference between A and B is always:

$$
\Delta = \frac{1}{60} - \frac{60}{6000} = \frac{1}{60} - \frac{1}{100} = \frac{1}{150} \approx 0.00\overline{6}°
$$

~300 m east–west and ~740 m north–south at 66°N.

**Identifiability.** If $ss \in [40, 59]$, then interpretation B would require
the original $dm = ss + 60 \in [100, 119]$, which is impossible (DDMMmm uses
hundredths of a minute, max 99). Hence $x_A$ is uniquely correct — these are the
**high-confidence** records. For $ss \in [0, 39]$, both $dm = ss$ and
$dm = ss + 60$ are valid, giving the **ambiguous** records.

---

### Algorithm {#sec-algo}

**Greedy forward propagation** processes each vessel's track in chronological
order. At step $i \to i+1$:

1. **Predict** the next position by great-circle dead-reckoning:
   $d = v_i \times \tfrac{1852}{60} \times \Delta t$ metres in the direction
   $\theta_i$ from the current resolved position.
2. **Score** candidates $A_{i+1}$ and $B_{i+1}$ by Haversine distance to the
   prediction.
3. **Select** the closer candidate; ties go to A.
4. **Fallback** to A when $\Delta t > 30$ min or speed/heading is missing,
   resetting the propagation chain.

**Forward–backward smoother** runs two independent passes (forward: left to
right; backward: right to left with reversed heading) and combines:

- If both passes choose the same interpretation → take it.
- If they disagree → score both candidates by the **sum** of distances to the
  forward and backward predictions; take the minimum.

The two passes agreed on **78%** of ambiguous records in the development subset;
the additional smoothness gain from the second pass over the greedy forward was
modest (0–1.3° median turn angle, gear-dependent).

**Limitations.**

- Errors can propagate forward (an incorrect early choice biases downstream
  decisions until the next gap).
- The 30-minute gap threshold is arbitrary.
- The resolver makes no use of the heading at the *destination* record.
- Performance degrades for slow or stationary vessels, where step distance
  falls below the ~300–740 m ambiguity.

---

### Functions {#sec-functions}

All utility functions used in this document. Copy to reuse.

```{r funs-display}
#| eval: false

# ── Coordinate conversion ─────────────────────────────────────────────────────

geo_convert <- function(x, inverse = FALSE, na = FALSE) {
  if (!inverse) {
    i   <- sign(x); x <- abs(x)
    min <- (x / 100) - trunc(x / 10000) * 100
    if (na) min <- ifelse(min > 60, NA, min)
    return((i * (x + (200 / 3) * min)) / 10000)
  } else {
    i  <- sign(x); x <- abs(x)
    p1 <- floor(x); p2 <- floor((x - p1) * 60)
    p3 <- round((x - p1 - p2 / 60) * 100 * 60)
    return(i * (p1 * 10000 + p2 * 100 + p3))
  }
}

extract_ss <- function(w) {
  w <- abs(w); dd <- floor(w); mm <- floor((w - dd) * 60)
  round(((w - dd) * 60 - mm) * 60)
}

recover_coords <- function(w, sign_w = 1) {
  w <- abs(w); dd <- floor(w); mm <- floor((w - dd) * 60)
  ss <- round(((w - dd) * 60 - mm) * 60)
  sign_w * (dd + mm / 60 + ss / 6000)
}

recover_coords_wacky <- function(w, sign_w = 1) {
  w <- abs(w); dd <- floor(w); mm <- floor((w - dd) * 60)
  ss <- round(((w - dd) * 60 - mm) * 60)
  sign_w * (dd + (mm - 1) / 60 + (ss + 60) / 6000)
}

classify_confidence <- function(ss_lon, ss_lat) {
  dplyr::case_when(
    ss_lon >= 40 & ss_lat >= 40 ~ "high",
    ss_lon  < 40 & ss_lat  < 40 ~ "low",
    TRUE                         ~ "partial"
  )
}

add_recovery_cols <- function(df) {
  df |>
    dplyr::mutate(
      ss_lon     = extract_ss(lengd),
      ss_lat     = extract_ss(breidd),
      confidence = classify_confidence(ss_lon, ss_lat),
      lon_A      = recover_coords(lengd,  sign(lengd)),
      lat_A      = recover_coords(breidd, sign(breidd)),
      lon_B      = dplyr::if_else(ss_lon < 40,
                                  recover_coords_wacky(lengd,  sign(lengd)), lon_A),
      lat_B      = dplyr::if_else(ss_lat < 40,
                                  recover_coords_wacky(breidd, sign(breidd)), lat_A)
    )
}

# ── Geodetic helpers ──────────────────────────────────────────────────────────

haversine_m <- function(lon1, lat1, lon2, lat2) {
  R <- 6371000; phi1 <- lat1 * pi / 180; phi2 <- lat2 * pi / 180
  dphi <- (lat2 - lat1) * pi / 180; dlam <- (lon2 - lon1) * pi / 180
  a <- sin(dphi/2)^2 + cos(phi1) * cos(phi2) * sin(dlam/2)^2
  2 * R * asin(sqrt(a))
}

bearing_deg <- function(lon1, lat1, lon2, lat2) {
  phi1 <- lat1*pi/180; phi2 <- lat2*pi/180; dlam <- (lon2-lon1)*pi/180
  (atan2(sin(dlam)*cos(phi2),
         cos(phi1)*sin(phi2) - sin(phi1)*cos(phi2)*cos(dlam)) * 180/pi + 360) %% 360
}

project_pos <- function(lon, lat, speed_kt, hdg_deg, dt_min) {
  dist_m <- speed_kt * (1852/60) * dt_min; R <- 6371000
  phi1 <- lat*pi/180; lam1 <- lon*pi/180; brg <- hdg_deg*pi/180; d_R <- dist_m/R
  phi2 <- asin(sin(phi1)*cos(d_R) + cos(phi1)*sin(d_R)*cos(brg))
  lam2 <- lam1 + atan2(sin(brg)*sin(d_R)*cos(phi1), cos(d_R)-sin(phi1)*sin(phi2))
  list(lon = lam2*180/pi, lat = phi2*180/pi)
}

# ── Track resolvers (see Appendix — Algorithm for description) ────────────────

resolve_track    <- function(df) { ... }   # greedy forward
resolve_track_fb <- function(df) { ... }   # forward-backward smoother
```
